### 2019-09-24 manuscript/ms.Rmd
add manuscript with abstract

---
title: "Demographic Species Distribution Models"
author:
  - Nick Golding
  - School of BioSciences, University of Melbourne, Parkville VIC, Australia
bibliography: dsdm.bib
csl: mee.csl
output:
  pdf_document:
  word_document:
    fig_caption: true
    highlight: null
---

Word Count: `r wordcountaddin::word_count()`

# Abstract

1. Quantifying the distributions and abundances of species is a problem central to ecology. Correlative Species Distribution Models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they have questionable ability to predict how distributions will change in response to environmental change, habitat loss, or human interventions.

2. I propose Demographic Species Distribution Models (DSDMs): matrix population models that are statistically fitted to species distribution data, to form a mechanistically-based statistical model for species distributions. This approach differs from existing mechanistic SDMs in that it makes use of commonly available species distribution data (abundance, presence-only, or presence-absence), explicitly models how vital rates such as fecundity and survival vary over space, and enables - but does not require - the incorporation of prior knowledge of a species' ecology.

3. Example applications mapping the Carolina Chickadee in North Ameerica, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates. Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.

4. DSDMs can be extended to consider features of population dynamics such as dispersal, density dependence, and biotic interactions. Being specifically designed for efficient statistical inference, they are a promising replacement for correlative SDMs where study aims require the incorporating of more ecological mechanism.

```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
# set up knitr options
knitr::opts_chunk$set(fig.path = 'figs/',
               message = FALSE,
               warning = FALSE,
               fig.align = 'center',
               dev = c('png'),
               cache = TRUE)
```

# Introduction

# Fitting Demographic Models to Distribution Data

# Computation

# Example 1. A DSDM for the Carolina Chickadee

# Example 2. A DSDM for *Protea repens*

### Supporting information

The code required to reproduce this manuscript is archived at [<code to be archived on acceptance>](), and hosted at [<github link>]().

### Author's contributions

NG developed the model, software and example analyses and wrote the manuscript.

### Acknowledgements

This work was funded by a McKenzie fellowship from the University of Melbourne and an ARC DECRA fellowship (DE180100635).

### Data accessibility

Carolina Chickadee data were downloaded from the North American Breeding Bird Survey FTP site. Protea data are made available as an appendix to @merowUsingIntegralProjection2014, they can be downloaded by searching for artefact ECOG-00839 at [http://www.ecography.org/readers/appendix]()

### Figures

```{r flow_diagram, echo = FALSE, fig.cap = "Figure 1. Title"}
# knitr::include_graphics("figs/diagrams.png")
```

### References

### 2019-09-24 manuscript/ms.Rmd
add schematic figure to ms

@@
 
+Advancing distribution modelling
+
+Distribution models underpin decision making in conservation1 and public health.
+The accuracy and interpretability of distribution modelling methods therefore has real-world impact.
+Correlative models (SDM and geostatistics) are the most widely used methods as they describe and interpolate distributions well with minimal prior knowledge of the system5.
+However their arbitrary, flexible functional forms make them poor at predicting to novel conditions9.
+Correlative approaches do not explicitly model the processes (demography, dispersal and interactions) the drive distributions, so are poor at incorporating prior ecological knowledge and cannot predict how distributions will change if these processes are altered, such as by human interventions11.
+Advancing distribution modelling to incorporate these processes will improve decision making in conservation and public health research12.
+
+Existing methods for mapping demography
+
+Spatial simulation models such as metapopulation viability analyses and spatial infectious disease simulations are used in conservation and epidemiology to anticipate how unstable population dynamics will evolve over a landscape.
+These approaches do not map spatial variation in demographic rates so cannot be used to predict distributions.
+Dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes12.
+DRMs model the intrinsic population growth rate rather than demographic processes so suffer the same extrapolation issues as correlative SDMs and cannot be queried to predict the effects of changing demography.
+Matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a Leslie matrix representing key demographic processes13.
+Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing Leslie matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution14.
+
 # Fitting Demographic Models to Distribution Data
 
+I have developed demographic species distribution models (DSDMs) to further extend this approach, treating the spatial IPM as a single model and fitting it to observed distribution data.
+DSDMs therefore unite two fields of quantitative ecology: statistical distribution modelling and demographic modelling (Fig.
+1).
+I have already developed a basic DSDM and the software required to perform efficient statistical inference on it (Box 1, Fig.
+2).
+This DECRA will extend this basic DSDM to incorporate three key ecological processes, enabling me to build DSDMs to solve three applied problems in public health and conservation (Fig.
+3).
+
+Parameter identifiability
+
+Like many ecologically-realistic models, DSDM parameters are in general not uniquely identified, so estimates may be correlated or even multimodal.
+This non-identifiability is ecologically informative as it arises from ignorance of the species’ ecology and highlights areas where more research is needed.
+This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions15.
+However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge16.
+This proposal uses two new tools to resolve this issue and fit DSDMs to large datasets: Hamiltonian Monte Carlo & high performance automatic differentiation.
+
 # Computation
+
+Hamiltonian Monte Carlo (HMC) is a Markov chain Monte Carlo method that can sample effectively from highly correlated and multi-modal parameter distributions16, but requires gradient calculations.
+Automatic differentiation (AD, also known as back-propagation) is a computational technique to derive these gradients.
+Whilst HMC & AD modelling software is available20, it is not optimised for the large-scale linear algebra required for DSDMs.
+High-performance AD software enables complex machine learning models (e.g. deep neural networks) to be fit to very large datasets; facilitating recent dramatic advances in artificial intelligence.
+Google’s recently released open source, high performance AD software TensorFlow is optimised for massive linear algebra21.
+I have combined HMC with TensorFlow in an R package greta19 for flexible specification of and sampling from statistical models in R.
+Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.
+
+Advancing DSDMs
+
+Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
+Where such information is available (e.g.
+from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
+The DSDMs I propose are based on Leslie matrices, which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories23.
+Like other Leslie matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models:
+
+*Density dependence* can be represented in DSDMs by modelling demographic rates as a function of the current population state23 (the abundances of each life stage) as well as covariates.
+For example, the current density of individuals could determine fecundity for the next iteration of the dynamics.
+The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.
+
+*Dispersal* can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured Leslie matrix, with dispersal represented by transition probabilities between states in different patches24.
+
+*Biotic interactions* between species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single Leslie matrix, with inter-species density dependence25.
+By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
 
@@
 
-```{r flow_diagram, echo = FALSE, fig.cap = "Figure 1. Title"}
-# knitr::include_graphics("figs/diagrams.png")
+```{r flow_diagram, echo = FALSE, fig.cap = "Demographic species distribution models in the context of other models of distribution and demography.
+A) Correlative species distribution models (SDMs) use a correlative mathematical relationship - such as a logit-linear model - to predict a species' distribution from spatial covariates.
+The predicted distribution (for a given set of parameter values) is compared to observed distribution data using an observation model - such as a Bernoulli distribution, for presence-absence data.
+Statistical inference is performed to infer the correlative model parameters from the distribution data.
+B) Matrix population models (MPMs) and integral projections models (IPMs) combine either direct estimates, or correlative models, of vital rates like survival and fecundity to predict demographic quantities such as the population growth rate.
+Spatial IPMs apply these models to covariate values at each pixel in the landscape, and therefore map the distributions of vital rates and species.
+Unlike correlative SDMs, the parameters of MPMs, IPMs, and spatial IPMs are inferred from data on population dynamics at specific locations, rather than distribution data.
+Demographic SDMs (proposed here) combine correlative SDMs and spatial IPMs; replacing the arbitrary correlative model in a correlative SDM with an MPM or IPM, and inferring model parameters by fitting the whole model statistically to observed distributions.", out.width = "\\textwidth"}
+knitr::include_graphics("../figures/schematic.png")
 ```

### 2019-09-24 schematic/mini_figs.R

@@ -1,0 +1,101 @@
+# generate mini figures for schematic figure
+set.seed(2019-09-24)
+
+# covariate data
+library(raster)
+library(fields)
+
+# download a shapefule for Victoria, Australia and convert it into a raster
+states <- raster::getData("GADM", country = "AUS", level = 1)
+vic <- states[states$NAME_1 == "Victoria", ]
+r <- raster(vic)
+res(r) <- c(0.1, 0.1)
+template <- rasterize(vic, r)
+idx <- which(!is.na(getValues(template)))
+
+# generate covariates (Gaussian processes with exponential covariance, as usual)
+coords <- xyFromCell(template, idx)
+n_cells <- length(idx)
+v <- matrix(rnorm(n_cells * 3), n_cells, 3)
+d <- rdist.earth(coords)
+Sigma <- exp(-d / 40) + diag(n_cells) * 1e-3
+L <- t(chol(Sigma))
+f <- L %*% v
+covs <- stack(template, template, template)
+covs[idx] <- f
+
+# vital rate rasters
+fec <- surv <- template
+X <- cbind(1, covs[idx])
+beta_fec <- c(0, rnorm(3, 0, 0.1))
+fec[idx] <- exp(X %*% beta_fec)[, 1]
+beta_surv <- c(-1, rnorm(3, 0, 1))
+surv[idx] <- plogis(X %*% beta_surv)[, 1]
+
+# distribution data
+# analytic solution to thedominant eigenvalue for a simple MPM
+lambda <- sqrt(surv * (fec  * 0.7 + 1))
+prob <- 1 - exp(-0.5 * lambda)
+loc <- sample.int(n_cells, 100)
+pa <- rbinom(length(loc), 1, prob[idx][loc])
+points <- cbind(coords[loc, ], pa)
+
+# predicted distribution
+pred <- template
+m <- glm(pa ~ ., data = as.data.frame(X[loc, -1]), family = binomial)
+pred[idx] <- predict(m, as.data.frame(X[, -1]), type = "response")
+
+# plots
+plot_raster <- function (r, col = cols()) {
+  image(
+    r,
+    maxpixels = Inf,
+    col = col,
+    asp = 1,
+    axes = FALSE,
+    xlab = "",
+    ylab = "",
+    bg = NA
+  )
+}
+cols <- function (pal = viridis::magma, n = 1000, rev = TRUE, fraction = 0.5) {
+  cols <- pal(n)
+  if (rev) {
[…]

### 2019-09-24 manuscript/ms.Rmd
more writing and shuffling in the ms

@@
     highlight: null
+header-includes: \usepackage{float}
 ---
@@
 Quantifying the distributions and abundances of species is a problem central to ecology.
-Correlative Species Distribution Models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they have questionable ability to predict how distributions will change in response to environmental change, habitat loss, or human interventions.
+Correlative species distribution models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they have questionable ability to predict how distributions will change in response to environmental change, habitat loss, or human interventions.
 
 2.
-I propose Demographic Species Distribution Models (DSDMs): matrix population models that are statistically fitted to species distribution data, to form a mechanistically-based statistical model for species distributions.
+I propose demographic species distribution models (DSDMs): matrix population models that are statistically fitted to species distribution data, to form a mechanistically-based statistical model for species distributions.
 This approach differs from existing mechanistic SDMs in that it makes use of commonly available species distribution data (abundance, presence-only, or presence-absence), explicitly models how vital rates such as fecundity and survival vary over space, and enables - but does not require - the incorporation of prior knowledge of a species' ecology.
@@
 # set up knitr options
-knitr::opts_chunk$set(fig.path = 'figs/',
+knitr::opts_chunk$set(fig.path = "figs/",
                message = FALSE,
                warning = FALSE,
-               fig.align = 'center',
-               dev = c('png'),
+               fig.align = "center",
+               fig.pos = "H",
+               dev = c("png"),
                cache = TRUE)
@@
 
-Advancing distribution modelling
+Species distribution modelling is widely used in ecology, with models informing ecological theory [@elithSpeciesDistributionModels2009], and the resulting maps used in a number of applied disciplines, including conservation [@guisanPredictingSpeciesDistributions2013] and public health [@purseTrackingDistributionImpacts2015].
+Correlative statistical models, using environmental variables as predictors, are very widely used for predicting species distributions as they describe and interpolate distributions well, with minimal required prior knowledge of the system [@elithSpeciesDistributionModels2009].
 
-Distribution models underpin decision making in conservation1 and public health.
-The accuracy and interpretability of distribution modelling methods therefore has real-world impact.
-Correlative models (SDM and geostatistics) are the most widely used methods as they describe and interpolate distributions well with minimal prior knowledge of the system5.
-However their arbitrary, flexible functional forms make them poor at predicting to novel conditions9.
-Correlative approaches do not explicitly model the processes (demography, dispersal and interactions) the drive distributions, so are poor at incorporating prior ecological knowledge and cannot predict how distributions will change if these processes are altered, such as by human interventions11.
-Advancing distribution modelling to incorporate these processes will improve decision making in conservation and public health research12.
+However the arbitrary, flexible functional forms of these correlative species distribution models (SDMs) make them poor at predicting to novel conditions [@dormannCorrelationProcessSpecies2012].
+Moreover, models of species distributions are often needed to predict new distributions in response to changes other than the environmental predictors that correlative SDMs rely upon. For instance
 
-Existing methods for mapping demography
+Whilst users typically have some knowledge of the processes driving these aspects of species distributions - such as the species' life history, approximate survival and fecundity rates or dispersal ability - incorporating this knowledge into correlative statistical models is problematic.
+On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed pyhsiological modelling [@kearneyModellingSpeciesDistributions2008].
 
-Spatial simulation models such as metapopulation viability analyses and spatial infectious disease simulations are used in conservation and epidemiology to anticipate how unstable population dynamics will evolve over a landscape.
-These approaches do not map spatial variation in demographic rates so cannot be used to predict distributions.
-Dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes12.
-DRMs model the intrinsic population growth rate rather than demographic processes so suffer the same extrapolation issues as correlative SDMs and cannot be queried to predict the effects of changing demography.
-Matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a Leslie matrix representing key demographic processes13.
+When no single modelling approach is available that fits their needs, species distribution modellers needing to incorporate more ecological mechanism into their predictions must employ a number of tricks to shoehorn their ecological knowledge into correlative SDMs.
+For instance, manipulating distribution data to remove the signal of dispersal limitation before model fitting [@elithArtModellingRangeshifting2010], or treating correlative SDM predictions as a proxy for habitat carrying capacity in a secondary spatial population simulation model [@andersonDynamicsRangeMargins2009].
+Whilst these approaches can be effective, they often make strong and questionable assumptions (for example about the relationship between habitat suitability and carrying capacity) and require the modeller to focus on implementation details, rather than the ecology of the species they are studying.
+Models that explicitly consider the demographic processes relevant to distribution modelling task in hand are therefore appealing.
+
+A number of models have been made to improve the realism in species distribution models, whilst still enabling them to be fitted to the distribution data that are commonly available.
+For example, dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes and spatial variation in population growth rates [@pagelForecastingSpeciesRanges2012].
+DRMs model population growth rates using a correlative model and environmental covariates, just as a correlative SDM models the abundance or probability of presence of a species. 
+DRMs can also be fitted to distribution data, though they are intended for use with timeseries of abundance or presence-only data, since repeat observations are needed to accurately quantify population growth rates and dispersal.
+
+Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change - for example a conservation intervention like predator control - that affects a single vital rate.
+
+<explicitly modelling vital rates, because it's useful>
+
+<MPMs & IPMs do this at a single location - widely used in ecology, and even whole databases available>
+
+By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a Leslie matrix representing key demographic processes13.
 Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing Leslie matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution14.
 
-# Fitting Demographic Models to Distribution Data
+<SIPMs (DDMs in Merow - note on nomenclature) apply these to large spatial scales.
+Whilst they rely on vast amnounts field data (so are not applicable in most SDM situations), they provide a link between demography and distribution, so we could perform statistical inference on them using distribution data.
+This is what we call a DSDM - the subject of this manuscript.
 
-I have developed demographic species distribution models (DSDMs) to further extend this approach, treating the spatial IPM as a single model and fitting it to observed distribution data.
-DSDMs therefore unite two fields of quantitative ecology: statistical distribution modelling and demographic modelling (Fig.
-1).
-I have already developed a basic DSDM and the software required to perform efficient statistical inference on it (Box 1, Fig.
-2).
-This DECRA will extend this basic DSDM to incorporate three key ecological processes, enabling me to build DSDMs to solve three applied problems in public health and conservation (Fig.
-3).
+Whilst the concept is simple, it is not obvious that multiple spatial surfaces (vital rate distributions) can be inferred from a single spatial dataset of a species' distribution.
+Nor is it obvious that such a model can be fitted to data with sufficiently ease and speed to be useful to ecologists.
+The rest of this manuscript addresses these issues, introducing the basic model and describing a number of potential extensions.
+Example 1 illustrates a basic DSDM for presence/absence distribution data, requiring minimal prior knowledge and demonstrates that spatial variation in the two different vital rates can indeed be disentangled.
+Example 2 then develops a more complex DSDM incorporating detailed knowledge of a species' ecology, and illustrates that the relationships between vital rates and environmental variables learned from distribution data do broadly agree with field-derived estimates.
 
-Parameter identifiability
+# Demographic Species Distribution Models
+
+<general equations - a number of options for observation models>
+<blame link between growth rate and probability of presence on Jorn, point reader to section on extensions for density-dependent solution>
+<suggest doing Bayesian inference - propagate uncertainty and incorporate any prior information to help the model disentangle the spatial signal>
+
+```{r schematic, echo = FALSE, fig.cap = "Demographic species distribution models in the context of other models of distribution and demography.
+A) Correlative species distribution models (SDMs) use a correlative mathematical relationship - such as a logit-linear model - to predict a species' distribution from spatial covariates.
+The predicted distribution (for a given set of parameter values) is compared to observed distribution data using an observation model - such as a Bernoulli distribution, for presence-absence data.
+Statistical inference is performed to infer the correlative model parameters from the distribution data.
+B) Matrix population models (MPMs) and integral projections models (IPMs) combine either direct estimates, or correlative models, of vital rates like survival and fecundity to predict demographic quantities such as the population growth rate.
+Spatial IPMs apply these models to covariate values at each pixel in the landscape, and therefore map the distributions of vital rates and species.
+Unlike correlative SDMs, the parameters of MPMs, IPMs, and spatial IPMs are inferred from data on population dynamics at specific locations, rather than distribution data.
+Demographic SDMs (proposed here) combine correlative SDMs and spatial IPMs; replacing the arbitrary correlative model in a correlative SDM with an MPM or IPM, and inferring model parameters by fitting the whole model statistically to observed distributions.
+Equations correspond to the model considered in Example 1, but represent a wide range of correlative, population dynamic, and observation models that could be used in this framework.", out.width = "\\textwidth"}
+knitr::include_graphics("../figures/schematic.png")
+```
+
+# Example 1. A DSDM for the Carolina Chickadee
+
+This box demonstrates the proposed approach by constructing and analysing a simple DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
+A two-stage matrix population model for the species has been developed for the USA [@ryuDevelopingPopulationModels2016], though this demographic model assumed demographic rates to be constant across the continent, rather than accounting for spatial variation in them.
+
+I constructed the DSDM as a single statistical model, using these national average fecundity and adult survival estimates as a prior distribution and modelling environmental variation in these rates via log- and logit-linear relationships with six environmental variables: cover of three forest types used for breeding habitat (fecundity), minimum temperature as a thermal limit (survival) and annual precipitation and mean temperature as proxies for insect abundance (fecundity and survival).
+Juvenile and adult survival differed by a constant value on the logit scale [@ryuDevelopingPopulationModels2016].
+
+For each location, these rates were used to generate a 2x2 Leslie matrix and iterated to estimate the intrinsic growth rate $\lambda$.
+As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
+The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the Breeding Bird Survey of North America in 2015 [@todo].
+
+Inference was by Hamiltonian Monte Carlo using the R package greta and its extension package greta.dynamics for modelling dynamical systems.
+Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence ($\hat{R}\le1.01$), and yield at least X effective samples, for all model parameters.
+This took Y seconds on a laptop computer.
+
+The fitted DSDM yields similar predictions of the current distribution, and has similar accuracy, to a correlative SDM (a Bayesian logistic regression fitted  by MCMC in a similar way) (Fig. 2A).
+Unlike a correlative SDM, the DSDM also predicts spatial variation in the survival (Fig. 2B) and fecundity (Fig. 2C) rates.
+The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda \ge 1$, and the model can predict the distribution under reduced fecundity (e.g.
+due to an introduced competitor) or survival (e.g.
+due to an introduced pathogen) (Fig.
+2D).
+
+```{r chickadee, echo = FALSE, fig.cap = "Figure 2.
+Spatial predictions from the fitted DSDM.
+A) Probability of presence predicted by the DSDM, with 2015 BBS presence (black dots) and absence (grey dots) records overlaid.
+B) Spatial variation in fecundity (juveniles per adult per year).
+C) Spatial variation in annual adult survival probability.
+D) The species’ range limits (where λ≥1) under this baseline model (light blue area), a unilateral 50% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
+The predictions in panels B, C and D cannot be made using a correlative SDM.", out.width = "\\textwidth"}
+# knitr::include_graphics("../figures/chickadee.png")
+```
+
+# Extending the basic DSDM
+
+Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
+Where such information is available (e.g.
+from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
+The DSDMs I propose are based on Leslie matrices, which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories23.
+Like other Leslie matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.
+
+## Density dependence
+
+Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state23 (the abundances of each life stage) as well as covariates.
+For example, the current density of individuals could determine fecundity for the next iteration of the dynamics.
+The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.
+
+## Dispersal
+
+Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured Leslie matrix, with dispersal represented by transition probabilities between states in different patches24.
+
+## Biotic interactions
+
+Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single Leslie matrix, with inter-species density dependence25.
+By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
+
+# Example 2. A DSDM for *Protea repens*
+
+Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty those predictions, it we do not have sufficient spatially-varying data.
+
+# Discussion
+
+## Parameter identifiability
 
@@
 This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions15.
+
+Example 1 indicates that identifiability issues do not affect predictions of the probability of presence - even if the data and priors do not provide sufficient information to disentangle vital rates.
+This example also suggests that a minimal amount of information is required to make the vital rates identifiable.
+These issues of identifiability warrant more detailed investigation, and healthy scepticism on the part of users.
+However since identifiability will depend on the nature of the data, model, and prior information, a detailed assessment of model identifiability is beyond the scope of this manuscript.
+
 However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge16.
@@
 
-# Computation
+##  Computation
 
@@
 Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.
-
-Advancing DSDMs
-
-Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
-Where such information is available (e.g.
-from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
-The DSDMs I propose are based on Leslie matrices, which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories23.
-Like other Leslie matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models:
-
-*Density dependence* can be represented in DSDMs by modelling demographic rates as a function of the current population state23 (the abundances of each life stage) as well as covariates.
-For example, the current density of individuals could determine fecundity for the next iteration of the dynamics.
-The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.
-
-*Dispersal* can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured Leslie matrix, with dispersal represented by transition probabilities between states in different patches24.
-
-*Biotic interactions* between species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single Leslie matrix, with inter-species density dependence25.
-By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
-
-# Example 1. A DSDM for the Carolina Chickadee
-
-# Example 2. A DSDM for *Protea repens*
 
@@
 
-```{r flow_diagram, echo = FALSE, fig.cap = "Demographic species distribution models in the context of other models of distribution and demography.
-A) Correlative species distribution models (SDMs) use a correlative mathematical relationship - such as a logit-linear model - to predict a species' distribution from spatial covariates.
-The predicted distribution (for a given set of parameter values) is compared to observed distribution data using an observation model - such as a Bernoulli distribution, for presence-absence data.
-Statistical inference is performed to infer the correlative model parameters from the distribution data.
-B) Matrix population models (MPMs) and integral projections models (IPMs) combine either direct estimates, or correlative models, of vital rates like survival and fecundity to predict demographic quantities such as the population growth rate.
-Spatial IPMs apply these models to covariate values at each pixel in the landscape, and therefore map the distributions of vital rates and species.
-Unlike correlative SDMs, the parameters of MPMs, IPMs, and spatial IPMs are inferred from data on population dynamics at specific locations, rather than distribution data.
-Demographic SDMs (proposed here) combine correlative SDMs and spatial IPMs; replacing the arbitrary correlative model in a correlative SDM with an MPM or IPM, and inferring model parameters by fitting the whole model statistically to observed distributions.", out.width = "\\textwidth"}
-knitr::include_graphics("../figures/schematic.png")
-```
+[To be moved here later]
 

### 2019-10-02 R/fetch_data.R
add download of bbs data and tidy drake plan

@@ -12,18 +12,14 @@ download_merow <- function () {
   file.remove(zip_file) -> .
 
   # find the path to the RData object and return that
-  list.files(
+  file_path <- list.files(
     "data/raw/ECOG-00839 Appendix/",
     pattern = ".Rdata$",
     full.names = TRUE
   )
 
-}
-
-# read in the RData object and output a tidied up version with the objects we
-# want
-tidy_merow <- function (file_path) {
-
+  # read in the RData object and output a tidied up version with the objects we
+  # want
   load(file_path)
 
   # extract the relevant parts of the cape floristic region atlas data
@@ -49,8 +45,116 @@ tidy_merow <- function (file_path) {
       summer_smd = smdsum,
       winter_smd = smdwin,
       mean_annual_precip = map
-    ) -> occurrence
+    )
 
-  occurrence
+}
+
+# get landcover and bioclim covariates for the BBS example and
+# remove extranous areas
+download_bbs_rasters <- function () {
+
+  # get CarolineWren rasters from maxlike package and convert to a raster
+  data(carw, envir = environment())
+  rl <- lapply(carw.data$raster.data, function(x) {
+    m <- matrix(x, nrow = carw.data$dim[1], ncol = carw.data$dim[2],
+                byrow = TRUE)
+    raster::raster(m)
+  })
+  rs <- raster::stack(rl)
+  extent(rs) <- carw.data$ext
+  names(rs) <- names(carw.data$raster.data)
+  projection(rs) <- CRS("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=m +no_defs")
+  covs1 <- raster::projectRaster(rs, crs = "+init=epsg:4326")
+
+  # bioclim layers
+  covs2 <- raster::getData("worldclim", var = "bio", res = 10)
+  covs2 <- raster::crop(covs2, c(-128, -66, 21, 52))
+
+  # align them
+  covs1 <- resample(covs1, covs2[[1]])
+  covs2 <- mask(covs2, covs1[[1]])
+  covs <- stack(covs1, covs2)
+
+  covs
+  # writeRaster(covs, file = "data/clean/bbs_covs")
+}
+
+clean_bbs <- function(file, AOU, years = 2015) {
+
[…]

### 2019-10-02 R/modelling.R

@@ -20,7 +20,7 @@ tf_log_normalise <- function (x) {
 
 # load the parameter estimates from Merow et al. and format them as the
 # parameters of normal priors
-prior <- function (model = c("seedling_survival", "adult_survival", "growth",
+protea_prior <- function (model = c("seedling_survival", "adult_survival", "growth",
                              "flowering", "seedheads", "seeds", "germination", "offspring_size"),
                    type = c("environment", "size"),
                    informative = TRUE) {
@@ -194,7 +194,7 @@ interpolate_survival <- function (adult_survival_matrix,
 # dimensions as A . Note that this is only necessary because greta's sweep
 # doesn't yet handle arrays (with more than 2 dimensions), otherwise we could
 # just do: sweep(A, B, MARGIN = c(1, 3), FUN = FUN)
-sweep_3 <- function(A, B, FUN = c('-', '+', '/', '*')) {
+sweep_3 <- function(A, B, FUN = c("-", "+", "/", "*")) {
   n <- dim(A)[1]
   m <- dim(A)[2]
   Al <- aperm(A, c(2, 3, 1))
@@ -207,7 +207,7 @@ sweep_3 <- function(A, B, FUN = c('-', '+', '/', '*')) {
 }
 
 # build and return a greta model for the DSDM
-build_model <- function (occurrence, informative_priors = FALSE) {
+build_protea_model <- function (occurrence, informative_priors = FALSE) {
 
   # subset occurrence data to where we observe presence/absence, and subset to
   # a few observations (for now)
@@ -229,7 +229,7 @@ build_model <- function (occurrence, informative_priors = FALSE) {
 
   beta_priors <- lapply(
     names(formulae),
-    prior,
+    protea_prior,
     type = "environment",
     informative = informative_priors
   )
@@ -400,13 +400,13 @@ run_mcmc <- function (model_list) {
 
 # use the model and MCMC parameter samples to visualise the fitted vital rate
 # relationships (using calculate and relevant design matrices)
-plot_relationships <- function(draws, model_list, occurrence) {
+plot_protea_relationships <- function(draws, model_list, occurrence) {
 
 }
 
 # use the model and MCMC parameter samples to make posterior predictions of
 # the probability of presence in new places (using calculate and relevant design matrices)
-make_predictions <- function(draws, model_list, occurrence) {
+make_protea_predictions <- function(draws, model_list, occurrence) {
 
 }
 

### 2019-10-02 main.R

@@ -7,12 +7,14 @@ set.seed(2019-06-19)
 # packages ----------------------------------------------------------------
 
 # reproducible workflow
+library(future.apply)
 library(drake)
 
 # data munging and plotting
 library(raster)
 library(RColorBrewer)
 library(dplyr)
+library(maxlike)
 
 # modelling
 library(greta.dynamics)
@@ -21,37 +23,49 @@ library(greta.dynamics)
 
 source("R/fetch_data.R")
 
-read_data <- drake_plan(
-  merow_file = download_merow(),
-  occurrence = tidy_merow(merow_file),
-  tidy_file = saveRDS(occurrence, "data/clean/occurrence.RDS")
+read_all_data <- drake_plan(
+
+  # bbs analysis
+  bbs_covs = download_bbs_rasters(),
+  tidy_covs = writeRaster(bbs_covs, "data/clean/bbs_covs", overwrite = TRUE),
+  bbs_occurrence = download_bbs(),
+  tidy_occ = saveRDS(bbs_occurrence, "data/clean/bbs_occ.RDS"),
+
+  # protea analysis
+  protea_occurrence = download_merow(),
+  tidy_file = saveRDS(protea_occurrence, "data/clean/protea_occurrence.RDS")
+
 )
 
-make(read_data)
+# download BBS data chunks in parallel
+old_plan <- future::plan()
+future::plan(multisession)
+make(read_all_data)
+future::plan(old_plan)
 
 # plot data  --------------------------------------------------------
 
 source("R/plotting.R")
 
-plot_data <- drake_plan(
-  occurrence = readRDS(file_in("data/clean/occurrence.RDS")),
-  covs_fig = plot_covs(occurrence, file_out("figures/covs.png")),
-  occ_fig = plot_occ(occurrence, file_out("figures/occ.png"))
+plot_protea_data <- drake_plan(
+  protea_occurrence = readRDS(file_in("data/clean/protea_occurrence.RDS")),
+  protea_covs_fig = plot_covs(protea_occurrence, file_out("figures/protea_covs.png")),
+  protea_occ_fig = plot_occ(protea_occurrence, file_out("figures/protea_occ.png"))
 )
 
-make(plot_data)
+make(plot_protea_data)
 
 
 # fit model ---------------------------------------------------------------
 
 source("R/modelling.R")
 
-fit_model <- drake_plan(
-  occurrence = readRDS(file_in("data/clean/occurrence.RDS")),
-  model_list = build_model(occurrence),
[…]

### 2019-10-02 manuscript/ms.Rmd
add citations into old text

@@
 
-By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a Leslie matrix representing key demographic processes13.
-Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing Leslie matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution14.
+By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
+Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution [@merowUsingIntegralProjection2014].
 
@@
 
-For each location, these rates were used to generate a 2x2 Leslie matrix and iterated to estimate the intrinsic growth rate $\lambda$.
+For each location, these rates were used to generate a 2x2 transition matrix (a Lefkovitch matrix) and iterated to estimate the intrinsic growth rate $\lambda$.
 As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
-The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the Breeding Bird Survey of North America in 2015 [@todo].
+The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
 
@@
 from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
-The DSDMs I propose are based on Leslie matrices, which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories23.
-Like other Leslie matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.
+The DSDMs I propose are based on transition matrices (including the matrices used in both MPMs and IPMs), which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories [@ellnerIntegralProjectionModels].
+Like other transition matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.
 
@@
 
-Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state23 (the abundances of each life stage) as well as covariates.
+Density dependence can be represented in DSDMs by modelling demographic rates as a function of the current population state [@ellnerIntegralProjectionModels] (the abundances of each life stage) as well as covariates.
 For example, the current density of individuals could determine fecundity for the next iteration of the dynamics.
@@
 
-Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured Leslie matrix, with dispersal represented by transition probabilities between states in different patches24.
+Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured transition matrix, with dispersal represented by transition probabilities between states in different patches [@neubertDEMOGRAPHYDISPERSALCALCULATION2000].
 
@@
 
-Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single Leslie matrix, with inter-species density dependence25.
+Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single transition matrix, with inter-species density dependence [@fujiwaraCoexistenceCompetingStagestructured2011].
 By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
@@
 This non-identifiability is ecologically informative as it arises from ignorance of the species’ ecology and highlights areas where more research is needed.
-This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions15.
+This ignorance can be captured in prediction uncertainty, more realistically representing how well we can predict distributions [@bealeIncorporatingUncertaintyPredictive2012].
 
@@
 
-However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge16.
+However, poor identification complicates statistical inference as maximum likelihood methods will be inaccurate and Markov chain Monte Carlo samplers may not converge [@monnahanFasterEstimationBayesian2017].
 This proposal uses two new tools to resolve this issue and fit DSDMs to large datasets: Hamiltonian Monte Carlo & high performance automatic differentiation.
@@
 
-Hamiltonian Monte Carlo (HMC) is a Markov chain Monte Carlo method that can sample effectively from highly correlated and multi-modal parameter distributions16, but requires gradient calculations.
+Hamiltonian Monte Carlo (HMC) is a Markov chain Monte Carlo method that can sample effectively from highly correlated and multi-modal parameter distributions [@monnahanFasterEstimationBayesian2017], but requires gradient calculations.
 Automatic differentiation (AD, also known as back-propagation) is a computational technique to derive these gradients.
-Whilst HMC & AD modelling software is available20, it is not optimised for the large-scale linear algebra required for DSDMs.
+Whilst HMC & AD modelling software is available [@carpenterStanProbabilisticProgramming2017], it is not optimised for the large-scale linear algebra required for DSDMs.
 High-performance AD software enables complex machine learning models (e.g. deep neural networks) to be fit to very large datasets; facilitating recent dramatic advances in artificial intelligence.
-Google’s recently released open source, high performance AD software TensorFlow is optimised for massive linear algebra21.
-I have combined HMC with TensorFlow in an R package greta19 for flexible specification of and sampling from statistical models in R.
+Google’s recently released open source, high performance AD software TensorFlow is optimised for massive linear algebra [@abadiTensorFlowLargeScaleMachine2016].
+I have combined HMC with TensorFlow in an R package greta [@goldingGretaSimpleScalable2019] for flexible specification of and sampling from statistical models in R.
 Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.

### 2019-10-02 R/modelling.R
copy over the rest of the BBS analysis

@@ -1,3 +1,325 @@
+# modelling functions for boht the BBS and protea analyses
+
+# given a vector greta array, where each element corresponds to a cell in the
+# raster, an mcmc.list object of posterior samples, return a raster giving the
+# posterior mean of the greta array in each cells
+map_variable <- function (greta_array, draws, raster_template) {
+  vals <- calculate(greta_array, draws)
+  vals_mat <- as.matrix(vals)
+  mean <- colMeans(vals_mat)
+  map <- raster_template$raster
+  map[raster_template$idx] <- mean
+  map
+}
+
+# scale covariates by named vectors of means and standard deviations
+scale_covs <- function (covs, means, sds) {
+  cols_scale <- match(names(means), colnames(covs))
+  covs_sub <- covs[, cols_scale]
+  covs_sub <- sweep(covs_sub, 2, means, "-")
+  covs_sub <- sweep(covs_sub, 2, sds, "/")
+  covs[, cols_scale] <- covs_sub
+  covs
+}
+
+surv_covs <- c("bio1", "bio12", "bio6")
+fec_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio12")
+
+# calculate survival, given covariates
+get_survival <- function (covs, params) {
+  x_survival <- covs[, surv_covs]
+  survival_logit <- x_survival %*% params$beta_survival + params$default_logit_survival
+  survival_adult <- ilogit(survival_logit)
+  juvenile_offset <- qlogis(0.489) - qlogis(0.710)
+  survival_juvenile <- ilogit(survival_logit + juvenile_offset)
+  list(adult = survival_adult,
+       juvenile = survival_juvenile)
+}
+
+get_fecundity <- function (covs, params) {
+  x_fecundity <- covs[, fec_covs]
+  fecundity_log <- x_fecundity %*% params$beta_fecundity + params$default_log_fecundity
+  exp(fecundity_log)
+}
+
+get_lambda <- function(survival, fecundity) {
+
+  # build n x m x m 3D array containing Leslie matrices for different locations
+  top_row <- cbind(fecundity * survival$juvenile,
+                   fecundity * survival$adult)
+  bottom_row <- cbind(survival$juvenile,
+                      survival$adult)
+  matrices <- abind(top_row, bottom_row, along = 3)
+
[…]

### 2019-10-02 main.R

@@ -27,13 +27,13 @@ read_all_data <- drake_plan(
 
   # bbs analysis
   bbs_covs = download_bbs_rasters(),
-  tidy_covs = writeRaster(bbs_covs, "data/clean/bbs_covs", overwrite = TRUE),
+  tidy_covs = writeRaster(bbs_covs, file_out("data/clean/bbs_covs.grd"), overwrite = TRUE),
   bbs_occurrence = download_bbs(),
-  tidy_occ = saveRDS(bbs_occurrence, "data/clean/bbs_occ.RDS"),
+  tidy_occ = saveRDS(bbs_occurrence, file_out("data/clean/bbs_occ.RDS")),
 
   # protea analysis
   protea_occurrence = download_merow(),
-  tidy_file = saveRDS(protea_occurrence, "data/clean/protea_occurrence.RDS")
+  tidy_file = saveRDS(protea_occurrence, file_out("data/clean/protea_occurrence.RDS"))
 
 )
 
@@ -42,6 +42,32 @@ old_plan <- future::plan()
 future::plan(multisession)
 make(read_all_data)
 future::plan(old_plan)
+
+# fit models ---------------------------------------------------------------
+
+source("R/modelling.R")
+
+fit_bbs_model <- drake_plan(
+  bbs_occurrence = readRDS(file_in("data/clean/bbs_occ.RDS")),
+  bbs_covs = brick(file_in("data/clean/bbs_covs.grd")),
+  model_list = build_bbs_model(bbs_occurrence, bbs_covs),
+  draws_list = run_mcmc(model_list),
+  predictions = make_bbs_predictions(model_list, draws_list, bbs_covs),
+  plots = plot_bbs_maps(predictions, model_list),
+  stats = compare_bbs_predictions(predictions, model_list)
+)
+
+make(fit_bbs_model)
+
+fit_protea_model <- drake_plan(
+  protea_occurrence = readRDS(file_in("data/clean/protea_occurrence.RDS")),
+  model_list = build_protea_model(protea_occurrence),
+  draws_list = run_mcmc(model_list),
+  relationships_fig = plot_protea_relationships(model_list, draws_list, protea_occurrence),
+  predictions = make_protea_predictions(model_list, draws_list, protea_occurrence)
+)
+
+make(fit_protea_model)
 
 # plot data  --------------------------------------------------------
 
@@ -55,18 +81,4 @@ plot_protea_data <- drake_plan(
 
 make(plot_protea_data)
 
-
[…]

### 2019-10-02 manuscript/ms.Rmd
flesh out example 1 text

@@
                cache = TRUE)
+```
+
+```{r stats, echo = FALSE, cache = FALSE, eval = TRUE}
+# read in statistics from modelling to insert in text
+
+bbs_stats <- readRDS("../data/clean/bbs_stats.RDS")
+bbs_stats <- lapply(bbs_stats, round, 3)
+
 ```
@@
 
-This box demonstrates the proposed approach by constructing and analysing a simple DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
-A two-stage matrix population model for the species has been developed for the USA [@ryuDevelopingPopulationModels2016], though this demographic model assumed demographic rates to be constant across the continent, rather than accounting for spatial variation in them.
+This section demonstrates the construction, fitting and analysis of a basic DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
+A two-stage matrix population model for the species has previously been developed for the USA [@ryuDevelopingPopulationModels2016], and used to carry out a population viability analysis.
+Parameters of the model - fecundity and survival rates for adults and juveniles - were estimated from mark-recapture data across the United States, and these estimates and the resulting matrix therefore represent a single spatial average of these rates over the continent.
+The model structure is:
 
-I constructed the DSDM as a single statistical model, using these national average fecundity and adult survival estimates as a prior distribution and modelling environmental variation in these rates via log- and logit-linear relationships with six environmental variables: cover of three forest types used for breeding habitat (fecundity), minimum temperature as a thermal limit (survival) and annual precipitation and mean temperature as proxies for insect abundance (fecundity and survival).
-Juvenile and adult survival differed by a constant value on the logit scale [@ryuDevelopingPopulationModels2016].
+\begin{align}
+\tilde{\mathbf{A}} = \begin{pmatrix} \widetilde{F} \cdot \widetilde{S}_J & \widetilde{F} \cdot \widetilde{S}_A \\ \widetilde{S}_J & \widetilde{S}_A \end{pmatrix}
+\end{align}
 
-For each location, these rates were used to generate a 2x2 transition matrix (a Lefkovitch matrix) and iterated to estimate the intrinsic growth rate $\lambda$.
-As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
-The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
+where I use $\tilde{\mathbf{A}}$ to denote a transition matrix representing the continental average population, and $\widetilde{F}$, $\widetilde{S}_j$, and $\widetilde{S}_a$ are similarly spatially-averaged annual fecundity, juvenile survival, and adult survival rates respectively.
+Tables 2-4 in @ryuDevelopingPopulationModels2016 give the following parameter estimates, along with estimates of variability expressed here as standard deviations: $\widetilde{F}$ = 0.58 ($\pm$ 0.25), $\widetilde{S}_j$ = 0.710 ($\pm$ 0.195), $\widetilde{S}_a$ = 0.489 ($\pm$ 0.195)
 
-Inference was by Hamiltonian Monte Carlo using the R package greta and its extension package greta.dynamics for modelling dynamical systems.
+Whilst these estimates provide detailed information with which to predict population trajectories in the US as a whole, they do not provide any information about how vital rates varies across the continent, and don't help use to predict the species' distribution.
+To this end, I constructed a DSDM based on this model, but where the population in each location $i$ has a distinct set of vital rates and resulting population transition matrix:
+
+\begin{align}
+\mathbf{A}_i = \begin{pmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{pmatrix}
+\end{align}
+
+I construct log-linear and logit-linear models for the fecundity and survival rates, such that they vary spatially as a function of environmental covariates, just as the probability of presence does in a correlative SDM:
+
+\begin{align}
+ln(F_i) &= ln(\widetilde{F}) + \mathbf{X}_{Fi}\beta_F \\ \nonumber
+logit(S_{Ai}) &= logit(\widetilde{S}_A) + \mathbf{X}_{Si}\beta_S \\ \nonumber
+logit(S_{Ji}) &= logit(S_{Ai}) - 0.94 \\ \nonumber
+\widetilde{F} &\sim N^+(-0.545, 0.25^2) \\ \nonumber
+\widetilde{S}_A &\sim N^{(0,1)}(0.895, 0.195^2) \\ \nonumber
+\beta_{Si} &\sim N(0, 10) \\ \nonumber
+\beta_{Fi} &\sim N(0, 10)
+\end{align}
+
+where $\mathbf{X}_{F}$ and $\mathbf{X}_{S}$ are matrices of covariates for the fecundity and survival submodels respectively, and $\beta_F$ and $\beta_S$ are the corresponding vectors of regression coefficients, to which we assign minimally-informative priors.
+These design matrices do not include dummy intercept columns, but instead covariates are centred so that their average value is zero over the contiguous USA, and the intercept is modelled via a separate parameter representing the spatial average of these rates.
+We use the estimates of @ryuDevelopingPopulationModels2016 as informative priors for these continental average rates, with normal distributions truncated to the appropriate range of values (positive values for fecundity, and the unit interval for survival).
+Rather than separately modelling spatial variation in adult and juvenile fecundity, we assume that the environmental predictors of survival are the same in both stages, but that juvenile survival is consistently lower than adult survival.
+We fix the difference between adult and juvenile survival (on the logit scale to maintain the 0-1 constraints on survival parameters) to be the difference between the continental average estimates ($logit(0.710) - logit(0.489) = 0.94$).
+
+Finally, we specify and observation model for the presence-absence data in the breeding-bird survey:
+
+\begin{align}
+y_i &\sim Bernoulli(p_i) \\ \nonumber
+p_i &= 1 - e^{-N_i} \\ \nonumber
+N_i &= \alpha \lambda_i  \\ \nonumber
+\lambda_i &= eigval(\mathbf{A_i}) \\ \nonumber
+\alpha &\sim U(0, \infty)
+\end{align}
+
+where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$.
+We make a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure.
+This assumption is made in other population models such as dynamic range models [@pagelForecastingSpeciesRanges2012], and as we discuss in the text below DSDMs can easily incorporate more complex and realistic forms of density-dependence.
+The parameter $\alpha$ represents the product of both this constant of proportionality, and the fraction of individuals in the population detected in this year of the survey, and we assign it an improper uniform prior on all positive values.
+This yields $N_i$, the expected number of individuals detected during the survey, from which we can calculate the Bernoulli probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
+
+We fitted the model to a random subset of 500 presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
+We used as six environmental covariates as predictors in the fecundity and survival models, with partial overlap i the predictors between the models.
+Percentage cover of deciduous forest, coniferous forest, and mixed forest (data provided in [@royleLikelihoodAnalysisSpecies2012]) were used as covariates for fecundity, since they represent available breeding habitat.
+Minimum temperature (from @fickWorldClimNew1km2017) was used in the survival model, to capture thermal limits on survival of the species, and annual precipitation and mean temperature from the same data source were used in both the fecundity and survival models, as proxies for insect abundance, which is expected to drive both rates in this insectovorous species.
+
+Inference was by Hamiltonian Monte Carlo using the R package greta [@goldingGretaSimpleScalable2019] and its extension package greta.dynamics [@goldingGretaDynamicsModelling2019] for modelling dynamical systems.
 Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence ($\hat{R}\le1.01$), and yield at least X effective samples, for all model parameters.
@@
 
-The fitted DSDM yields similar predictions of the current distribution, and has similar accuracy, to a correlative SDM (a Bayesian logistic regression fitted  by MCMC in a similar way) (Fig. 2A).
+The fitted DSDM yields predictions of the probability of presence like those from a correlative SDM (Fig.
+2A).
+The predicted probabilities are very similar and have similar accuracy: predictions from a maximum likelihood logistic regression model fitted to the same data had a correlation of `r bbs_stats$glm_dsdm_correl` with the DSDM predictions, and similar AUC and correlation when compared with the true data in the 1,933 survey locations not used to train the model (GLM: AUC = `r bbs_stats$glm_auc`, correlation = `r bbs_stats$glm_pa_correl`, DSDM: AUC = `r bbs_stats$dsdm_auc`, correlation = `r bbs_stats$dsdm_pa_correl`).
 Unlike a correlative SDM, the DSDM also predicts spatial variation in the survival (Fig. 2B) and fecundity (Fig. 2C) rates.
-The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda \ge 1$, and the model can predict the distribution under reduced fecundity (e.g.
+The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda_i \ge 1$, and the model can predict the distribution under reduced fecundity (e.g.
 due to an introduced competitor) or survival (e.g.
@@
 
-```{r chickadee, echo = FALSE, fig.cap = "Figure 2.
-Spatial predictions from the fitted DSDM.
+```{r bbs, echo = FALSE, fig.cap = "Spatial predictions from the fitted DSDM.
 A) Probability of presence predicted by the DSDM, with 2015 BBS presence (black dots) and absence (grey dots) records overlaid.
@@
 C) Spatial variation in annual adult survival probability.
-D) The species’ range limits (where λ≥1) under this baseline model (light blue area), a unilateral 50% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
+D) The species' range limits (where $\\lambda_i \\ge 1$) under this baseline model (light blue area), a unilateral 50\\% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
 The predictions in panels B, C and D cannot be made using a correlative SDM.", out.width = "\\textwidth"}
-# knitr::include_graphics("../figures/chickadee.png")
+knitr::include_graphics("../figures/bbs_fig.png")
 ```
+
+Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure.
+As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller.
+Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatiall-yvarying vital rates.
 
@@
 
-Basic DSDMs, like the model in Example 1, do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
+Basic DSDMs like the model in Example 1 do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
 Where such information is available (e.g.

### 2019-10-02 R/modelling.R
code tweaks and compare DSDM with GLM for BBS data

@@ -12,18 +12,9 @@ map_variable <- function (greta_array, draws, raster_template) {
   map
 }
 
-# scale covariates by named vectors of means and standard deviations
-scale_covs <- function (covs, means, sds) {
-  cols_scale <- match(names(means), colnames(covs))
-  covs_sub <- covs[, cols_scale]
-  covs_sub <- sweep(covs_sub, 2, means, "-")
-  covs_sub <- sweep(covs_sub, 2, sds, "/")
-  covs[, cols_scale] <- covs_sub
-  covs
-}
-
 surv_covs <- c("bio1", "bio12", "bio6")
 fec_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio12")
+all_covs <- unique(c(surv_covs, fec_covs))
 
 # calculate survival, given covariates
 get_survival <- function (covs, params) {
@@ -61,71 +52,8 @@ get_prob <- function (lambdas, params) {
   icloglog(params$likelihood_intercept + log(lambdas))
 }
 
-
-build_bbs_model <- function (occ, covs) {
-
-  vals <- extract(covs, occ[, c('Longitude', 'Latitude')])
-  occ <- cbind(occ, vals)
-
-  # subset, remove a text column, and convert to a matrix
-  dat <- occ[sample(seq_len(nrow(occ)), 500),-1]
-  dat <- as.matrix(dat)
-
-  all_covs <- c(surv_covs, fec_covs)
-
-  # data size
-  ncov_fecundity <- length(fec_covs)
-  ncov_survival <- length(surv_covs)
-
-  # get means and sds for scaling/rescaling
-  dat_means <- colMeans(dat[, all_covs])
-  dat_sds <- apply(dat[, all_covs], 2, sd)
-
-  # centre & scale covs to make priors meaningful
-  dat <- scale_covs(dat, dat_means, dat_sds)
-
-  # parameters
-  params <- list(
-    # sum of regression priors should have variance of 1, as a form of shrinkage
-    beta_fecundity = normal(0, 1 / ncov_fecundity, dim = ncov_fecundity),
-    beta_survival = normal(0, 1 / ncov_survival, dim = ncov_survival),
-    # no prior on the likelihood intercept
-    likelihood_intercept = variable(),
-    # uncertainty in the population means for survival and fecundity
-    default_logit_survival = normal(qlogis(0.710), sqrt(0.03806)),
-    default_log_fecundity = normal(log(0.58), 0.25)
-  )
-
-  # get all the components
-  survival <- get_survival(dat, params)
[…]

### 2019-10-02 main.R

@@ -15,6 +15,7 @@ library(raster)
 library(RColorBrewer)
 library(dplyr)
 library(maxlike)
+library(pROC)
 
 # modelling
 library(greta.dynamics)
@@ -50,11 +51,14 @@ source("R/modelling.R")
 fit_bbs_model <- drake_plan(
   bbs_occurrence = readRDS(file_in("data/clean/bbs_occ.RDS")),
   bbs_covs = brick(file_in("data/clean/bbs_covs.grd")),
-  model_list = build_bbs_model(bbs_occurrence, bbs_covs),
+  data_list = split_data(bbs_occurrence, bbs_covs),
+  model_list = build_bbs_model(data_list),
   draws_list = run_mcmc(model_list),
-  predictions = make_bbs_predictions(model_list, draws_list, bbs_covs),
+  draws_summary = check_draws(draws_list, "bbs"),
+  predictions = make_bbs_predictions(model_list, draws_list, data_list),
   plots = plot_bbs_maps(predictions, model_list),
-  stats = compare_bbs_predictions(predictions, model_list)
+  stats = compare_bbs_predictions(predictions, data_list),
+  save = saveRDS(stats, file_out("data/clean/bbs_stats.RDS"))
 )
 
 make(fit_bbs_model)
@@ -63,6 +67,7 @@ fit_protea_model <- drake_plan(
   protea_occurrence = readRDS(file_in("data/clean/protea_occurrence.RDS")),
   model_list = build_protea_model(protea_occurrence),
   draws_list = run_mcmc(model_list),
+  draws_summary = check_draws(draws_list, "protea"),
   relationships_fig = plot_protea_relationships(model_list, draws_list, protea_occurrence),
   predictions = make_protea_predictions(model_list, draws_list, protea_occurrence)
 )

### 2019-10-03 manuscript/ms.Rmd
code tweaks and compare DSDM with GLM for BBS data
rescale against MAPS survey locations yielding vital rate estimates

@@
 \widetilde{S}_A &\sim N^{(0,1)}(0.895, 0.195^2) \\ \nonumber
-\beta_{Si} &\sim N(0, 10) \\ \nonumber
-\beta_{Fi} &\sim N(0, 10)
+\beta_{Si} &\sim N(0, 1) \\ \nonumber
+\beta_{Fi} &\sim N(0, 1)
 \end{align}
+
+TODO:
+ - change this to have generic intercepts ($\alpha_F$ and $\alpha_S$), then define those intercepts below.
+ - reword the below dso that the general formulation doesn't talk about the priors for the intercepts
+ - introduce all the parts of the model here first, then have a section below this where we talk about prior ($\alpha_F \sim N(0.58, 0.25 ^ 2)$ etc.)
+ - reword the centring of X to make it clear we are centreing against the (weighted) locations where the fecundity and survival estimates were made.
 

### 2019-10-03 R/fetch_data.R

@@ -79,6 +79,61 @@ download_bbs_rasters <- function () {
   # writeRaster(covs, file = "data/clean/bbs_covs")
 }
 
+# download the station names where this species was collected to derive vital
+# rate estimates, and the coordinates of those stations
+download_maps_locations <- function () {
+
+  # get the data used to fit vital rates for the Carolina Chickadee
+  cach_url <- "https://raw.githubusercontent.com/Akcakaya/MAPS-to-Models/master/Public%20dataset/CACH_BandDataAllMonths.csv"
+  cach_data <- read.csv(cach_url)
+
+  # convert to weights (how much the inferred rates were due to that station)
+  station_counts <- table(cach_data$station)
+  station_weights <- station_counts / sum(station_counts)
+
+  # scrape the station information
+  stations_url <- "https://www.birdpop.org/pages/mapsMap.php"
+  stations_text <- readLines(stations_url)
+  stations_html <- xml2::read_html(paste(stations_text, collapse = "\n"))
+  stations_list <- xml2::as_list(stations_html)
+
+  # pull out the section of the JS code with the station data
+  js_functions <- stations_list$html$head[[11]][[1]]
+  js_function_split <- strsplit(js_functions, "\n")[[1]]
+  circles <- js_function_split[32:1317]
+
+  # parse these to find the station code and the coordinates
+  ids <- vapply(circles, get_code, "STATION ID", FUN.VALUE = character(1))
+  lats <- vapply(circles, get_code, "LATITUDE", FUN.VALUE = character(1))
+  lons <- vapply(circles, get_code, "LONGITUDE", FUN.VALUE = character(1))
+  names(ids) <- names(lats) <- names(lons) <- NULL
+
+  stations <- data.frame(
+    id = ids,
+    lat = as.numeric(lats),
+    lon = as.numeric(lons),
+    stringsAsFactors = FALSE
+  )
+
+  # subset this to the stations of interest and add the weights
+  idx <- match(names(station_weights), stations$id)
+  stations <- stations[idx, ]
+  stations$weight <- station_weights
+  stations
+
+}
+
+get_code <- function(text, field) {
+
+  start <- paste0(field, ":</span> ")
+  end <- "<br>"
+
+  text <- strsplit(text, start)[[1]][2]
+  text <- strsplit(text, end)[[1]][1]
+  text
+
+}
+
 clean_bbs <- function(file, AOU, years = 2015) {
 
[…]

### 2019-10-03 R/modelling.R

@@ -53,7 +53,7 @@ get_prob <- function (lambdas, params) {
 }
 
 # randomly split data into training and test datasets, and compute other useful data things
-split_data <- function (occ, covs) {
+split_data <- function (occ, covs, maps_coords) {
 
   # covariates for all pixels
   raster_template <- list(
@@ -61,19 +61,25 @@ split_data <- function (occ, covs) {
     idx = which(!is.na(getValues(covs[[7]])))
   )
 
-  # scale covs to have zero mean (so intercept can be defined as a spatial average) and variance 1
-  covs <- scale(covs)
+  # scale covs to have zero mean at the average maps location (so intercept can
+  # be defined as a spatial average over maps data) and variance 1 (realtive to )
+  all_vals <- getValues(covs)
+  station_vals <- extract(covs, maps_coords[, c("lon", "lat")])
+  means <- apply(station_vals, 2, weighted.mean, maps_coords$weight)
+  diffs <- sweep(all_vals, 2, means, "-")
+  sds <- sqrt(colMeans(diffs ^ 2, na.rm = TRUE))
+  all_vals_scaled <- sweep(diffs, 2, sds, "/")
+  covs <- setValues(covs, all_vals_scaled)
 
+  # extract all scaled covariate values, and combine with
   coords <- occ[, c("Longitude", "Latitude")]
   vals <- extract(covs, coords)
   occ <- occ[, -(1:2)]
   all <- cbind(occ, vals)
   all <- na.omit(all)
 
-
   # subset, remove a text column, and convert to a matrix
   train_idx <- sample.int(nrow(all), 500)
-
 
   list(
     train = all[train_idx, ],

### 2019-10-03 main.R

@@ -31,6 +31,8 @@ read_all_data <- drake_plan(
   tidy_covs = writeRaster(bbs_covs, file_out("data/clean/bbs_covs.grd"), overwrite = TRUE),
   bbs_occurrence = download_bbs(),
   tidy_occ = saveRDS(bbs_occurrence, file_out("data/clean/bbs_occ.RDS")),
+  maps_coords = download_maps_locations(),
+  tidy_maps = saveRDS(maps_coords, file_out("data/clean/maps_stations.RDS")),
 
   # protea analysis
   protea_occurrence = download_merow(),
@@ -51,14 +53,16 @@ source("R/modelling.R")
 fit_bbs_model <- drake_plan(
   bbs_occurrence = readRDS(file_in("data/clean/bbs_occ.RDS")),
   bbs_covs = brick(file_in("data/clean/bbs_covs.grd")),
-  data_list = split_data(bbs_occurrence, bbs_covs),
+  maps_coords = readRDS(file_in("data/clean/maps_stations.RDS")),
+  data_list = split_data(bbs_occurrence, bbs_covs, maps_coords),
   model_list = build_bbs_model(data_list),
   draws_list = run_mcmc(model_list),
   draws_summary = check_draws(draws_list, "bbs"),
   predictions = make_bbs_predictions(model_list, draws_list, data_list),
   plots = plot_bbs_maps(predictions, model_list),
   stats = compare_bbs_predictions(predictions, data_list),
-  save = saveRDS(stats, file_out("data/clean/bbs_stats.RDS"))
+  save = saveRDS(stats, file_out("data/clean/bbs_stats.RDS")),
+  print(stats)
 )
 
 make(fit_bbs_model)

### 2019-10-09 R/modelling.R
add analytic lambda model run

@@ -33,18 +33,30 @@ get_fecundity <- function (covs, params) {
   exp(fecundity_log)
 }
 
-get_lambda <- function(survival, fecundity) {
+get_lambda <- function(survival, fecundity, analytic = FALSE) {
 
-  # build n x m x m 3D array containing Leslie matrices for different locations
-  top_row <- cbind(fecundity * survival$juvenile,
-                   fecundity * survival$adult)
-  bottom_row <- cbind(survival$juvenile,
-                      survival$adult)
-  matrices <- abind(top_row, bottom_row, along = 3)
+  # use either the analytic solution (specific to this simple model) or the
+  # numeric solution (which is more geneeral to DSDMs)
+  if (analytic) {
 
-  # loop across observations, iterating matrices (only 10 times, for speed) to get implied intrinsic growth
-  iterated <- greta.dynamics::iterate_matrix(matrices, niter = 10)
-  iterated$lambda
+    lambda <- survival$adult * (0.94 * fecundity + 1)
+
+  } else {
+
+    # build n x m x m 3D array containing Leslie matrices for different locations
+    top_row <- cbind(fecundity * survival$juvenile,
+                     fecundity * survival$adult)
+    bottom_row <- cbind(survival$juvenile,
+                        survival$adult)
+    matrices <- abind(top_row, bottom_row, along = 3)
+
+    # loop across observations, iterating matrices to get implied intrinsic growth
+    iterated <- greta.dynamics::iterate_matrix(matrices)
+    lambda <- iterated$lambda
+
+  }
+
+  lambda
 
 }
 
@@ -114,7 +126,7 @@ build_bbs_model <- function (data_list) {
   # get all the components
   survival <- get_survival(data_list$train, params)
   fecundity <- get_fecundity(data_list$train, params)
-  lambdas <- get_lambda(survival, fecundity)
+  lambdas <- get_lambda(survival, fecundity, analytic = analytic)
   prob <- get_prob(lambdas, params)
 
   # likelihood for presence-absence data
@@ -149,7 +161,7 @@ make_bbs_predictions <- function (model_list, draws_list, data_list) {
   # get prediction greta arrays
   survival_predict <- get_survival(covs_predict, params)
   fecundity_predict <- get_fecundity(covs_predict, params)
-  lambdas_predict <- get_lambda(survival_predict, fecundity_predict)
[…]

### 2019-10-09 main.R

@@ -16,6 +16,7 @@ library(RColorBrewer)
 library(dplyr)
 library(maxlike)
 library(pROC)
+library(truncdist)
 
 # modelling
 library(greta.dynamics)
@@ -51,18 +52,30 @@ future::plan(old_plan)
 source("R/modelling.R")
 
 fit_bbs_model <- drake_plan(
+
+  # prep data
   bbs_occurrence = readRDS(file_in("data/clean/bbs_occ.RDS")),
   bbs_covs = brick(file_in("data/clean/bbs_covs.grd")),
   maps_coords = readRDS(file_in("data/clean/maps_stations.RDS")),
-  data_list = split_data(bbs_occurrence, bbs_covs, maps_coords),
-  model_list = build_bbs_model(data_list),
-  draws_list = run_mcmc(model_list),
-  draws_summary = check_draws(draws_list, "bbs"),
-  predictions = make_bbs_predictions(model_list, draws_list, data_list),
-  plots = plot_bbs_maps(predictions, model_list),
-  stats = compare_bbs_predictions(predictions, data_list),
-  save = saveRDS(stats, file_out("data/clean/bbs_stats.RDS")),
-  print(stats)
+  bbs_data_list = split_data(bbs_occurrence, bbs_covs, maps_coords),
+
+  # model fitting
+  bbs_model_list = build_bbs_model(bbs_data_list),
+  bbs_draws_list = run_mcmc(bbs_model_list),
+  bbs_draws_summary = check_draws(bbs_draws_list, "bbs"),
+
+  # also fit the analytic version
+  bbs_analytic_model_list = build_bbs_model(bbs_data_list, analytic = TRUE),
+  bbs_analytic_draws_list = run_mcmc(bbs_analytic_model_list),
+  bbs_analytic_draws_summary = check_draws(bbs_analytic_draws_list, "bbs"),
+
+  # prediction
+  bbs_predictions = make_bbs_predictions(bbs_model_list,
+                                         bbs_draws_list,
+                                         bbs_data_list),
+  bbs_plots = plot_bbs_maps(bbs_predictions, bbs_data_list),
+  bbs_stats = compare_bbs_predictions(bbs_predictions, bbs_data_list),
+
 )
 
 make(fit_bbs_model)

### 2019-10-09 manuscript/ms.Rmd
add variable gamma; use all covs in all models; use cached variables in ms

--- 2019-10-02 manuscript/ms.Rmd
@@
 
+```{r stats, echo = FALSE, cache = FALSE, eval = TRUE}
+
+# get all cached drake objects
+library(storr)
+cache <- storr_rds(drake::find_cache())
+
+# bbs objects
+
+# data size
+bbs_data <- cache$get("bbs_data_list")
+bbs_n_train <- nrow(bbs_data$train)
+bbs_n_test <- nrow(bbs_data$test)
+
+# timings
+bbs_time <- cache$get("bbs_draws_list")$timing["elapsed"]
+bbs_analytic_time <- cache$get("bbs_analytic_draws_list")$timing["elapsed"]
+
+# effective sample sizes
+bbs_draws_sry <- cache$get("bbs_draws_summary")
+bbs_neff <- round(min(bbs_draws_sry$n_eff), -3)
+
+# the R hat below which all parameter estimates fall (print 1.01 if it's less
+# than that)
+bbs_max_rhat <- max(bbs_draws_sry$r_hat$psrf[, 1])
+bbs_max_rhat <- max(bbs_max_rhat, 1.01)
+
+# format parameter estimates
+bbs_means <- bbs_draws_sry$summary$statistics[, "Mean"]
+bbs_cis <- bbs_draws_sry$summary$quantiles[, c("2.5%", "97.5%")]
+bbs_cov_names <- cache$get("all_covs")
+bbs_params <- cbind(data.frame(mean = bbs_means), bbs_cis)
+bbs_params <- bbs_params[seq_len(2 * length(bbs_cov_names)), ]
+
+text <- sprintf("%s (%s - %s)",
+                format(bbs_params$mean, digits = 2, trim = FALSE),
+                format(bbs_params$`2.5%`, digits = 1, trim = FALSE),
+                format(bbs_params$`97.5%`, digits = 1, trim = FALSE))
+
+name_lookup <- matrix(
+  c("pcMix", "mixed forest cover",
+    "pcDec", "deciduous forest cover",
+    "pcCon", "coniferous forest cover",
+    "bio1", "annual mean temperature",
+    "bio6", "min temperature of coldest month",
+    "bio12", "annual precipitation"),
+  ncol = 2, byrow = TRUE
+)
+
+bbs_coef_mat <- matrix(text, ncol = 2)
+rownames(bbs_coef_mat) <- name_lookup[match(bbs_cov_names, name_lookup[, 1]), 2]
+colnames(bbs_coef_mat) <- c("fecundity", "survival")
+
+# statistics comparing DSDM and GLM
+bbs_stats <- cache$get("bbs_stats")
+bbs_stats <- lapply(bbs_stats, round, 3)
+
+```
+
 # Introduction
@@
 <SIPMs (DDMs in Merow - note on nomenclature) apply these to large spatial scales.
-Whilst they rely on vast amnounts field data (so are not applicable in most SDM situations), they provide a link between demography and distribution, so we could perform statistical inference on them using distribution data.
-This is what we call a DSDM - the subject of this manuscript.
+Whilst they rely on vast amnounts field data (so are not applicable in most SDM situations), they provide a link between demography and distribution, making it possible to perform statistical inference on them using distribution data.
+This is what I call a DSDM - the subject of this manuscript.>
 
@@
 
-This box demonstrates the proposed approach by constructing and analysing a simple DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
-A two-stage matrix population model for the species has been developed for the USA [@ryuDevelopingPopulationModels2016], though this demographic model assumed demographic rates to be constant across the continent, rather than accounting for spatial variation in them.
+This section demonstrates the construction, fitting and analysis of a basic DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
+A two-stage matrix population model for the species has previously been developed for the USA [@ryuDevelopingPopulationModels2016], and used to carry out a population viability analysis.
+Parameters of the model - fecundity and survival rates for adults and juveniles - were estimated from mark-recapture data at a number of survey stations across the United States.
+These estimates and the resulting matrix therefore represent a single (spatial) average population over the survey stations where the species was detected.
+The model structure is:
 
-I constructed the DSDM as a single statistical model, using these national average fecundity and adult survival estimates as a prior distribution and modelling environmental variation in these rates via log- and logit-linear relationships with six environmental variables: cover of three forest types used for breeding habitat (fecundity), minimum temperature as a thermal limit (survival) and annual precipitation and mean temperature as proxies for insect abundance (fecundity and survival).
-Juvenile and adult survival differed by a constant value on the logit scale [@ryuDevelopingPopulationModels2016].
+\begin{align}
+\tilde{\mathbf{A}} = \begin{pmatrix} \widetilde{F} \cdot \widetilde{S}_J & \widetilde{F} \cdot \widetilde{S}_A \\ \widetilde{S}_J & \widetilde{S}_A \end{pmatrix}
+\end{align}
 
-For each location, these rates were used to generate a 2x2 transition matrix (a Lefkovitch matrix) and iterated to estimate the intrinsic growth rate $\lambda$.
-As in a dynamic range model, species abundance was assumed to be proportional to $\lambda$ (this simplistic assumption will be improved upon in the proposed project) and probability of presence at each location derived via the complementary log-log link, with a spatially-constant detection rate parameter.
-The model specification was completed with a Bernoulli likelihood and non-informative priors, and fitted to 500 presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
+where I use $\tilde{\mathbf{A}}$ to denote a transition matrix representing this spatially-averaged population, and $\widetilde{F}$, $\widetilde{S}_J$, and $\widetilde{S}_A$ are similarly spatially-averaged annual fecundity, juvenile survival, and adult survival rates respectively.
+Tables 2-4 in @ryuDevelopingPopulationModels2016 give the following parameter estimates, along with estimates of variability expressed here as standard deviations: $\widetilde{F}$ = 0.58 ($\pm$ 0.25), $\widetilde{S}_A$ = 0.710 ($\pm$ 0.195), $\widetilde{S}_J$ = 0.489 ($\pm$ 0.195)
 
-Inference was by Hamiltonian Monte Carlo using the R package greta and its extension package greta.dynamics for modelling dynamical systems.
-Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence ($\hat{R}\le1.01$), and yield at least X effective samples, for all model parameters.
-This took Y seconds on a laptop computer.
+Whilst these estimates provide detailed information with which to predict population trajectories for the US as a whole, they do not provide any information about how vital rates vary across the continent, and so don't help us to predict the species' distribution.
+To this end, I constructed a DSDM based on this model, but where the population in each location $i$ has a distinct set of vital rates and resulting population transition matrix:
 
-The fitted DSDM yields similar predictions of the current distribution, and has similar accuracy, to a correlative SDM (a Bayesian logistic regression fitted  by MCMC in a similar way) (Fig. 2A).
+\begin{align}
+\mathbf{A}_i = \begin{pmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{pmatrix}
+\end{align}
+
+I constructeed log-linear and logit-linear models for the fecundity and survival rates, such that they vary spatially as a function of environmental covariates, just as the probability of presence does in a correlative SDM:
+
+\begin{align}
+ln(F_i) &= \alpha_F + \mathbf{X}_i\beta_F \\ \nonumber
+logit(S_{Ai}) &= \alpha_S + \mathbf{X}_i\beta_S \\ \nonumber
+logit(S_{Ji}) &= logit(S_{Ai}) - \gamma
+\end{align}
+
+where $\mathbf{X}$ is a matrix of environmental covariates for the fecundity and survival submodels, $\alpha_F$ and $\alpha_S$ are the corresponding intercept parameters, and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
+Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale (to maintain the 0-1 constraints on survival parameters).
+
+I specify an observation model for the presence-absence data in the breeding-bird survey:
+
+\begin{align}
+y_i &\sim Bernoulli(p_i) \\ \nonumber
+p_i &= 1 - e^{-N_i} \\ \nonumber
+N_i &= \tau \lambda_i  \\ \nonumber
+\lambda_i &= eigval(\mathbf{A_i})
+\end{align}
+
+where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$.
+Imake a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure.
+This simplifying assumption has been made in other spatial population models such as dynamic range models [@pagelForecastingSpeciesRanges2012].
+As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
+The parameter $\tau$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey).
+This yields $N_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
+
+This model provides a very general DSDM structure for presence-absence data that could be applied to a wide array of species.
+I refined it for this application by encoding knowledge of the ecology of the Carolina Chickadee in prior distributions over the model parameters:
+
+\begin{align}
+exp(\alpha_F) &\sim N^+(-0.545, 0.25^2) \\ \nonumber
+logit(\alpha_S)^{-1} &\sim N^{(0,1)}(0.895, 0.195^2) \\ \nonumber
+\gamma &\sim N(0.99, 1.42) \\ \nonumber
+\beta_{S,k} &\sim N(0, 1) \\ \nonumber
+\beta_{F,k} &\sim N(0, 1) \\ \nonumber
+\tau &\sim U(0, \infty)
+\end{align}.
+
+The spatially-averaged survival and fecundity estimates from @ryuDevelopingPopulationModels2016 are used to define priors on the intercepts for the vital rate submodels.
+The means and standard deviations of the estimates define normally-distributed priors on the average fecundity and adult survival over the support of those parameters ($N^+$ and $N^{(0,1)}$ denoting truncation to positive values and to the unit interval), and these are transformed via the link function to be unconstrained.
+The logit-difference between adult and juvenile survival $\gamma$ was assigned a normal prior with parameters (estimated via Monte Carlo simulation) that approximate a $N(0.489, 0.195 ^ 2)$ prior distribution over the intercept on survival for juveniles: $logit^{-1}(\alpha_S - \gamma)$, corresponding to the juvenile survival estimate from @ryuDevelopingPopulationModels2016.
+Because these intercepts correspond to the average vital rates over locations where the species was surveyed, I centred the values in the matrix of covariate values $\mathbf{X}$ to have zero mean over the locations of the survey sites from which the recapture data were obtained that were used to estimate vital rates (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016).
+Consequently, a population in a site with the environmental conditions matching the average environmental conditions of these survey sites would have vital rates corresponding to those given by @ryuDevelopingPopulationModels2016, and spatial variation about these estimates is modelled by the regression coefficients.
+I also standardised the covariate values to have standard deviation 1 (relative to these survey station means) over the whole study area, and assigned minimally informative standard normal priors to each element $k$ in the two vectors of regression coefficients.
+In the absence of any meaningful prior information for the compound parameter $\tau$, I assigned it an improper uniform prior on all positive values, and fixed $\gamma$ - the logit-difference between adult and juvenile survival - to be the difference between the continental average estimates ($logit(0.710) - logit(0.489) = 0.94$).
+
+I fitted the model to a random subset of `r bbs_n_train` presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
+I used six environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species), minimum temperature (to capture thermal limits on survival of the species), annual precipitation and mean temperature (as proxies for insect abundance and plant growth, food sources for this species).
+The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the environmental layers were bioclim layers 6, 1, and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017].
+In general, ecological knowledge could be used to select separate covariates to use for each of the submodels.
+However, in the absence of a strong belief that any of these covariates would affect only one vital rate, all six covariates were used in both submodels.
+
+Inference was by Hamiltonian Monte Carlo using the R package greta [@goldingGretaSimpleScalable2019] and its extension package greta.dynamics [@goldingGretaDynamicsModelling2019] for modelling dynamical systems (discussed in *Computation* below).
+Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence (`r sprintf("$\\hat{R}\\le%s$", format(bbs_max_rhat, digits = 3))`), and yield at least `r round(bbs_neff)` effective samples, for all model parameters.
+This took `r round(bbs_time)` seconds on a laptop computer.
+Table 1 reports estimates of these parameters.
+
+```{r bbs_beta, echo = FALSE}
+knitr::kable(bbs_coef_mat, caption = "Posterior means and 95% credible intervals for regression coefficients in the fecundity and survival submodels ($\\beta_F$ and $\\beta_S$).
+All covariates are standardised, enabling comparison of coefficient estimates as effect sizes.")
+```
+
+The fitted DSDM yields predictions of the probability of presence like those from a correlative SDM (Fig.
+2A): predictions from a maximum likelihood logistic regression model fitted to the same data had a correlation of `r bbs_stats$glm_dsdm_correl` with the DSDM predictions, and similar AUC and correlation when compared with the true data in the `r bbs_n_test` survey locations not used to train the model (GLM: AUC = `r bbs_stats$glm_auc`, correlation = `r bbs_stats$glm_pa_correl`, DSDM: AUC = `r bbs_stats$dsdm_auc`, correlation = `r bbs_stats$dsdm_pa_correl`).
 Unlike a correlative SDM, the DSDM also predicts spatial variation in the survival (Fig. 2B) and fecundity (Fig. 2C) rates.
-The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda \ge 1$, and the model can predict the distribution under reduced fecundity (e.g.
+The species’ range limits (in this case due to the fundamental niche) can be mapped by identifying areas where $\lambda_i \ge 1$, and the model can predict the distribution under reduced fecundity (e.g.
 due to an introduced competitor) or survival (e.g.
@@
 
-```{r chickadee, echo = FALSE, fig.cap = "Figure 2.
-Spatial predictions from the fitted DSDM.
-A) Probability of presence predicted by the DSDM, with 2015 BBS presence (black dots) and absence (grey dots) records overlaid.
+```{r bbs, echo = FALSE, fig.cap = "Spatial predictions from the fitted DSDM.
+A) Probability of presence predicted by the DSDM, with 2015 BBS presence (black dots) and absence (ringed white dots) records overlaid.
 B) Spatial variation in fecundity (juveniles per adult per year).
 C) Spatial variation in annual adult survival probability.
-D) The species’ range limits (where λ≥1) under this baseline model (light blue area), a unilateral 50% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
+D) The species' range limits (where $\\lambda_i \\ge 1$) under this baseline model (light blue area), a unilateral 50\\% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
 The predictions in panels B, C and D cannot be made using a correlative SDM.", out.width = "\\textwidth"}
-# knitr::include_graphics("../figures/chickadee.png")
+knitr::include_graphics("../figures/bbs_fig.png")
 ```
+
+Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure.
+As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller.
+Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatiall-yvarying vital rates.
 
@@
 
-Biotic interactionsbetween species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single transition matrix, with inter-species density dependence [@fujiwaraCoexistenceCompetingStagestructured2011].
+Biotic interactions between species can be incorporated in a DSDM by representing the interacting dynamics of multiple species in a single transition matrix, with inter-species density dependence [@fujiwaraCoexistenceCompetingStagestructured2011].
 By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
@@
 
-Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty those predictions, it we do not have sufficient spatially-varying data.
+Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty those predictions, there is no spatially-varying information on vital rates with which to test those assumptions for this species.
+The time and cost required to robustly estimate how a species' vital rates vary with environmental covariates means that such estimates have been produced for relatively few species - one of the motivations for DSDMs.
 
@@
 
+report posterior correlations for the BBS model
+
 ##  Computation
@@
 Using greta, ecologists can easily build and fit DSDMs on desktop computers or high-performance compute clusters.
+
+Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer.
+For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = S_{Ai}(0.94 F_i + 1)$.
+Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
+Such convenient and computationally efficient analytic solutions will not be available for all DSDM models however - particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal.
+Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species.
+Consequently, the ability to efficiently analyse large numbers of population transition matrices during each iteration of the inference algorithm is critical to ensuring DSDMs are a practically useful tool for ecologists.
+
+The greta extension package greta.dynamics provides methods for efficiently analysing multiple population transition matrices to determine their intrinsic growth rates and stable stage distributions.
+The package also enables efficient analysis of dynamic transition matrices, where vital rates and therefor the matrix at each iteration depend on the previous population size or temporally-varying covariates - enabling modelling of density-dependent or temporally stochastic demographics in DSDMs.
+Whilst dispersal can already be implemented by convolving population transition matrices with a dispersal matrix, future versions of the R package will facilitate more computationally efficient algorithms for accounting for dispersal - leveraging the inherent sparsity in dispersal matrices.
 

### 2019-10-09 R/modelling.R

@@ -12,17 +12,15 @@ map_variable <- function (greta_array, draws, raster_template) {
   map
 }
 
-surv_covs <- c("bio1", "bio12", "bio6")
-fec_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio12")
-all_covs <- unique(c(surv_covs, fec_covs))
+# use all covariates in both models
+surv_covs <- fec_covs <- all_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio6", "bio12")
 
 # calculate survival, given covariates
 get_survival <- function (covs, params) {
   x_survival <- covs[, surv_covs]
   survival_logit <- x_survival %*% params$beta_survival + params$default_logit_survival
   survival_adult <- ilogit(survival_logit)
-  juvenile_offset <- qlogis(0.489) - qlogis(0.710)
-  survival_juvenile <- ilogit(survival_logit + juvenile_offset)
+  survival_juvenile <- ilogit(survival_logit - params$gamma)
   list(adult = survival_adult,
        juvenile = survival_juvenile)
 }
@@ -77,7 +75,8 @@ split_data <- function (occ, covs, maps_coords) {
   # be defined as a spatial average over maps data) and variance 1 (realtive to )
   all_vals <- getValues(covs)
   station_vals <- extract(covs, maps_coords[, c("lon", "lat")])
-  means <- apply(station_vals, 2, weighted.mean, maps_coords$weight)
+  # means <- apply(station_vals, 2, weighted.mean, maps_coords$weight)
+  means <- colMeans(station_vals)
   diffs <- sweep(all_vals, 2, means, "-")
   sds <- sqrt(colMeans(diffs ^ 2, na.rm = TRUE))
   all_vals_scaled <- sweep(diffs, 2, sds, "/")
@@ -103,21 +102,58 @@ split_data <- function (occ, covs, maps_coords) {
 
 }
 
-build_bbs_model <- function (data_list) {
+# estimate the prior for gamma, from the difference between two
+# logit-transformed truncated normal distributions
+est_gamma <- function (n = 1e5) {
+  S_A <- truncdist::rtrunc(n, "norm", 0, 1, 0.71, sqrt(0.03806))
+  S_J <- truncdist::rtrunc(n, "norm", 0, 1, 0.489, sqrt(0.03806))
+  gamma <- qlogis(S_A) - qlogis(S_J)
+  list(mean = mean(gamma),
[…]

### 2019-10-11 manuscript/ms.Rmd
explain new prior; bold 'significant' coefficients

@@
 bbs_draws_sry <- cache$get("bbs_draws_summary")
-bbs_neff <- round(min(bbs_draws_sry$n_eff), -3)
+bbs_neff <- round(min(bbs_draws_sry$n_eff), -1)
 
@@
 bbs_cis <- bbs_draws_sry$summary$quantiles[, c("2.5%", "97.5%")]
-bbs_cov_names <- cache$get("all_covs")
+bbs_cov_names <- c("intercept", cache$get("all_covs"))
 bbs_params <- cbind(data.frame(mean = bbs_means), bbs_cis)
@@
 
-text <- sprintf("%s (%s - %s)",
-                format(bbs_params$mean, digits = 2, trim = FALSE),
-                format(bbs_params$`2.5%`, digits = 1, trim = FALSE),
-                format(bbs_params$`97.5%`, digits = 1, trim = FALSE))
+mean_text <- format(round(bbs_params$mean, 3))
+ci_text <- sprintf(
+  "(%s\\ -\\ %s)",
+  format(round(bbs_params$`2.5%`, 2)),
+  format(round(bbs_params$`97.5%`, 2))
+)
+
+sig <- sign(bbs_params$`97.5%`) == sign(bbs_params$`2.5%`)
+mean_text[sig] <- paste0("\\textbf{", mean_text[sig], "}")
+text <- paste(mean_text, ci_text)
 
 name_lookup <- matrix(
-  c("pcMix", "mixed forest cover",
+  c("intercept", "intercept",
+    "pcMix", "mixed forest cover",
     "pcDec", "deciduous forest cover",
@@
 
-This section demonstrates the construction, fitting and analysis of a basic DSDM for the Carolina Chickadee *Poecile carolinensis* in the contiguous USA.
-A two-stage matrix population model for the species has previously been developed for the USA [@ryuDevelopingPopulationModels2016], and used to carry out a population viability analysis.
-Parameters of the model - fecundity and survival rates for adults and juveniles - were estimated from mark-recapture data at a number of survey stations across the United States.
-These estimates and the resulting matrix therefore represent a single (spatial) average population over the survey stations where the species was detected.
-The model structure is:
+This section demonstrates the construction, fitting and analysis of a basic DSDM the Carolina Chickadee *Poecile carolinensis*, a small passerine bird, for the contiguous USA.
+A two-stage matrix population model for the species has previously been developed [@ryuDevelopingPopulationModels2016], and used to carry out a population viability analysis.
+Fecundity and survival rates for adults and juveniles were estimated from mark-recapture data at a number of survey stations across the United States, and represent a single (spatial) average population over the survey stations where the species was detected.
+The spatially static matrix population model from this previous analysis was defined by transition matrix $\mathbf{A}$:
 
 \begin{align}
-\tilde{\mathbf{A}} = \begin{pmatrix} \widetilde{F} \cdot \widetilde{S}_J & \widetilde{F} \cdot \widetilde{S}_A \\ \widetilde{S}_J & \widetilde{S}_A \end{pmatrix}
+\mathbf{A} = \begin{pmatrix} F \cdot S_J & F \cdot S_A \\ S_J & S_A \end{pmatrix}
 \end{align}
 
-where I use $\tilde{\mathbf{A}}$ to denote a transition matrix representing this spatially-averaged population, and $\widetilde{F}$, $\widetilde{S}_J$, and $\widetilde{S}_A$ are similarly spatially-averaged annual fecundity, juvenile survival, and adult survival rates respectively.
-Tables 2-4 in @ryuDevelopingPopulationModels2016 give the following parameter estimates, along with estimates of variability expressed here as standard deviations: $\widetilde{F}$ = 0.58 ($\pm$ 0.25), $\widetilde{S}_A$ = 0.710 ($\pm$ 0.195), $\widetilde{S}_J$ = 0.489 ($\pm$ 0.195)
+where and $F$, $S_J$, and $S_A$ are single spatial estiamtes of annual fecundity, juvenile survival, and adult survival rates respectively.
+Tables 2-4 in @ryuDevelopingPopulationModels2016 give the following parameter estimates, along with estimates of variability expressed here as standard deviations: $F$ = 0.58 ($\pm$ 0.25), $S_J$ = 0.49 ($\pm$ 0.20), $S_A$ = 0.71 ($\pm$ 0.20).
 
 Whilst these estimates provide detailed information with which to predict population trajectories for the US as a whole, they do not provide any information about how vital rates vary across the continent, and so don't help us to predict the species' distribution.
-To this end, I constructed a DSDM based on this model, but where the population in each location $i$ has a distinct set of vital rates and resulting population transition matrix:
+To this end, I constructed a DSDM based on this model, but where the population in each location $i$ has a distinct set of vital rates, each given by a vital rate regression to model how vital rates vary according to environmental covariates:
 
 \begin{align}
-\mathbf{A}_i = \begin{pmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{pmatrix}
-\end{align}
-
-I constructeed log-linear and logit-linear models for the fecundity and survival rates, such that they vary spatially as a function of environmental covariates, just as the probability of presence does in a correlative SDM:
-
-\begin{align}
-ln(F_i) &= \alpha_F + \mathbf{X}_i\beta_F \\ \nonumber
-logit(S_{Ai}) &= \alpha_S + \mathbf{X}_i\beta_S \\ \nonumber
+\mathbf{A}_i &= \begin{pmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{pmatrix} \\ \nonumber
+ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
+logit(S_{Ai}) &= \mathbf{X}_i\beta_S \\ \nonumber
 logit(S_{Ji}) &= logit(S_{Ai}) - \gamma
@@
 
-where $\mathbf{X}$ is a matrix of environmental covariates for the fecundity and survival submodels, $\alpha_F$ and $\alpha_S$ are the corresponding intercept parameters, and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
+where $\mathbf{X}$ is a design matrix of environmental covariates and an intercept dummy variable for the fecundity and survival submodels and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
 Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale (to maintain the 0-1 constraints on survival parameters).
@@
 p_i &= 1 - e^{-N_i} \\ \nonumber
-N_i &= \tau \lambda_i  \\ \nonumber
+N_i &= \alpha \lambda_i  \\ \nonumber
 \lambda_i &= eigval(\mathbf{A_i})
@@
 As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
-The parameter $\tau$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey).
+The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey).
 This yields $N_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
@@
 \begin{align}
-exp(\alpha_F) &\sim N^+(-0.545, 0.25^2) \\ \nonumber
-logit(\alpha_S)^{-1} &\sim N^{(0,1)}(0.895, 0.195^2) \\ \nonumber
-\gamma &\sim N(0.99, 1.42) \\ \nonumber
-\beta_{S,k} &\sim N(0, 1) \\ \nonumber
-\beta_{F,k} &\sim N(0, 1) \\ \nonumber
-\tau &\sim U(0, \infty)
+\beta_{F} &\sim N(\mu_F, \mathbf{\Sigma}_F) \\ \nonumber
+\beta_{S} &\sim N(\mu_S, \mathbf{\Sigma}_S) \\ \nonumber
+\gamma &\sim N(1.17, 2.10) \\ \nonumber
+\alpha &\sim U(0, \infty)
 \end{align}.
 
-The spatially-averaged survival and fecundity estimates from @ryuDevelopingPopulationModels2016 are used to define priors on the intercepts for the vital rate submodels.
-The means and standard deviations of the estimates define normally-distributed priors on the average fecundity and adult survival over the support of those parameters ($N^+$ and $N^{(0,1)}$ denoting truncation to positive values and to the unit interval), and these are transformed via the link function to be unconstrained.
-The logit-difference between adult and juvenile survival $\gamma$ was assigned a normal prior with parameters (estimated via Monte Carlo simulation) that approximate a $N(0.489, 0.195 ^ 2)$ prior distribution over the intercept on survival for juveniles: $logit^{-1}(\alpha_S - \gamma)$, corresponding to the juvenile survival estimate from @ryuDevelopingPopulationModels2016.
-Because these intercepts correspond to the average vital rates over locations where the species was surveyed, I centred the values in the matrix of covariate values $\mathbf{X}$ to have zero mean over the locations of the survey sites from which the recapture data were obtained that were used to estimate vital rates (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016).
-Consequently, a population in a site with the environmental conditions matching the average environmental conditions of these survey sites would have vital rates corresponding to those given by @ryuDevelopingPopulationModels2016, and spatial variation about these estimates is modelled by the regression coefficients.
-I also standardised the covariate values to have standard deviation 1 (relative to these survey station means) over the whole study area, and assigned minimally informative standard normal priors to each element $k$ in the two vectors of regression coefficients.
-In the absence of any meaningful prior information for the compound parameter $\tau$, I assigned it an improper uniform prior on all positive values, and fixed $\gamma$ - the logit-difference between adult and juvenile survival - to be the difference between the continental average estimates ($logit(0.710) - logit(0.489) = 0.94$).
+The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels.
+These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs.
+I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{Aj}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
+First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{Aj}) \sim N(1.12, 1.28)$, and $logit(S_{Jj}) \sim N(-0.05, 0.85)$.
+These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed.
+The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$.
+The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{Aj})$: 1.12, and 1.28.
+Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
+The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
+In the absence of any meaningful prior information for the compound parameter $\alpha$, I assigned it an improper uniform prior on all positive values.
 
@@
 However, in the absence of a strong belief that any of these covariates would affect only one vital rate, all six covariates were used in both submodels.
+These covariate values were centred and scaled (before calculation of the priors on $\beta_F$ and $\beta_S$) to enable comparison of regreesion coefficients as effect sizes.
 
@@
 This took `r round(bbs_time)` seconds on a laptop computer.
-Table 1 reports estimates of these parameters.
+Table 1 reports estamited posterior distributions over these parameters.
 
-```{r bbs_beta, echo = FALSE}
-knitr::kable(bbs_coef_mat, caption = "Posterior means and 95% credible intervals for regression coefficients in the fecundity and survival submodels ($\\beta_F$ and $\\beta_S$).
-All covariates are standardised, enabling comparison of coefficient estimates as effect sizes.")
+```{r bbs_beta, echo = FALSE, cache = FALSE}
+knitr::kable(bbs_coef_mat, align = "r", caption = "Posterior means and 95% credible intervals for regression coefficients in the fecundity and survival submodels ($\\beta_F$ and $\\beta_S$).
+All covariates are standardised, enabling comparison of coefficient estimates as effect sizes.
+Posterior means are bolded where credible intervals do not overlap zero.")
 ```
@@
 Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer.
-For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = S_{Ai}(0.94 F_i + 1)$.
+For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{Ji} + S_{Ai}$.
 Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.

### 2019-10-11 R/modelling.R
implement new beta prior; fix analytic eigenvaluee

@@ -17,8 +17,8 @@ surv_covs <- fec_covs <- all_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio6"
 
 # calculate survival, given covariates
 get_survival <- function (covs, params) {
-  x_survival <- covs[, surv_covs]
-  survival_logit <- x_survival %*% params$beta_survival + params$default_logit_survival
+  x_survival <- covs[, c("intercept", surv_covs)]
+  survival_logit <- x_survival %*% params$beta_survival
   survival_adult <- ilogit(survival_logit)
   survival_juvenile <- ilogit(survival_logit - params$gamma)
   list(adult = survival_adult,
@@ -26,18 +26,18 @@ get_survival <- function (covs, params) {
 }
 
 get_fecundity <- function (covs, params) {
-  x_fecundity <- covs[, fec_covs]
-  fecundity_log <- x_fecundity %*% params$beta_fecundity + params$default_log_fecundity
+  x_fecundity <- covs[, c("intercept", fec_covs)]
+  fecundity_log <- x_fecundity %*% params$beta_fecundity
   exp(fecundity_log)
 }
 
 get_lambda <- function(survival, fecundity, analytic = FALSE) {
 
   # use either the analytic solution (specific to this simple model) or the
-  # numeric solution (which is more geneeral to DSDMs)
+  # numeric solution (which is more general to DSDMs)
   if (analytic) {
 
-    lambda <- survival$adult * (0.94 * fecundity + 1)
+    lambda <- fecundity * survival$juvenile + survival$adult
 
   } else {
 
@@ -71,20 +71,19 @@ split_data <- function (occ, covs, maps_coords) {
     idx = which(!is.na(getValues(covs[[7]])))
   )
 
-  # scale covs to have zero mean at the average maps location (so intercept can
-  # be defined as a spatial average over maps data) and variance 1 (realtive to )
-  all_vals <- getValues(covs)
-  station_vals <- extract(covs, maps_coords[, c("lon", "lat")])
-  # means <- apply(station_vals, 2, weighted.mean, maps_coords$weight)
-  means <- colMeans(station_vals)
-  diffs <- sweep(all_vals, 2, means, "-")
-  sds <- sqrt(colMeans(diffs ^ 2, na.rm = TRUE))
-  all_vals_scaled <- sweep(diffs, 2, sds, "/")
-  covs <- setValues(covs, all_vals_scaled)
+  # scale covs to have mean 0 and sd 1 over the whole study area. Handle the
[…]

### 2019-10-11 manuscript/ms.Rmd
add more text on alternative specifications

@@
 y_i &\sim Bernoulli(p_i) \\ \nonumber
-p_i &= 1 - e^{-N_i} \\ \nonumber
-N_i &= \alpha \lambda_i  \\ \nonumber
+p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
+\Lambda_i &= \alpha \lambda_i  \\ \nonumber
 \lambda_i &= eigval(\mathbf{A_i})
@@
 The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey).
-This yields $N_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
+This yields $\Lambda_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
 
@@
 This took `r round(bbs_time)` seconds on a laptop computer.
-Table 1 reports estamited posterior distributions over these parameters.
+Table 1 reports estimates of the posterior distributions over these parameters.
 
@@
 
-The fitted DSDM yields predictions of the probability of presence like those from a correlative SDM (Fig.
+Prediction of the DSM to new locations proceeds as with any Bayesian model by substituting the design matrix $\mathbf{X}$ for observed data with a design matrix corresponding to the prediction locations, and for each posterior sample of parameter values, calculating the value of the variable of interest.
+This yields a posterior distribution over predictions that incorporates uncertainty in parameter values.
+The fitted DSDM yields (posterior mean) predictions of the probability of presence similar to those from a correlative SDM (Fig.
 2A): predictions from a maximum likelihood logistic regression model fitted to the same data had a correlation of `r bbs_stats$glm_dsdm_correl` with the DSDM predictions, and similar AUC and correlation when compared with the true data in the `r bbs_n_test` survey locations not used to train the model (GLM: AUC = `r bbs_stats$glm_auc`, correlation = `r bbs_stats$glm_pa_correl`, DSDM: AUC = `r bbs_stats$dsdm_auc`, correlation = `r bbs_stats$dsdm_pa_correl`).
@@
 As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller.
-Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatiall-yvarying vital rates.
+Unlike in purely correlative models however, these assumptions are made clear by explicitly considering spatially-varying vital rates.
 
-# Extending the basic DSDM
+# Alternative DSDM specifications
 
-Basic DSDMs like the model in Example 1 do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
-Where such information is available (e.g.
-from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
+The DSDM structure given in equations 2-3 is the simplest version. It can easily be extend to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
+
+## Alternative observation models
+
+DSDMs explicitly consider the density of individuals at each location through the relative abundance variable $\Lambda_i$.
+This provides a link to the most commonly used observations for modelling species distributions in ecology.
+An observation model for count data can be specified simply by replacing the bernoulli sampling distribution in equation 3 with an appropriatecount (such as a Poisson or negative binomial distribution) with $\Lambda_i$ as the expected count.
+A model for presence-only data can be formed by considering the data as arising from a Poisson point process with $\Lambda_i$ as the intensity at location $i$ [@rennerPointProcessModels2015].
+Integration areas for each observation, and any modelled or prespecified sampling bias can be provided via $\alpha$.
+Similarly, where repeat surveys are available, a model for $\alpha$ can be used to estimate the detection probability as in an occupancy-detection analysis [@mackenzieEstimatingSiteOccupancy2002].
+
+# Temporal dynamics
+
+The example DSDM given above is temporally static: the covariate values, vital rate parameters, transition matrix and intrinsic growth rate at each site are assumed to be constant through time.
+Long-term temporal trends driven by changing environmental covariates can be accommodated in this model simply be taking the index $i$ to represent observations at distinct locations *and times*, with the observation vector and design matrix extended accordingly.
+Shorter-term temporal trends and the impacts of environmental stochasticity on species persistence can also be included by using of dynamic (temporally-varying) transition matrices:
+
+\begin{align}
+\Lambda_{t,i} &= \alpha N_{t,i}  \\ \nonumber
+N_{t,i} &= \mathbf{A(\mathbf{X}_{t,i})}N_{t-1,i}
+\end{align}
+
+where the new variable $N_{t,i}$ gives the expected population size at time $t$, which is also used as an index to environmental covariates $\mathbf{X}$ and observations.
+Rather than the intrisic growth rate computed form the static matrix, this formulation explicitly calculates population size at each time $t$ from the deterministic dynamics that define the matrix population model.
+In this formulation of the DSDM, the initial population at each site $N_{0,i}$ must be provided.
+This could either be specified as a parameter in the model, via a known initial population size, or as an arbitrary fixed value accompanied by a sufficient 'burn in' period that parameter estimates are not influenced.
+
+# Continuous stages
+
+Can use IPMs too...
+
 The DSDMs I propose are based on transition matrices (including the matrices used in both MPMs and IPMs), which are well studied in ecology and have been used to study a wide range of plant and animal species with very different life histories [@ellnerIntegralProjectionModels].
-Like other transition matrix approaches, the basic DSDMs I have already developed can also be extended to explicitly model three key ecological processes that are poorly captured by existing distribution models: density dependence, dispersal, and biotic interactions.
 
@@
 
+ add notation for A(N_t)
+ note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps (T) to get a temporally-static density-dependent model.
+In practice, T can be determined adaptively, iterating a sufficient number of times for N_i to converge and then terminating.
+
 ## Dispersal
@@
 Dispersal can be incorporated in DSDMs by representing an entire metapopulation in a single sparse, block-structured transition matrix, with dispersal represented by transition probabilities between states in different patches [@neubertDEMOGRAPHYDISPERSALCALCULATION2000].
+
+Two approaches (with equation)
+- for models using the eigenvalue convolve A with a dispersal matrix D (using a mask for dispersing stages)
+- or for density-dependent or temporally dynamic models interleave dispersal and reproduction stages
+note that in practice we would compute the eigenvalue by iteration for computational convenience.
+
+Note that the initial populations can be specified in only some patches to represent dispersal limitation.
+For example in the CaCh example setting zero initial population in the pacific north west - where there is a patch of apparently suitable habitat but where the species does not exist (presumably due to dispersal limitation).
 
@@
 By modelling demographic rates of one species as dependent on the abundance of another species, DSDMs can represent the effects of competition, facilitation, predation and parasitism.
+
+add example matrix structure as an equation
 
@@
 # Discussion
+
+# Prior specification
+
+Basic DSDMs like the model in Example 1 do not require detailed knowledge of a species population dynamics or the relationship between specific demographic rates and the environment in order to make predictions that are comparable to correlative SDMs.
+Where such information is available (e.g.
+from studies or expert opinion) it can be incorporated to improve DSDM predictions either via informative priors or by directly incorporating predictions from other models.
+
+Determining the most common and useful types of prior information, and coming up with straightforward strategies to incorporate them is an area for future research.
+
+The affine transformation provided in example 1 is one approach.
+An integrated likelihood is another option - the analysis of recapture data to inform vital rates at specific sites could be performed simultaneously with estimation of the DSDM...
 

### 2019-10-11 R/modelling.R
use formula to define design matrix

@@ -16,18 +16,16 @@ map_variable <- function (greta_array, draws, raster_template) {
 surv_covs <- fec_covs <- all_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio6", "bio12")
 
 # calculate survival, given covariates
-get_survival <- function (covs, params) {
-  x_survival <- covs[, c("intercept", surv_covs)]
-  survival_logit <- x_survival %*% params$beta_survival
+get_survival <- function (X, params) {
+  survival_logit <- X %*% params$beta_survival
   survival_adult <- ilogit(survival_logit)
   survival_juvenile <- ilogit(survival_logit - params$gamma)
   list(adult = survival_adult,
        juvenile = survival_juvenile)
 }
 
-get_fecundity <- function (covs, params) {
-  x_fecundity <- covs[, c("intercept", fec_covs)]
-  fecundity_log <- x_fecundity %*% params$beta_fecundity
+get_fecundity <- function (X, params) {
+  fecundity_log <- X %*% params$beta_fecundity
   exp(fecundity_log)
 }
 
@@ -62,10 +60,15 @@ get_prob <- function (lambdas, params) {
   icloglog(params$likelihood_intercept + log(lambdas))
 }
 
-# randomly split data into training and test datasets, and compute other useful data things
-split_data <- function (occ, covs, maps_coords) {
+get_design_matrix <- function(formula, vals) {
+  vals_df <- as.data.frame(vals)
+  model.matrix(formula, vals_df)
+}
 
-  # covariates for all pixels
+# randomly split data into training and test datasets, and compute other useful data things
+prep_data <- function (formula, occ, covs, maps_coords) {
+
+  # template raster with minimal mask and index to cells with values
   raster_template <- list(
     raster = covs[[7]] * 0,
     idx = which(!is.na(getValues(covs[[7]])))
@@ -76,28 +79,39 @@ split_data <- function (occ, covs, maps_coords) {
   covs <- scale(covs)
 
   # get covariate values for maps locations
-  maps_covs <- extract(covs, maps_coords[, c("lon", "lat")])
-  maps_covs <- cbind(intercept = 1, maps_covs)
+  maps_vals <- extract(covs, maps_coords[, c("lon", "lat")])
+  x_maps <- get_design_matrix(formula, maps_vals)
 
[…]

### 2019-10-11 main.R

@@ -53,11 +53,14 @@ source("R/modelling.R")
 
 fit_bbs_model <- drake_plan(
 
-  # prep data
+  # read in data
   bbs_occurrence = readRDS(file_in("data/clean/bbs_occ.RDS")),
   bbs_covs = brick(file_in("data/clean/bbs_covs.grd")),
   maps_coords = readRDS(file_in("data/clean/maps_stations.RDS")),
-  bbs_data_list = split_data(bbs_occurrence, bbs_covs, maps_coords),
+
+  # prepare the train/test split and design matrices
+  bbs_formula = ~ 1 + pcMix + pcDec + pcCon + bio1 + bio6 + bio12,
+  bbs_data_list = prep_data(bbs_formula, bbs_occurrence, bbs_covs, maps_coords),
 
   # model fitting
   bbs_model_list = build_bbs_model(bbs_data_list),

### 2019-10-11 manuscript/ms.Rmd
use Beverton-Holt carrying capacity model

@@
 bbs_cis <- bbs_draws_sry$summary$quantiles[, c("2.5%", "97.5%")]
-bbs_cov_names <- c("intercept", cache$get("all_covs"))
+bbs_cov_names <- colnames(bbs_data$x_train)
 bbs_params <- cbind(data.frame(mean = bbs_means), bbs_cis)
@@
 
-I specify an observation model for the presence-absence data in the breeding-bird survey:
+I specify an observation model for presence-absence data:
 
@@
 y_i &\sim Bernoulli(p_i) \\ \nonumber
-p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
-\Lambda_i &= \alpha \lambda_i  \\ \nonumber
+p_i &= 1 - e^{-\psi \Lambda_i} \\ \nonumber
+\Lambda_i &= ln(1 + e^{(\lambda_i - 1) M})  \\ \nonumber
 \lambda_i &= eigval(\mathbf{A_i})
@@
 where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$.
-Imake a statistically convenient but biologically naive assumption that the size of each population is proportional to its intrinsic growth rate - a relationship that would arise from a constant (not density-dependent) competitive pressure.
-This simplifying assumption has been made in other spatial population models such as dynamic range models [@pagelForecastingSpeciesRanges2012].
-As discussed in the text below, DSDMs can easily incorporate more complex and realistic forms of density-dependence as required.
-The parameter $\alpha$ represents the product of this constant of proportionality with the individual-level detection rate (fraction of individuals in the population detected in this year of the survey).
-This yields $\Lambda_i$, the expected number of individuals detected during the survey, which gives the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
+From the growth rate, the relative abundance of individuals $\Lambda_i$ is calculated from an expression that combined the carrying capacity expression from the Beverton-Holt model: $K_i = (\lambda_i - 1) M$, where $M$ quantifies how competition limits population sizes, transformed via the softplus function $ln(1 + e^{K_i})$ so that $\Lambda_i$ is positive for all parameter values.
+The softplus transformation is necessary since the Beverton-Holt expression would otherwise yield zero or negative relative abundances when $\lambda_i <= 1$), and can be thought of a smooth version of the Heaviside function, with $\Lambda_i$ almost identical to $K_i$ for most $lambda_i > 1$.
+This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
+The parameter $\psi$ is an individual-level detection rate so that $\psi \Lambda_i$ gives the expected number of individuals detected during the survey at observation $i$.
+This leads us to the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
 
@@
 \gamma &\sim N(1.17, 2.10) \\ \nonumber
-\alpha &\sim U(0, \infty)
+M &\sim U(0, \infty) \\ \nonumber
+\psi &\sim U(0, 1)
 \end{align}.
@@
 The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
-In the absence of any meaningful prior information for the compound parameter $\alpha$, I assigned it an improper uniform prior on all positive values.
+In the absence of any meaningful prior information for the competition effect parameter $M$ or the detection rate parameter $\psi$, I assigned the former an improper uniform prior on all positive values, and the latter a uniform prior on the unit interval.
 

### 2019-10-11 R/modelling.R

@@ -56,8 +56,31 @@ get_lambda <- function(survival, fecundity, analytic = FALSE) {
 
 }
 
-get_prob <- function (lambdas, params) {
-  icloglog(params$likelihood_intercept + log(lambdas))
+# carrying capacity under the Ricker model
+ricker_k <- function (lambda, M) {
+  lambda / M
+}
+
+# carrying capacity under the Beverton-Holt model
+bh_k <- function (lambda, M) {
+  (lambda - 1) * M
+}
+
+# calculate the carrying capacity from lambda and scaling parameters, rectified
+# to always be positive (and differentiable)
+carrying_capacity <- function (lambda, params) {
+  log1pe(bh_k(lambda, params$M))
+}
+
+# probability of observed occupancy from growth rate and ancillary parameters
+get_prob <- function (lambda, params) {
+
+  # relative abundance at each site
+  Lambda <- carrying_capacity(lambda, params)
+
+  # corresponding probability of detection
+  icloglog(log(params$psi) + log(Lambda))
+
 }
 
 get_design_matrix <- function(formula, vals) {
@@ -244,15 +267,16 @@ build_bbs_model <- function (data_list, analytic = FALSE) {
     # submodel regression coefficients
     beta_fecundity = beta_parameter(priors$beta_fecundity),
     beta_survival = beta_parameter(priors$beta_survival),
-    # no prior on the likelihood intercept
-    alpha = variable(lower = 0),
+    # no prior on the competition effect of the Beverton holt model
+    M = variable(lower = 0),
+    # uniform prior on the detection probabilities
+    psi = uniform(0, 1),
     # uncertainty in the logit-difference between adult and juvenile survival
     gamma = normal(
       priors$gamma$mean,
       priors$gamma$sd
     )
   )
-  params$likelihood_intercept <- log(params$alpha)
 
   # get all the components
   survival <- get_survival(data_list$x_train, params)
@@ -268,7 +292,8 @@ build_bbs_model <- function (data_list, analytic = FALSE) {
   m <- model(
     params$beta_fecundity,
     params$beta_survival,
-    params$likelihood_intercept,
+    params$psi,
+    params$M,
     params$gamma
   )
 

### 2019-10-11 manuscript/ms.Rmd
use probability lambda>1 as range, and tidy plot and table

@@
 name_lookup <- matrix(
-  c("intercept", "intercept",
-    "pcMix", "mixed forest cover",
-    "pcDec", "deciduous forest cover",
-    "pcCon", "coniferous forest cover",
-    "bio1", "annual mean temperature",
-    "bio6", "min temperature of coldest month",
-    "bio12", "annual precipitation"),
+  c("(Intercept)", "Intercept",
+    "pcMix", "Mixed forest cover",
+    "pcDec", "Deciduous forest cover",
+    "pcCon", "Coniferous forest cover",
+    "bio1", "Annual mean temperature",
+    "bio6", "Min. temperature of coldest month",
+    "bio12", "Annual precipitation"),
   ncol = 2, byrow = TRUE
@@
 rownames(bbs_coef_mat) <- name_lookup[match(bbs_cov_names, name_lookup[, 1]), 2]
-colnames(bbs_coef_mat) <- c("fecundity", "survival")
+colnames(bbs_coef_mat) <- c("fecundity", "adult survival")
 
@@
 A) Probability of presence predicted by the DSDM, with 2015 BBS presence (black dots) and absence (ringed white dots) records overlaid.
-B) Spatial variation in fecundity (juveniles per adult per year).
+B) Spatial variation in annual fecundity.
 C) Spatial variation in annual adult survival probability.
-D) The species' range limits (where $\\lambda_i \\ge 1$) under this baseline model (light blue area), a unilateral 50\\% reduction in fecundity (medium pink area overlaid) or in survival (dark green area overlaid).
+D) The species' range limits (where the posterior probability $\\lambda_i \\ge 1$ exceeds 0.5) under this baseline model (light blue area), a unilateral 50\\% reduction in fecundity (medium pink area overlaid) or a 25\\% reduction in survival (dark green area overlaid).
 The predictions in panels B, C and D cannot be made using a correlative SDM.", out.width = "\\textwidth"}

### 2019-10-11 R/modelling.R

@@ -318,27 +318,32 @@ make_bbs_predictions <- function (model_list, draws_list, data_list) {
   prob_predict <- get_prob(lambdas_predict, params)
 
   # map posterior mean estimates of these
-  lambda_map <- map_variable(lambdas_predict, draws, raster_template)
   prob_pres_map <- map_variable(prob_predict, draws, raster_template)
   fec_map <- map_variable(fecundity_predict, draws, raster_template)
   surv_map <- map_variable(survival_predict$adult, draws, raster_template)
 
   # decrease survival by 50% across N America
-  survival_predict_low <- lapply(survival_predict, "*", 0.5)
+  survival_predict_low <- lapply(survival_predict, "*", 0.75)
   lambdas_low_surv <- get_lambda(survival_predict_low, fecundity_predict, analytic = TRUE)
-  lambda_low_surv_map <- map_variable(lambdas_low_surv, draws, raster_template)
 
-  # decrease fecundity by 50% across N Americ
+    # decrease fecundity by 50% across N Americ
   lambdas_low_fec <- get_lambda(survival_predict, fecundity_predict * 0.5, analytic = TRUE)
-  lambda_low_fec_map <- map_variable(lambdas_low_fec, draws, raster_template)
+
+  # compute the posterior probability of lambda exceeding 1, then plot
+  in_niche <- lambdas_predict >= 1
+  in_niche_low_surv <- lambdas_low_surv >= 1
+  in_niche_low_fec <- lambdas_low_fec >= 1
+  prob_in_niche_map <- map_variable(in_niche, draws, raster_template)
+  prob_in_niche_low_surv_map <- map_variable(in_niche_low_surv, draws, raster_template)
+  prob_in_niche_low_fec_map <- map_variable(in_niche_low_fec, draws, raster_template)
 
   list(
-    lambda = lambda_map,
+    prob_in_niche = prob_in_niche_map,
     prob_present = prob_pres_map,
     fecundity = fec_map,
     survival = surv_map,
-    lambda_low_survival = lambda_low_surv_map,
-    lambda_low_fecundity = lambda_low_fec_map
+    prob_in_niche_low_survival = prob_in_niche_low_surv_map,
+    prob_in_niche_low_fecundity = prob_in_niche_low_fec_map
   )
 
 }
@@ -346,9 +351,9 @@ make_bbs_predictions <- function (model_list, draws_list, data_list) {
 plot_bbs_maps <- function (predictions, data_list) {
 
   # get range limits
[…]

### 2019-10-11 manuscript/ms.Rmd
remove nonidentified parameter and add text

@@
 
-where and $F$, $S_J$, and $S_A$ are single spatial estiamtes of annual fecundity, juvenile survival, and adult survival rates respectively.
+where and $F$, $S_J$, and $S_A$ are estimates of annual fecundity, juvenile survival, and adult survival rates respectively for the entire North American population of the species.
 Tables 2-4 in @ryuDevelopingPopulationModels2016 give the following parameter estimates, along with estimates of variability expressed here as standard deviations: $F$ = 0.58 ($\pm$ 0.25), $S_J$ = 0.49 ($\pm$ 0.20), $S_A$ = 0.71 ($\pm$ 0.20).
@@
 y_i &\sim Bernoulli(p_i) \\ \nonumber
-p_i &= 1 - e^{-\psi \Lambda_i} \\ \nonumber
-\Lambda_i &= ln(1 + e^{(\lambda_i - 1) M})  \\ \nonumber
+p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
+\Lambda_i &= ln(1 + e^{K_i})  \\ \nonumber
+K_i &= (\lambda_i - 1) \alpha \\ \nonumber
 \lambda_i &= eigval(\mathbf{A_i})
@@
 where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$.
-From the growth rate, the relative abundance of individuals $\Lambda_i$ is calculated from an expression that combined the carrying capacity expression from the Beverton-Holt model: $K_i = (\lambda_i - 1) M$, where $M$ quantifies how competition limits population sizes, transformed via the softplus function $ln(1 + e^{K_i})$ so that $\Lambda_i$ is positive for all parameter values.
-The softplus transformation is necessary since the Beverton-Holt expression would otherwise yield zero or negative relative abundances when $\lambda_i <= 1$), and can be thought of a smooth version of the Heaviside function, with $\Lambda_i$ almost identical to $K_i$ for most $lambda_i > 1$.
+The theoretical carrying capacity at each site $K_i$ is computed from the carrying capacity relationship of the Beverton-Holt model.
+$K_i$ has negative values $\lambda_i < 1$, so it is transformed to an always-positive relative abundance parameter $\Lambda_i$ via the softplus function.
+The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ almost identical to $K_i$ for most $lambda_i > 1$, rapidly approaches zero below $lambda_i = 1$.
 This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
-The parameter $\psi$ is an individual-level detection rate so that $\psi \Lambda_i$ gives the expected number of individuals detected during the survey at observation $i$.
-This leads us to the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
+The parameter $\alpha$ controls the relationship between the intrinsic population growth rate and relative abundance.
+It is a compound of two parameters that are only identified together: the Beverton-Holt density-dependence parameter relating growth rates to carrying capacities; and the detection rate - the fraction of individuals in the total population that are detected during a presence-absence survey at site $i$.
+From $\Lambda_i$, we obtain the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
 
-This model provides a very general DSDM structure for presence-absence data that could be applied to a wide array of species.
+This model provides a general DSDM structure for presence-absence data that is applicable to a range species.
 I refined it for this application by encoding knowledge of the ecology of the Carolina Chickadee in prior distributions over the model parameters:
@@
 \gamma &\sim N(1.17, 2.10) \\ \nonumber
-M &\sim U(0, \infty) \\ \nonumber
-\psi &\sim U(0, 1)
+alpha &\sim U(0, \infty)
 \end{align}.
@@
 The prior for $\gamma$, the logit-difference between adult and juvenile survival, is given by the difference between the implied normal priors on logit-survival of adults and juveniles, which is also normal.
-In the absence of any meaningful prior information for the competition effect parameter $M$ or the detection rate parameter $\psi$, I assigned the former an improper uniform prior on all positive values, and the latter a uniform prior on the unit interval.
+In the absence of any meaningful prior information for the competition effects or detection rates, I assigned $\alpha$ an improper uniform prior on all positive values.
 
@@
 I used six environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species), minimum temperature (to capture thermal limits on survival of the species), annual precipitation and mean temperature (as proxies for insect abundance and plant growth, food sources for this species).
-The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the environmental layers were bioclim layers 6, 1, and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017].
+The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the climatic covariates were bioclim layers 6, 1, and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017].
 In general, ecological knowledge could be used to select separate covariates to use for each of the submodels.
@@
 Inference was by Hamiltonian Monte Carlo using the R package greta [@goldingGretaSimpleScalable2019] and its extension package greta.dynamics [@goldingGretaDynamicsModelling2019] for modelling dynamical systems (discussed in *Computation* below).
-Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence (`r sprintf("$\\hat{R}\\le%s$", format(bbs_max_rhat, digits = 3))`), and yield at least `r round(bbs_neff)` effective samples, for all model parameters.
+Four chains were run for 1000 iterations each (after 1000 discarded iterations during which the sampler was burnt-in and tuned), which was sufficent to ensure convergence (`r sprintf("$\\hat{R}\\le%s$", format(bbs_max_rhat, digits = 3))`), and yield at least `r round(bbs_neff)` effective samples for every model parameter.
 This took `r round(bbs_time)` seconds on a laptop computer.
@@
 
-Prediction of the DSM to new locations proceeds as with any Bayesian model by substituting the design matrix $\mathbf{X}$ for observed data with a design matrix corresponding to the prediction locations, and for each posterior sample of parameter values, calculating the value of the variable of interest.
+Prediction of the DSDM to new locations proceeds as with any Bayesian model by substituting the design matrix $\mathbf{X}$ for observed data with a design matrix corresponding to the prediction locations, and for each posterior sample of parameter values, calculating the value of the variable of interest.
 This yields a posterior distribution over predictions that incorporates uncertainty in parameter values.
@@
 
-The DSDM structure given in equations 2-3 is the simplest version. It can easily be extend to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
+The DSDM structure given in equations 2-3 is the simplest version.
+It can easily be extend to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
+Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are amenable to efficient MCMC methods.
+Specifically, models with likelihoods that are analytically tractable, and can be differentiated with respect to parameter values.
+This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], thought it necessarily restricts the scope of these models to those with only continuous parameters.
+The model variations I propose below encompass a wide range of features whist remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to large datasets.
 
@@
 This provides a link to the most commonly used observations for modelling species distributions in ecology.
-An observation model for count data can be specified simply by replacing the bernoulli sampling distribution in equation 3 with an appropriatecount (such as a Poisson or negative binomial distribution) with $\Lambda_i$ as the expected count.
+An observation model for count data can be specified simply by replacing the bernoulli sampling distribution in equation 3 with an appropriate count (such as a Poisson or negative binomial distribution) with $\Lambda_i$ as the expected count.
 A model for presence-only data can be formed by considering the data as arising from a Poisson point process with $\Lambda_i$ as the intensity at location $i$ [@rennerPointProcessModels2015].
@@
 The example DSDM given above is temporally static: the covariate values, vital rate parameters, transition matrix and intrinsic growth rate at each site are assumed to be constant through time.
-Long-term temporal trends driven by changing environmental covariates can be accommodated in this model simply be taking the index $i$ to represent observations at distinct locations *and times*, with the observation vector and design matrix extended accordingly.
-Shorter-term temporal trends and the impacts of environmental stochasticity on species persistence can also be included by using of dynamic (temporally-varying) transition matrices:
+Long-term temporal trends driven by changing environmental covariates can be accommodated in this model simply by considering each time point $t$ as an independent set of observations, such that each set of environmaental conditions $\mathbf{X}_{i,t}$ results in a separate transition matrix $\mathbf{A}_{i,t}$ and corresponds to observed responses $y_{i,t}$.
+This approach assumes the time periods are sufficiently far apart that the populations have returned to the carrying capacity implied by the contemporary environmental conditions.
+
+With shorter temporal intervals, the impacts of environmental stochasticity on population sizes can be included by explicitly iterating populations through the succession of transition matrices at times $t$:
 
 \begin{align}
-\Lambda_{t,i} &= \alpha N_{t,i}  \\ \nonumber
-N_{t,i} &= \mathbf{A(\mathbf{X}_{t,i})}N_{t-1,i}
+\Lambda_{i,t} &= \alpha N_{i,t}  \\ \nonumber
+N_{i,t} &= \mathbf{A}_{i,t}N_{i,t-1}
 \end{align}
 
-where the new variable $N_{t,i}$ gives the expected population size at time $t$, which is also used as an index to environmental covariates $\mathbf{X}$ and observations.
-Rather than the intrisic growth rate computed form the static matrix, this formulation explicitly calculates population size at each time $t$ from the deterministic dynamics that define the matrix population model.
+where the new variable $N_{t,i}$ gives the expected population size at time $t$.
 In this formulation of the DSDM, the initial population at each site $N_{0,i}$ must be provided.
@@
 The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.
+this cna be achieved by iterating populations a sufficient number of time $K$ for the populations to converge on stable population sizes.
 
- add notation for A(N_t)
+\begin{align}
+\Lambda_{i} &= \alpha N_{i,K}  \\ \nonumber
+N_{i,K} &= \mathbf{A}(N_{i,k-1})N_{i,k-1}
+\end{align}
+
+for iteration $k$ in $K$.
  note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps (T) to get a temporally-static density-dependent model.
@@
 
+Include an appendix with notation and greta code for all of the model variants described above
+
 Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer.

### 2019-10-11 R/modelling.R

@@ -56,30 +56,21 @@ get_lambda <- function(survival, fecundity, analytic = FALSE) {
 
 }
 
-# carrying capacity under the Ricker model
-ricker_k <- function (lambda, M) {
-  lambda / M
-}
-
-# carrying capacity under the Beverton-Holt model
-bh_k <- function (lambda, M) {
-  (lambda - 1) * M
-}
-
 # calculate the carrying capacity from lambda and scaling parameters, rectified
 # to always be positive (and differentiable)
-carrying_capacity <- function (lambda, params) {
-  log1pe(bh_k(lambda, params$M))
-}
 
 # probability of observed occupancy from growth rate and ancillary parameters
 get_prob <- function (lambda, params) {
 
-  # relative abundance at each site
-  Lambda <- carrying_capacity(lambda, params)
+  # (relative) Beverton-Holt theoretical carrying capacity (Ricker would be
+  # lambda * alpha)
+  K <- (lambda - 1) * params$alpha
 
-  # corresponding probability of detection
-  icloglog(log(params$psi) + log(Lambda))
+  # rectify to positive relative abundance for all values
+  Lambda <- log1pe(K)
+
+  # probability of detection under Poisson sampling
+  1 - exp(-Lambda)
 
 }
 
@@ -267,10 +258,8 @@ build_bbs_model <- function (data_list, analytic = FALSE) {
     # submodel regression coefficients
     beta_fecundity = beta_parameter(priors$beta_fecundity),
     beta_survival = beta_parameter(priors$beta_survival),
-    # no prior on the competition effect of the Beverton holt model
-    M = variable(lower = 0),
-    # uniform prior on the detection probabilities
-    psi = uniform(0, 1),
+    # no prior on the combined detection/competition effect
+    alpha = variable(lower = 0),
     # uncertainty in the logit-difference between adult and juvenile survival
     gamma = normal(
       priors$gamma$mean,
@@ -292,8 +281,7 @@ build_bbs_model <- function (data_list, analytic = FALSE) {
   m <- model(
     params$beta_fecundity,
     params$beta_survival,
-    params$psi,
-    params$M,
+    params$alpha,
     params$gamma
   )
 

### 2019-10-16 manuscript/ms.Rmd
update schematic in line with new model

@@
 Statistical inference is performed to infer the correlative model parameters from the distribution data.
-B) Matrix population models (MPMs) and integral projections models (IPMs) combine either direct estimates, or correlative models, of vital rates like survival and fecundity to predict demographic quantities such as the population growth rate.
+B) Matrix population models (MPMs) and integral projection models (IPMs) combine either direct estimates, or correlative models, of vital rates like survival and fecundity to predict demographic quantities such as the population growth rate.
 Spatial IPMs apply these models to covariate values at each pixel in the landscape, and therefore map the distributions of vital rates and species.
@@
 Demographic SDMs (proposed here) combine correlative SDMs and spatial IPMs; replacing the arbitrary correlative model in a correlative SDM with an MPM or IPM, and inferring model parameters by fitting the whole model statistically to observed distributions.
-Equations correspond to the model considered in Example 1, but represent a wide range of correlative, population dynamic, and observation models that could be used in this framework.", out.width = "\\textwidth"}
+Equations correspond roughly to the model considered in Example 1, but represent a wide range of correlative, population dynamic, and observation models that could be used in this framework, with choice of function $f()$ dictating the relationship between the species' transition matrix and relative abundance.", out.width = "\\textwidth"}
 knitr::include_graphics("../figures/schematic.png")
@@
 \begin{align}
-\mathbf{A} = \begin{pmatrix} F \cdot S_J & F \cdot S_A \\ S_J & S_A \end{pmatrix}
+\mathbf{A} = \begin{bmatrix} F \cdot S_J & F \cdot S_A \\ S_J & S_A \end{bmatrix}
 \end{align}
@@
 \begin{align}
-\mathbf{A}_i &= \begin{pmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{pmatrix} \\ \nonumber
+\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{bmatrix} \\ \nonumber
 ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
@@
 where $\mathbf{X}$ is a design matrix of environmental covariates and an intercept dummy variable for the fecundity and survival submodels and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
-Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale (to maintain the 0-1 constraints on survival parameters).
+Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale - to maintain the (0,1) constraint on survival probabilities.
 
@@
 $K_i$ has negative values $\lambda_i < 1$, so it is transformed to an always-positive relative abundance parameter $\Lambda_i$ via the softplus function.
-The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ almost identical to $K_i$ for most $lambda_i > 1$, rapidly approaches zero below $lambda_i = 1$.
+The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ closely approximates $K_i$ for most $lambda_i > 1$, and rapidly approaches zero below $lambda_i = 1$.
 This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
@@
 
-The DSDM structure given in equations 2-3 is the simplest version.
-It can easily be extend to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
+The DSDM structure given in equations 2-3 is simple version for presence-absence data which makes a number of assumptions about the species ecology.
+This structure can easily be extended to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
+I detail a number of proposed variants below, and provide full notation and example code in Appendix A.
+
 Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are amenable to efficient MCMC methods.
-Specifically, models with likelihoods that are analytically tractable, and can be differentiated with respect to parameter values.
+Specifically: models with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values.
 This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], thought it necessarily restricts the scope of these models to those with only continuous parameters.
-The model variations I propose below encompass a wide range of features whist remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to large datasets.
+The model variations I propose below encompass a wide range of features whist remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.
 
@@
 This provides a link to the most commonly used observations for modelling species distributions in ecology.
-An observation model for count data can be specified simply by replacing the bernoulli sampling distribution in equation 3 with an appropriate count (such as a Poisson or negative binomial distribution) with $\Lambda_i$ as the expected count.
+An observation model for count data can be specified simply by replacing the Bernoulli sampling distribution in equation 3 with an appropriate count distribution (such as Poisson or negative binomial) and using $\Lambda_i$ as the expectation.
 A model for presence-only data can be formed by considering the data as arising from a Poisson point process with $\Lambda_i$ as the intensity at location $i$ [@rennerPointProcessModels2015].
-Integration areas for each observation, and any modelled or prespecified sampling bias can be provided via $\alpha$.
+Integration areas for each observation, and a model or prespecified site value for spatial sampling bias can be provided via $\alpha$.
 Similarly, where repeat surveys are available, a model for $\alpha$ can be used to estimate the detection probability as in an occupancy-detection analysis [@mackenzieEstimatingSiteOccupancy2002].

### 2019-10-16 manuscript/appendix_a.Rmd
start appendix with all model details

---
title: "Appendix A - Alternative DSDM formulations"
author:
  - Nick Golding
  - School of BioSciences, University of Melbourne, Parkville VIC, Australia
bibliography: dsdm.bib
csl: mee.csl
output:
  pdf_document:
  word_document:
    fig_caption: true
    highlight: null
---

```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
# set up knitr options
knitr::opts_chunk$set(message = FALSE,
               warning = FALSE,
               eval = FALSE,
               cache = FALSE)
```

This appendix provides notation and, where available, computer code for R packages greta and greta.dynamics [@goldingGretaDynamicsModelling2019; @goldingGretaSimpleScalable2019] to implement the DSDMs proposed in the manuscript *Demographic Species Distribution Models* by Nick Golding.

Note that statistical notation is written in the standard way, with the first line giving the sampling distribution, and subsequent lines defining the variables on whhich they depend, however the code uses the reverse order since in greta those variables must be defined bfore they can be used.

# Basic DSDM

For ease of reference, I repeat the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript. It presents the model in its three parts, each of which can be modified independently. Specification of priors for parameters in the model are not discussed here. See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.

## Observation model

The observation model for the basic DSDM relates the species' relative abundance (I.e. assumed to be within a multiplicative constant of the true abundance) $\Lambda_i$ at site $i$ to the probability of (observed) presence $p_i$ during a single presence-absence survey, and the presence-absence data are assumed to be independently Bernoulli-distributed following these probabilities as in a presence-absence species distribution model.

Notation:

\begin{align}
y_i &\sim Bernoulli(p_i) \\ \nonumber
p_i &= 1 - e^{-\Lambda_i}
\end{align}

Code:

```{r basic_obs}
p <- 1 - exp(-Lambda)
distribution(y) <- bernoulli(p)
```

Given vectors `Lambda` and `y` each with length `n`, the number of sites.

## Abundance relationship

The abundance relationship relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$. The intrinsic growth rate $\lambda_i$ of the population is used to compute the theoretical carrying capacity $K_i$ according to the Beverton-Holt model, with competition effect parameter $\alpha$. This theoretical carrying capacity is negative when $\lambda_i < 1$, so it is rectified by the softplus function to always yield a positive abundance $\Lambda_i$, which is close to zero when $\lambda_i < 1$, and approximately equal to $K_i$ when $\lambda_i > 1$.

Notation:

\begin{align}
\Lambda_i &= ln(1 + e^{K_i})  \\ \nonumber
K_i &= (\lambda_i - 1) \alpha \\ \nonumber
\lambda_i &= eigval(\mathbf{A_i})
\end{align}

Code:

```{r basic_abundance}
analysis <- iterate_matrix(A)
lambda <- analysis$lambda
K <- (lambda - 1) * alpha
Lambda <- log1pe(K)
```

Given a 3D array `A` with dimensions `n` x `m` x `m`, where `n` is the number of sites, and `m` the number of stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar. When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix

## Transition matrix and vital rate regressions

The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows). Thus for each adult in the previous time step $F_i \cdot S_{Ai}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{Ai}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult. For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].

Each of these vital rates at each site $i$ is given by a linear model against a design matrix $\mathbf{X}_i$ representing environmental covariates, an intercept (coded as a column of all ones), and any transformations of those covariates

Notation:

\begin{align}
\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{bmatrix} \\ \nonumber
ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
logit(S_{Ai}) &= \mathbf{X}_i\beta_S \\ \nonumber
logit(S_{Ji}) &= logit(S_{Ai}) - \gamma
\end{align}

Code:

```{r basic_transition}
log_F <- X %*% beta_F
logit_S_A <- X %*% beta_S
F <- exp(log_F)
S_A <- ilogit(logit_S_A)
S_J <- ilogit(logit_S_A - gamma)

top <- cbind(F * S_J, F * S_A)
bottom <- cbind(S_J, S_A)
A <- abind(top_row, bottom_row, along = 3)
```

Given an `n` x `k` matrix `X`, length `k` vectors `beta_S` and `beta_F`, and a scalar `gamma`, where `n` is the number of sites and `k` the number of columns (covariates plus an intercept) of the design matrix.

# References

### 2019-10-16 manuscript/appendix_a.Rmd
add observation models section appendix

@@
 
+# Alternative observation models
+
+Below I provide alternative observation models for count, and presence-only data.
+occupancy detection observation models can be implemnted using other greta extension packages, but are byond the scope of this manuscript.
+
+## Poisson count data
+
+The relative abundance variable $Lambda_i$ is the parameter of the poisson distribution, so a poisson count model is simply given by:
+
+Notation:
+\begin{align}
+y_i &\sim Poisson(\Lambda_i)
+\end{align}
+
+Code:
+```{r poisson_obs}
+distribution(y) <- poisson(Lambda)
+```
+Where `y` is a vector of counts
+
+## Negative binomial count data
+
+Where count data are overdispersed with respect to the expected counts $\Lambda_i$, this overdispersion can be modelled with a negative binomial distribution as:
+
+Notation:
+\begin{align}
+p_i &= r / (r + \Lambda_i) \\ \nonumber
+y_i &\sim Negative Binomial(r, p_i)
+\end{align}
+Where the negative binomial is parameterised in terms of dispersion parameter $r$ representing the target number of successful trials, and $p_i$ the probability of success in each trial.
+
+Code:
+```{r negbin_obs}
+p <- r / (r + Lambda)
+distribution(y) <- negative_binomial(r, p)
+```
+Given a positive scalar `r`, and where y is a vector of counts.
+
+## Presence-only data
+
+We assume presence-only data (where records consist of locations where the species was detected) arise from an a thinned, inhomogenous Poisson point process, as described in [@rennerPointProcessModels2015].
+Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed homogenous across each site $i$) that *would* be seen if there were no spatial bias in rates of reporting.
+The expected number of presence-only observations in a location $i$ is therefore given by the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\gamma_i$ (also assumed homogenous across $i$).
+Values of $\gamma_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates.
+Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\gamma_i$ is not identified in this model, so any model for $\gamma_i$ should probably not have an intercept term.
+
+Notation:
+\begin{align}
+y_i &\sim Poisson(a_i\gamma_i\Lambda_i)
+\end{align}
+
+Code:
+```{r ppm_obs}
+distribution(y) <- poisson(a * gamma * Lambda)
+```
+Where `y` gives the number of presence-only records reported at site $i$
+
+## Integrated models
+
+Specification of integrated models (those with multiple likelihoods for different datatypes) is simple, simply by using multiple `distribution()` statements.
+For example, a a combined presence-absence and abundance model might be defined as:
+
+```{r integrated_obs}
+p <- 1 - exp(-Lambda[pa_locations])
+distribution(y_pa) <- bernoulli(p)
+distribution(y_count) <- poisson(Lambda[count_locations])
+```
+Where `pa_locations` and `count_locations` are numeric indices to the sites with to presence-absence and count data respectively, and `y_pa` and `y_count` are the corresponding vectors of observed data.
+
 # References

### 2019-10-16 manuscript/ms.Rmd
add observation models section appendix

@@
 
+Since these observation models all relate to $\Lambda_i$, multiple likelihoods could be specified to form an *integrated* DSDM to be fitted to multiple datasets.
+The integration of multiple liklihoods could enable estimation of parameters such as spatial bias in reporting rates of presence-only data, or to leverage datasets which differ in their spatial and temporal extents or informativeness with respect to growth rates and vital rates.
+
 # Temporal dynamics

### 2019-10-16 manuscript/appendix_a.Rmd
add toc to appendix A; fix typos

@@
   pdf_document:
+    toc: true
+    toc_depth: 2
   word_document:
@@
 ```
+
+# Introduction
 
@@
 bottom <- cbind(S_J, S_A)
-A <- abind(top_row, bottom_row, along = 3)
+A <- abind(top, bottom, along = 3)
 ```
@@
 ```
-Given a positive scalar `r`, and where y is a vector of counts.
+Given a positive scalar `r`, and where y is a vector of counts. `r` and `p` correspond to the `size` and `prob` arguments of `stats::dnbinom` in R.
 

### 2019-10-16 manuscript/appendix_a.Rmd
patch up presence-only section

@@
 
-We assume presence-only data (where records consist of locations where the species was detected) arise from an a thinned, inhomogenous Poisson point process, as described in [@rennerPointProcessModels2015].
-Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed homogenous across each site $i$) that *would* be seen if there were no spatial bias in rates of reporting.
-The expected number of presence-only observations in a location $i$ is therefore given by the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\gamma_i$ (also assumed homogenous across $i$).
+Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogenous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015].
+The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model.
+Here I apply the piecewise constant approximation, defininthe observations $y_i$ to be counts of occurrence records in each cell $i$, having area $a_i$, where the cells represent a partition of the study area [@diggleSpatialSpatioTemporalLogGaussian2013;@simpsonGoingGridComputationally2011].
+That is they tesselate (but do not overlap) and together cover the entire study area.
+Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogenous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting.
+The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\gamma_i$ (also assumed homogenous across each cell).
 Values of $\gamma_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates.
@@
 ```
-Where `y` gives the number of presence-only records reported at site $i$
+Where `y` gives the number of presence-only records reported in cell $i$ and `a` and `gamma` are vectors giving the areas of the cells, and
 

### 2019-10-16 manuscript/appendix_a.Rmd
add abundance models section

@@
 
+# Alternative abundance relationships
+
+## Ricker model
+
+The abundance relationship in the DSDM presented in the manuscript derives from the Beverton-Holt model, with the carrying capacity $K_i$ proportional to the intrinsic growth rate $\lambda_i$ minus one.
+This ensures that $K_i$ is zero if $\lambda_i = 1$ (no growth) and increases as lambda increases, though it has the unfortunate consequence that $K_i$ has negative values for $\lambda_i < 1$, which must be rectified in that model.
+An alternative abundance relationship approach, given in [@pagelForecastingSpeciesRanges2012], derives from the Ricker model and has a linear relationship between $K_i$ and $\lambda_i$, yielding a relative abundance expression that does not need rectifying.
+Though with large enough $\alpha$, this approach can yield large values of $\Lambda_i$ even if $\lambda_i$ is well below 1, in which  case the species should not be able to persist.
+
+Notation:
+\begin{align}
+\Lambda_i &= \lambda_i \alpha \\ \nonumber
+\lambda_i &= eigval(\mathbf{A_i})
+\end{align}
+
+Code:
+```{r ricker_abundance}
+analysis <- iterate_matrix(A)
+lambda <- analysis$lambda
+Lambda <- lambda * alpha
+```
+
+## Density-dependent vital rates
+
+Both the Ricker and Beverton-Holt approaches to the abundance relationship assume that the carrying capacity can be expressed as some function of the intrinsic growth rate.
+This is convenient since the intrinsic growth rate is a well-studied and simple to compute quantity of a temporally-static transition matrix $A_i$.
+However, a more ecologically realistic approach is to explicitly model the vital rates as being density-dependent - that is having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point.
+This enables incorporation of more detailed ecological knowledge about how vital rates drive abundance, which may be important to consider if the answer to the ecological question of interest depends on density dependence.
+
+Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $m$ life stages (indexed with $j$) at site $i$, at time $t$.
+$\mathbf{n}_{i,t}$ is evolved through time by multiplication against a timepoint-specific transition matrix $\mathbf{A}_{i,t}$, contructed of timepoint-specific vital rates that depend on the total population density (i.e.
+across all life stages $m$) in the previous time point $\mathbf{N}_{i,t}$.
+For example, these density-dependent relationships could be a linear function of population density (controlled by parameters $\alpha_F$ and $\alpha_S$) on the link-scale of the vital rates.
+Since the transition matrix is no longer temporally-static, the population dynamics cannot be summarised by a population growth rate.
+Instead, iteration of the matrix using the density-dependent vital rate functions causes the vector $\mathbf{n}_{i,t}$ of population densities to converge on a static value after a sufficient number of iterations, $T$.
+This terminal density can be used directly as the relative abundance.
+As in the basic DSDM, the parameters $\alpha_F$ and $\alpha_S$ control both density-dependence and the detection rate, since these two processes cannot be differentiated without population growth data.
+
+Notation:
+\begin{align}
+\Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}}\\ \nonumber
+logit(S_{Ai,t}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}} \\ \nonumber
+logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
+\end{align}
+
+Code:
+```{r vital_dd_abundance}
+dynamic_A <- function(state, iter, log_F, logit_S, alpha_F, alpha_S, gamma) {
+
+  N <- rowSums(state)
+  log_F_t <- base_log_F + N * alpha_F
+  logit_S_A_t <- base_logit_S + N * alpha_S
+
+  F_t <- exp(log_F_t)
+  S_A_t <- ilogit(logit_S_A_t)
+  S_J_t <- ilogit(logit_S_A_t - gamma)
+
+  top <- cbind(F_t * S_J_t, F_t * S_A_t)
+  bottom <- cbind(S_J_t, S_A_t)
+  A_t <- abind(top, bottom, along = 3)
+  A_t
+
+}
+
+analysis <- iterate_dynamic_matrix(
+  matrix_function = dynamic_A, 
+  base_log_F = X %*% beta_F,
+  base_logit_S = X %*% beta_S,
+  alpha_F = alpha_F,
+  alpha_S = alpha_S,
+  gamma = gamma
+)
+
+Lambda <- analysis$stable_population
+```
+Given additional scalar parameters `alpha_F` and `alpha_S`.
+
+# Temporally-varying abundance
+
+## Long-term environmental variation
+
+## Short term environmenatal variation
+
+# Transition matrix models
+
+## Integral projection models
+
+## Multiple species
+
+## Dispersal
+
 # References

### 2019-10-16 manuscript/appendix_a.Rmd
add temporal variation section to appendix

@@
 K_i &= (\lambda_i - 1) \alpha \\ \nonumber
-\lambda_i &= eigval(\mathbf{A_i})
+\lambda_i &= eigval(\mathbf{A}_i)
 \end{align}
@@
 ```{r vital_dd_abundance}
-dynamic_A <- function(state, iter, log_F, logit_S, alpha_F, alpha_S, gamma) {
+A_t_function <- function(state, iter, log_F, logit_S, alpha_F, alpha_S, gamma) {
 
   N <- rowSums(state)
-  log_F_t <- base_log_F + N * alpha_F
-  logit_S_A_t <- base_logit_S + N * alpha_S
+  log_F_t <- base_log_F +  alpha_F * N
+  logit_S_A_t <- base_logit_S + alpha_S * N
 
@@
 analysis <- iterate_dynamic_matrix(
-  matrix_function = dynamic_A, 
+  matrix_function = A_t_function, 
   base_log_F = X %*% beta_F,
@@
 
+Given temporally-indexed response data $y_{i,t}$, temporal variation in environmental conditions and therefore vital rates can be included in a DSDM in one of two ways.
+The applicability of each of these depends on both the temporal scale of the data (the period of time betwen the observations) and the demographics of the species.
+
 ## Long-term environmental variation
 
-## Short term environmenatal variation
+If the length of time between observations is on the order of multiple generations opf the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environental conditions.
+In other words, that the population is at, or close to, the long-term carrying capacity of the species if the current environmental conditions were to be held static.
+In this situation, considering temporal change in environmental conditions is simply a matter of redifining the index $i$ in the basic DSDM to indeex unique combinations of time and site, instead of just site.
+The relative abundance parameter for each of these observations is therefore computed from the intrinsic growth rate of the transition matrix at the corresponding environmental conditions.
+
+## Short term environmental variation
+
+Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above.
+Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over all timesteps $t$ covering the timepoints for which there were observations as well as all intervening timesteps.
+Under this formulation, the population size parameters $\mathbf{n}_{i,t}$ will be affected by the initial population state $\mathbf{n}_{i,0}$.
+this can either be introduced as an unknown parameter in the model, or set to an arbitrary value, and adding a sufficient 'burn-in' period prior to the first observation such that its value has no little effect on the estimated model parameters.
+
+This model does not use static transition matrices and therefore cannot describe density dependence as a simple function of an intrinsic growth rate.
+It should therefore be combined with an explicit consideration of density-dependent vital rates as in the model above so that the population sizes $\mathbf{n}_{i,t}$ tend toward a long-term carrying capacity, rather than towards zero or infinity.
+
+Notation:
+\begin{align}
+y_{i,t} &\sim Bernoulli(p_{i,t}) \\ \nonumber
+p_{i,t} &= 1 - e^{-\Lambda_{i,t}} \\ \nonumber
+\Lambda_{i,t} &=  \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}}\\ \nonumber
+logit(S_{Ai,t}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}} \\ \nonumber
+logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
+\end{align}
+
+Note that this notation is very similar to the density-dependent vital rate model, except that both the observation model (the first three lines) and the design matrix $\mathbf{X}$ are temporally-indexed, and the relative abundances $\Lambda_{i,t}$ are given by the contemporary population sizes, not the terminal sizes.
+
+Code:
+```{r short_term_temporal}
+A_t_function <- function(state, iter, X, beta_F, beta_S, alpha_F, alpha_S, gamma) {
+
+  N <- rowSums(state)
+  log_F_t <- X %*% beta_F +  alpha_F * N
+  logit_S_A_t <- X %*% beta_S + alpha_S * N
+
+  F_t <- exp(log_F_t)
+  S_A_t <- ilogit(logit_S_A_t)
+  S_J_t <- ilogit(logit_S_A_t - gamma)
+
+  top <- cbind(F_t * S_J_t, F_t * S_A_t)
+  bottom <- cbind(S_J_t, S_A_t)
+  A_t <- abind(top, bottom, along = 3)
+  A_t
+
+}
+
+analysis <- iterate_dynamic_matrix(
+  matrix_function = A_t_function, 
+  initial_state = n_init,
+  niter = timesteps,
+  X = X_list,
+  beta_F = beta_F,
+  beta_S = beta_S,
+  alpha_F = alpha_F,
+  alpha_S = alpha_S,
+  gamma = gamma
+)
+
+Lambda <- analysis$all_states
+p <- 1 - exp(-Lambda)
+distribution(y) <- bernoulli(p)
+```
+Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is a $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.
 

### 2019-10-16 manuscript/appendix_a.Rmd
tidy up appendix; add regression model bit to main ms

@@
 
-This appendix provides notation and, where available, computer code for R packages greta and greta.dynamics [@goldingGretaDynamicsModelling2019; @goldingGretaSimpleScalable2019] to implement the DSDMs proposed in the manuscript *Demographic Species Distribution Models* by Nick Golding.
+This appendix provides statistical notation and computer code for the R packages greta and greta.dynamics [@goldingGretaDynamicsModelling2019; @goldingGretaSimpleScalable2019] to implement the broad range of DSDMs proposed in the manuscript *Demographic Species Distribution Models* by Nick Golding.
 
-Note that statistical notation is written in the standard way, with the first line giving the sampling distribution, and subsequent lines defining the variables on whhich they depend, however the code uses the reverse order since in greta those variables must be defined bfore they can be used.
+Note that statistical notation is written in the standard way with the first line giving the sampling distribution, and subsequent lines defining the variables on which previous lines depend, however the code uses the reverse order since in greta variables must be defined before they can be used to create other variables.
 
@@
 
-For ease of reference, I repeat the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript.
+For ease of reference, I repeat here the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript.
 It presents the model in its three parts, each of which can be modified independently.
@@
 
-Given vectors `Lambda` and `y` each with length `n`, the number of sites.
+Given vectors `Lambda` and `y` each with length $N$, the number of sites.
 
-## Abundance relationship
+## Abundance model
 
-The abundance relationship relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$.
+The abundance model relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$.
 The intrinsic growth rate $\lambda_i$ of the population is used to compute the theoretical carrying capacity $K_i$ according to the Beverton-Holt model, with competition effect parameter $\alpha$.
@@
 
-Given a 3D array `A` with dimensions `n` x `m` x `m`, where `n` is the number of sites, and `m` the number of stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar.
+Given a 3D array `A` with dimensions $N$x$M$x$M$, where $N$ is the number of sites, and $M$ the number of life stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar.
 When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix
 
-## Transition matrix and vital rate regressions
+## Population dynamic model
 
@@
 
-Given an `n` x `k` matrix `X`, length `k` vectors `beta_S` and `beta_F`, and a scalar `gamma`, where `n` is the number of sites and `k` the number of columns (covariates plus an intercept) of the design matrix.
+Given an $N$x$K$ matrix `X`, length $K$ vectors `beta_S` and `beta_F`, and a scalar `gamma`, where $N$ is the number of sites and $K$ the number of columns (covariates plus an intercept) of the design matrix.
 
-# Alternative observation models
+# Observation models
 
@@
 
-# Alternative abundance relationships
+# Abundance models
 
@@
 
-analysis <- iterate_dynamic_matrix(
+iterations <- iterate_dynamic_matrix(
   matrix_function = A_t_function, 
@@
 
-Lambda <- analysis$stable_population
+Lambda <- iterations$stable_population
 ```
@@
 
-# Temporally-varying abundance
+## Long-term temporal variation
 
@@
 The applicability of each of these depends on both the temporal scale of the data (the period of time betwen the observations) and the demographics of the species.
-
-## Long-term environmental variation
 
@@
 
-## Short term environmental variation
+## Short-term temporal variation
 
 Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above.
-Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over all timesteps $t$ covering the timepoints for which there were observations as well as all intervening timesteps.
+Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over timesteps $t$ corresponding to time points in the dataset (and all intermediate timesteps).
+Thus the population at each timestep will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.
+
 Under this formulation, the population size parameters $\mathbf{n}_{i,t}$ will be affected by the initial population state $\mathbf{n}_{i,0}$.
@@
 
-Lambda <- analysis$all_states
+Lambda <- iterations$all_states
 p <- 1 - exp(-Lambda)
@@
 ```
-Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is a $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.
+Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is an $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.
 
-# Transition matrix models
+# Population dynamic models
 
@@
 
-## Multiple species
+<to do, need to package up inteegration code into greta.dynamics?>
+
+## Biotic interactions
+
+<to do, just add multiple life stages>
 
@@
 
+<to do, two matrix multiplications (mention sparsity) can only be done with density-dependent vital rates>
+
 # References

### 2019-10-16 manuscript/ms.Rmd
tidy up appendix; add regression model bit to main ms

@@
 
+# Vital rate regression models
+
+The basic DSDM introduced in the manuscript only considers simple linear models of environmental variables for (appropriately transformed) vital rates.
+An advantage of specifying these models in a general-purpose statistical modelling software is that these simple models can be replaced with one of any number of statistical models, as appropriate.
+
+For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008].
+Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM.
+Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@cGretaGamGeneralised2019; @goldingGretaGpGaussian2019].
+These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables.
+Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.
+
 # Example 2. A DSDM for *Protea repens*

### 2019-10-18 manuscript/appendix_a.Rmd
add text and notation for static-transition-matrix dispersal

@@
 
-Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $m$ life stages (indexed with $j$) at site $i$, at time $t$.
+Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $M$ life stages (indexed with $j$) at site $i$, at time $t$.
 $\mathbf{n}_{i,t}$ is evolved through time by multiplication against a timepoint-specific transition matrix $\mathbf{A}_{i,t}$, contructed of timepoint-specific vital rates that depend on the total population density (i.e.
-across all life stages $m$) in the previous time point $\mathbf{N}_{i,t}$.
+across all life stages $M$) in the previous time point $\mathbf{N}_{i,t}$.
 For example, these density-dependent relationships could be a linear function of population density (controlled by parameters $\alpha_F$ and $\alpha_S$) on the link-scale of the vital rates.
@@
 \mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
-ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}}\\ \nonumber
-logit(S_{Ai,t}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}} \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
+logit(S_{Ai,t}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
 logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
@@
 \mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
-ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}}\\ \nonumber
-logit(S_{Ai,t}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^m{\mathbf{n}_{i,t,j}} \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
+logit(S_{Ai,t}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
 logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
@@
 
-## Dispersal
+Need iternatactions to be via density-dependent effects, because we can't have negative elements in the matrix?
+
+## Dispersal with a static transition matrix
+
+Dispersal of individuals between sites can be incorporated by expanding the matrix population model to simultaneously consider populations in all patches, with dispersal between patches quantified by a dispersal kernel.
+
+In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$.
+To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
+The the $i$^{th} block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this timestep.
+$d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.
+
+By connecting the populations at all sites with immigration and emigration potential, we now have a single intrinsic growth rate for the entire population rather than for th eopulation at each site.
+Rather than computing the relative abundance at each site $\Lambda_i$ as a function of an intrinsic growth rate, we can instead sum stable stage distributions over the stages in each site to yeild a stable population distribution over sites.
+$\Lambda_i$ can therefore be computed as proportional (with constant $\alpha$) to this relative population size.
+
+In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$.
+This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse.
+It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, such as cumputtion by iteration using sparse matrix multiplication.
+
+Notation:
+\begin{align}
+\Lambda_i &= \alpha v_i / \textstyle\sum_{i=1}^{N}{v_{i}} \\ \nonumber
+v_i &= \textstyle\sum_{j=1}^{M}{v^*_{j+(i-1)}} \\ \nonumber
+v^* &= eigvec(\mathbf{A}^*) \\ \nonumber
+\mathbf{A}^* &= \begin{bmatrix} \mathbf{A}_1 & \mathbf{M}_{1,2} & \dots & \mathbf{M}_{1,i'} & \dots & \mathbf{M}_{1,N}\\
+\mathbf{M}_{2,1} & \mathbf{A}_2 & \dots & \mathbf{M}_{2,i'} & \dots & \mathbf{M}_{2,N} \\
+\vdots & \vdots & \ddots & \vdots & &  \vdots \\
+\mathbf{M}_{i,1} & \mathbf{M}_{i,2} & \dots & \mathbf{A}_i & \dots & \mathbf{M}_{i,N} \\
+\vdots & \vdots &  & \vdots & \ddots & \vdots \\
+\mathbf{M}_{N,1} & \mathbf{M}_{N,2} & \dots & \mathbf{M}_{N,i'} & \dots & \mathbf{A}_N \end{bmatrix} \\ \nonumber
+\mathbf{M}_{i,i'} &= diag(d_{1,i,i'}, \dots, d_{j,i,i'}, \dots, d_{M,i,i'}) \\ \nonumber
+d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
+d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
+\end{align}
+
+Code:
+
+## Dispersal with density-dependent vital rates
 
@@
 
+Or do it with static analysis and stable size distribution for relative abundances?
+
+The order of the dispersal and population transition iterations can be reversed if pre-breeding migration is required.
+
+Notation:
+\begin{align}
+\mathbf{n}_{j,t+1} &= \mathbf{M}_j \mathbf{n}^*_{j,t} \\ \nonumber
+\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t}
+\mathbf{M}_j &= exp(-\tau_j \mathbf{D})
+\end{align}
+
+Code:
+
 # References

### 2019-10-18 manuscript/appendix_a.Rmd
add dynamic-transition-matrix dispersal; be explicit about code inputs

@@
 
+Each section below details a modelling option for a specific component of the DSDM - the observation model, aundance model, or population dynamic model.
+These components are largely interoperable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the eoclogical question at hand.
+The one exception to this is that abundance models using as input the intrinsic growth rate of the species (the Beverton-Holt model in the manuscript example, and the Ricker model introduced here) cannot be used in combination with the population dynamic models that have either dynamic transition matrices or dispersal, since these models do not yield a separate intrinsic growth rate for each site.
+With these models, the relative abundance is instead calculated directly from the dynamics, so these abundance models are not needed.
+
 Note that statistical notation is written in the standard way with the first line giving the sampling distribution, and subsequent lines defining the variables on which previous lines depend, however the code uses the reverse order since in greta variables must be defined before they can be used to create other variables.
@@
 
-Notation:
+##### Notation
 
@@
 
-Code:
+##### Code
+
+Input:
+
+- `y`: a vector of 1s and 0s encoding presence-absence observations
+- `Lambda`: a (positive) vector of the relative abundances for each observation
 
@@
 
-Code:
+##### Code
+
+Input:
+
+- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
+- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence
 
@@
 
-Code:
+##### Code
+
+Input:
+
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
+- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
+- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
 
@@
 
-Notation:
+##### Notation
+
 \begin{align}
@@
 
-Code:
+##### Code
+
+Input:
+
+- `y`: a vector of non-negative integers encoding count observations
+- `Lambda`: a (positive) vector of the relative abundances for each observation
+
 ```{r poisson_obs}
@@
 ```
-Where `y` is a vector of counts
 
@@
 \end{align}
+
 Where the negative binomial is parameterised in terms of dispersion parameter $r$ representing the target number of successful trials, and $p_i$ the probability of success in each trial.
 
-Code:
+##### Code
+
+Input:
+
+- `r`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
+- `Lambda`: a (positive) vector of the relative abundances for each observation
+- `y`: a vector of (non-negative) integers encoding count observations
+
 ```{r negbin_obs}
@@
 Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogenous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting.
-The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\gamma_i$ (also assumed homogenous across each cell).
-Values of $\gamma_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates.
-Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\gamma_i$ is not identified in this model, so any model for $\gamma_i$ should probably not have an intercept term.
+The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\tau_i$ (also assumed homogenous across each cell).
+Values of $\tau_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates.
+Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\tau_i$ is not identified in this model, so any model for $\tau_i$ should probably not have an intercept term.
 
-Notation:
+##### Notation
+
 \begin{align}
-y_i &\sim Poisson(a_i\gamma_i\Lambda_i)
+y_i &\sim Poisson(a_i\tau_i\Lambda_i)
 \end{align}
 
-Code:
+##### Code
+
+Input:
+
+- `y`: a vector of (non-negative) integers encoding the number of occurrences in each cell 
+- `a`: a vector giving the areas of each cell
+- `tau`: a (positive) vector of the relative reporting rates in each cell
+- `Lambda`: a (positive) vector of the relative abundances  in each cell
+
 ```{r ppm_obs}
-distribution(y) <- poisson(a * gamma * Lambda)
+distribution(y) <- poisson(a * tau * Lambda)
 ```
-Where `y` gives the number of presence-only records reported in cell $i$ and `a` and `gamma` are vectors giving the areas of the cells, and
 
@@
 Specification of integrated models (those with multiple likelihoods for different datatypes) is simple, simply by using multiple `distribution()` statements.
-For example, a a combined presence-absence and abundance model might be defined as:
+For example, a a combined presence-absence and abundance model might be defined as follows.
+
+##### Code
+
+Input:
+
+- `Lambda`: a vector of (positive) relative abundances for all sites, for both presence-absence and count data.
+- `y_pa`: a vector of 1s and 0s encoding presence-absence observations
+- `y_count`: a vector of (non-negative integers) encoding count observations
+- `pa_locations`: a vector of (positive) integers with the same length as `y_pa` giving the site numbers corresponding to presence-absence data
+- `count_locations`: a vector of (positive) integers with the same length as `y_count` giving the site numbers corresponding to count data
+- `alpha_count`: a (positive) scalar giving the detection rate for the count data, relative to the detection rat of the presence-absence data.
 
@@
 distribution(y_pa) <- bernoulli(p)
-distribution(y_count) <- poisson(Lambda[count_locations])
+distribution(y_count) <- poisson(alpha_count * Lambda[count_locations])
 ```
@@
 
-Code:
+##### Code
+
+Input:
+
+- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
+- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence
+
 ```{r ricker_abundance}
@@
 
-Code:
+##### Code
+
+Input:
+
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
+- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
+- `alpha_F`: a scalar for the effect of population density on log-fecundity
+- `alpha_S`: a scalar for the effect of population density on logit-survival
+- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
+
 ```{r vital_dd_abundance}
@@
 
-Code:
+##### Code
+
+Input:
+
+- `X_list`: a list of length $T$ of $N$x$K$ design matrices for each of the time points
+- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
+- `timesteps`: the value of $T$.
+- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
+- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
+- `alpha_F`: a scalar for the effect of population density on log-fecundity
+- `alpha_S`: a scalar for the effect of population density on logit-survival
+- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
+- `y`: an $N$x$T$ matrix of presence-absence data at each timestep and site
+
 ```{r short_term_temporal}
@@
 ```
-Given: `X_list` a list of length $T$ of design matrices for each of the time points, `n_init` is the initial population state, and `y` is an $N$x$T$ matrix of presence-absence at each timestep and site, and `timesteps` is the value of $T$.
 
@@
 This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse.
-It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, such as cumputtion by iteration using sparse matrix multiplication.
+It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g.
+by iteration using sparse matrix multiplication.
 
-Notation:
+#### Notation
+
 \begin{align}
@@
 
-Code:
+##### Code
 
-## Dispersal with density-dependent vital rates
+Input:
 
-<to do, two matrix multiplications (mention sparsity) can only be done with density-dependent vital rates>
+- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
+- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
+- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
+- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence
 
-Or do it with static analysis and stable size distribution for relative abundances?
+```{r static_dispersal}
+dispersal_kernel <- function (distance) {
+  exp(-tau * distance)
+}
 
-The order of the dispersal and population transition iterations can be reversed if pre-breeding migration is required.
+analysis <- iterate_matrix(
+  matrix = A,
+  distances = distance_matrix,
+  dispersal_kernel = dispersal_kernel
+)
 
-Notation:
+Lambda <- alpha * analysis$stable_abundance_distribution
+```
+
+## Dispersal with dynamic transition matrix
+
+Where the transition matrix is assumed to be dynamic - ie.
+to model the effects of environmental stochasticity, depsnity-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
+Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
+As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.
+
+I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each lifestage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
+For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
+
+The intermediate population states are computed as in earlier models by matrix multiplication of the state vectors $\mathbf{n}_{i,t}$ with the corresponding transition matrix $\mathbf{A}_i$.
+The ultimate population state for this time point is then given by matrix multiplications - separately for each of the *life stages* - of the intermediate population states against the dispersal matrix, formed from the distance matrix and dispersal function as in the dispersal model for static transition matrices.
+That is, the matrix multiplication is applied to the other dimension of the population state matrix.
+
+##### Notation
+
 \begin{align}
-\mathbf{n}_{j,t+1} &= \mathbf{M}_j \mathbf{n}^*_{j,t} \\ \nonumber
-\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t}
-\mathbf{M}_j &= exp(-\tau_j \mathbf{D})
+\mathbf{n}_{j,t+1} &= \mathbf{d}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
+\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
+d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
 \end{align}
 
-Code:
+where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilitiesfor life stage $j$, which sums to 1 across each column.
+
+##### Code
+
+Input:
+
+- `A_t_fun`: a function yielding for each time point a $N$x$M$x$M$ array of transition matrices for each site
+- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
+- `timesteps`: the number of timesteps to iterate for
+- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
+- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
+- `...`: additional named arguments to `A_t_fun`, which may be lists with different elements to be used at each time point
+
+```{r dynamic_dispersal}
+dispersal_kernel <- function (distance) {
+  exp(-tau * distance)
+}
+
+iterations <- iterate_dynamic_matrix(
+  matrix_function = A_t_function, 
+  initial_state = n_init,
+  niter = timesteps,
+  distances = distance_matrix,
+  dispersal_kernel = dispersal_kernel,
+  ...
+)
+
+Lambda <- iterations$all_states
+```
+
+Note that the final line could be replaced with: `Lambda <- iterations$stable_population` to yield the long-term stable population in the case that dispersal and density-dependent vital rates are required, but neither the effects of environmental stochasticity nor a time series of population sizes is required.
 

### 2019-10-18 manuscript/appendix_a.Rmd
add biotic interaction text and notation; double-subscript survival notation

@@
 The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows).
-Thus for each adult in the previous time step $F_i \cdot S_{Ai}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{Ai}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult.
+Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult.
 For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].
@@
 \begin{align}
-\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{Ji} & F_i \cdot S_{Ai} \\ S_{Ji} & S_{Ai}  \end{bmatrix} \\ \nonumber
+\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{J_i} & F_i \cdot S_{A_i} \\ S_{J_i} & S_{A_i}  \end{bmatrix} \\ \nonumber
 ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
-logit(S_{Ai}) &= \mathbf{X}_i\beta_S \\ \nonumber
-logit(S_{Ji}) &= logit(S_{Ai}) - \gamma
+logit(S_{A_i}) &= \mathbf{X}_i\beta_S \\ \nonumber
+logit(S_{J_i}) &= logit(S_{A_i}) - \gamma
 \end{align}
@@
 \mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
-ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
-logit(S_{Ai,t}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
-logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
+\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
+logit(S_{A_{i,t}}) &= \mathbf{X}_i \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
+logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
 \end{align}
@@
 \mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{Ji,t} & F_{i,t} \cdot S_{Ai,t} \\ S_{Ji,t} & S_{Ai,t}  \end{bmatrix}  \\ \nonumber
-ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}}\\ \nonumber
-logit(S_{Ai,t}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,t,j}} \\ \nonumber
-logit(S_{Ji,t}) &= logit(S_{Ai,t}) - \gamma
+\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
+ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
+logit(S_{A_{i,t}}) &= \mathbf{X}_{i,t} \beta_S + \alpha_S \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}} \\ \nonumber
+logit(S_{J_{i,t}}) &= logit(S_{A_{i,t}}) - \gamma
 \end{align}
@@
 
+## Species with complex demography
+
+<illustrate that any species for which a transition matrix can be formed can be modelled with a DSDM- at least to illustrate that multiple life stages can be considered.>
+
 ## Integral projection models
 
-<to do, need to package up inteegration code into greta.dynamics?>
+<to do, need to package up integration code into greta.dynamics?>
 
@@
 
-<to do, just add multiple life stages>
+Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix comprising the population transition matrices for all the species of interest in the diagonal blocks.
+For a two-species model $\mathbf{A}_i$ therefore has dimension $(M1+M2)$x$(M1+M2)$, where $M1$ and $M2$ are the number of life stages for species 1 and 2 respectively.
+Since an individual can neither change species or produce offspring of another species, the off-diagonal blocks are all zero.
+Interactions between these species can be encoded in density-dependent vital rates, with the density of both conspecific and intraspecific populations having an impact.
 
-Need iternatactions to be via density-dependent effects, because we can't have negative elements in the matrix?
+##### Notation
+
+\begin{align}
+\Lambda_{i,l} &= \mathbf{N}_{l,i,T} \\ \nonumber
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}_{i,t+1} &= \begin{bmatrix} \mathbf{A}_{1,i,t+1} & & \dots & & 0 \\
+& \ddots & & & \\
+\vdots & & \mathbf{A}_{l,i,t+1} & & \vdots \\
+& & & \ddots & \\
+0 & & \dots & & \mathbf{A}_{L,i,t+1} \end{bmatrix} \\ \nonumber
+\mathbf{A}_{l,i,t+1} &= \begin{bmatrix} F_{l,i,t} \cdot S_{J_{l,i,t}} & F_{l,i,t} \cdot S_{A_{l,i,t}} \\ S_{J_{l,i,t}} & S_{A_{l,i,t}}  \end{bmatrix}  \\ \nonumber
+ln(F_{l,i,t}) &= \mathbf{X}_i \beta_{F_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{F_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
+logit(S_{A_{l,i,t}}) &= \mathbf{X}_i \beta_{S_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{S_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
+logit(S_{J_{l,i,t}}) &= logit(S_{A_{l,i,t}}) - \gamma_l \\ \nonumber
+\mathbf{N}_{l,i,t} &= \textstyle\sum_{j=1+(l - 1)}^{lM}{\mathbf{n}_{i,j,t}}
+\end{align}
+
+##### Code
+
+Input:
+
+```{r biotic_interactions}
+
+```
+
+## Evolutionary dynamics
+
+As in the biotic interactions, model different genotypes with a separate set of life stages, but fecundity in off-diagonal blocks gives the probability of the offspring having a different genotype to the parent.
+Different genotypes can therefore have different vital rates.
+
+How to get separate abundances? intrinsic growth rate-derived abundance, multiplied by relative abundance of each of the genotypes.
+
+##### Notation
+
+\begin{align}
+a
+\end{align}
+
+##### Code
+
+Input:
+
+```{r evolutionary_dynamics}
+
+```
 

### 2019-10-18 manuscript/ms.Rmd
add biotic interaction text and notation; double-subscript survival notation

@@
 These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs.
-I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{Aj}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
-First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{Aj}) \sim N(1.12, 1.28)$, and $logit(S_{Jj}) \sim N(-0.05, 0.85)$.
+I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{A_j}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
+First, the means and variances of these estimates were used to calculate equivalent parameters of univariate normal distributions on the link-scale of the regression submodels, resulting in the implied priors: $ln(F_j) \sim N(-0.63, 0.18)$, $logit(S_{A_j}) \sim N(1.12, 1.28)$, and $logit(S_{J_j}) \sim N(-0.05, 0.85)$.
 These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed.
 The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$.
-The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{Aj})$: 1.12, and 1.28.
+The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{A_j})$: 1.12, and 1.28.
 Note that these priors do not provide any information on the direction of the relationships between vital rates and environmental covariates (the intercept elements of $\mu_F$ and $\mu_S$ are informed, but the prior mean for the other terms is $\approx 0$), but by providing information on plausible magnitudes of, and correlations between these parameters the prior favours regression equations that pass close to the vital rate estimates at $F_j$ and $S_j$.
@@
 Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM.
-Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@cGretaGamGeneralised2019; @goldingGretaGpGaussian2019].
+Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@goldingGretaGamGeneralised2019; @goldingGretaGpGaussian2019].
 These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables.
@@
 Bayesian inference on the DSDM for the Carolina Chickadee by MCMC took `r round(bbs_time)` seconds on a laptop computer.
-For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{Ji} + S_{Ai}$.
+For this particular model, the computation time can be sped up by using an analytic solution to the transition matrix eigenvalues: $\lambda_i = F_i S_{J_i} + S_{A_i}$.
 Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.

### 2019-10-18 manuscript/appendix_a.Rmd
add biotic interaction model

@@
 
-Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix comprising the population transition matrices for all the species of interest in the diagonal blocks.
-For a two-species model $\mathbf{A}_i$ therefore has dimension $(M1+M2)$x$(M1+M2)$, where $M1$ and $M2$ are the number of life stages for species 1 and 2 respectively.
-Since an individual can neither change species or produce offspring of another species, the off-diagonal blocks are all zero.
-Interactions between these species can be encoded in density-dependent vital rates, with the density of both conspecific and intraspecific populations having an impact.
+Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix that has the population transition matrices for each species on the diagonal.
+For model of $L$ species each with $M$ lifestages $\mathbf{A}_i$ therefore has dimension $LM$x$LM$.
+The off-diagonal blocks are all zero-matrices, since an individual can neither change species nor produce offspring of another species.
+Interactions between species can be encoded via density-dependent vital rates, with the density of both conspecific and intraspecific populations affecting the vital rates of each species.
+
+As in the single-species density-dependent vital rate model above the relative population density for each species is computed by iteration; at each timestep recalculating the transition matrix from population densities of all species and matrix-multiplying it by the current population density vector to obtain the subsequent population density vector.
+The transition matrix $\mathbf{A}_i$ is sparse; with only 1 in $L$ elements being non-zero.
+This iteration can therefore be carried out efficiently with sparse matrix-multiplication operations.
 
@@
 0 & & \dots & & \mathbf{A}_{L,i,t+1} \end{bmatrix} \\ \nonumber
-\mathbf{A}_{l,i,t+1} &= \begin{bmatrix} F_{l,i,t} \cdot S_{J_{l,i,t}} & F_{l,i,t} \cdot S_{A_{l,i,t}} \\ S_{J_{l,i,t}} & S_{A_{l,i,t}}  \end{bmatrix}  \\ \nonumber
-ln(F_{l,i,t}) &= \mathbf{X}_i \beta_{F_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{F_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
-logit(S_{A_{l,i,t}}) &= \mathbf{X}_i \beta_{S_l} + \textstyle\sum_{l'=1}^{L}{\alpha_{S_{l,l'}} \mathbf{N}_{l',i,t}}\\ \nonumber
-logit(S_{J_{l,i,t}}) &= logit(S_{A_{l,i,t}}) - \gamma_l \\ \nonumber
+\mathbf{A}_{l,i,t+1} &= \begin{bmatrix} \mathbf{F}_{l,i,t} \cdot \mathbf{S}_{J_{l,i,t}} & \mathbf{F}_{l,i,t} \cdot \mathbf{S}_{A_{l,i,t}} \\ \mathbf{S}_{J_{l,i,t}} & \mathbf{S}_{A_{l,i,t}}  \end{bmatrix}  \\ \nonumber
+ln(\mathbf{F}_{t}) &= \mathbf{X} \beta_F + \mathbf{N}_{t} \mathbf{\alpha}_F \\ \nonumber
+logit(\mathbf{S}_{A_{t}}) &= \mathbf{X} \beta_S + \mathbf{N}_{t} \mathbf{\alpha}_S \\ \nonumber
+logit(\mathbf{S}_{J_{l,i,t}}) &= logit(\mathbf{S}_{A_{l,i,t}}) - \gamma_l \\ \nonumber
 \mathbf{N}_{l,i,t} &= \textstyle\sum_{j=1+(l - 1)}^{lM}{\mathbf{n}_{i,j,t}}
 \end{align}
+
+where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are 3D arrays with dimension $L$x$N$x$T$ giving the fecundity, adult and juvenile survival and relative population density of each species $l$ at site $i$ at time $t$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ or regression coefficients for the submodels for each of species, and $L$x$L$ matrices $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$ of density-dependence effects between specie.
+I.e.
+each element $\mathbf{\alpha}_{F_{l,l'}}$ gives the effect of the density of the population of species $l'$ on the log-fecundity of species $l$.
+Off-diagonal elements of $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$  quantify fecundity- and survival-mediated interactions between species, whilst the diagonal elements quantify conspecific density-dependence.
 
@@
 Input:
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a $K$x$L$ matrix of regression coefficients for log-fecundity for each species
+- `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each species
+- `alpha_F`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on log-fecundity
+- `alpha_S`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on logit-survival
+- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival betwen adults and juveniles for each species
 
 ```{r biotic_interactions}
+# matrix of weights for each lifestage to the population density of each species
+# (for density-dependence effects)
 
+A_t_function <- function(state, iter, log_F, logit_S, alpha_F, alpha_S, gamma, dd_weights) {
+
+  # apply density-dependent effects
+  N <- state %*% dd_weights
+  log_F_t <-  base_log_F + N %*% alpha_F
+  logit_S_A_t <- base_logit_S + N %*% alpha_S
+
+  # get vital rates for each site/species
+  F_t <- exp(log_F_t)
+  S_A_t <- ilogit(logit_S_A_t)
+  S_J_t <- ilogit(sweep(logit_S_A, 2, gamma, "-"))
+
+  # combine into a single matrix such that that each species has adjacent lifestages
+  top <- rbind(F_t * S_J_t, F_t * S_A_t)
+  bottom <- rbind(S_J_t, S_A_t)
+  dim(top) <- c(nrow(N), ncol(N) * 2)
+  dim(bottom) <- c(nrow(N), ncol(N) * 2)
+  A_t <- abind(top, bottom, along = 3)
+  A_t
+
+}
+
+# weights matrix to aggregate stage-states for each species
+L <- length(gamma)
+dd_weights <- diag(L)[rep(1:L, each = M), ]
+
+iterations <- iterate_dynamic_matrix(
+  matrix_function = A_t_function, 
+  base_log_F = X %*% beta_F,
+  base_logit_S = X %*% beta_S,
+  alpha_F = alpha_F,
+  alpha_S = alpha_S,
+  gamma = gamma,
+  dd_weights = dd_weights
+)
+
+Lambda <- iterations$stable_population
 ```
@@
 It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g.
-by iteration using sparse matrix multiplication.
+by iteration using sparse matrix multiplication, as in the biotic interaction model.
 

### 2019-10-18 manuscript/appendix_a.Rmd
--amend

@@
 Input:
+
 - `X`: a $N$x$K$ design matrix for vital rate regression sub-models
@@
 ```{r biotic_interactions}
-# matrix of weights for each lifestage to the population density of each species
-# (for density-dependence effects)
-
 A_t_function <- function(state, iter, log_F, logit_S, alpha_F, alpha_S, gamma, dd_weights) {
 
-  # apply density-dependent effects
   N <- state %*% dd_weights
@@
 
-  # get vital rates for each site/species
   F_t <- exp(log_F_t)
@@
 
-  # combine into a single matrix such that that each species has adjacent lifestages
   top <- rbind(F_t * S_J_t, F_t * S_A_t)

### 2019-10-22 manuscript/appendix_a.Rmd
add evolutionary dynamics model

@@
 \Lambda_{i,l} &= \mathbf{N}_{l,i,T} \\ \nonumber
-\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{A}_{i,t+1} &= \begin{bmatrix} \mathbf{A}_{1,i,t+1} & & \dots & & 0 \\
+\mathbf{n}_{i,t+1} &= \mathbf{A}^*_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}^*_{i,t+1} &= \begin{bmatrix} \mathbf{A}_{1,i,t+1} & & \dots & & 0 \\
 & \ddots & & & \\
@@
 
-As in the biotic interactions, model different genotypes with a separate set of life stages, but fecundity in off-diagonal blocks gives the probability of the offspring having a different genotype to the parent.
-Different genotypes can therefore have different vital rates.
+Similarly to the biotic interactions model, we can consider a population of single species but with $L$ different genotypes with a blocked transition matrix.
+In this model, the genotypes are linked via reproduction rather than density-dependent vital rates.
+tThe $L$x$L$ matrix $\mathbf{m}$ quantifies mutation rates between different genotypes.
+For an individual of genotype $l$, each of its offspring has probability $\mathbf{m}_{l,l'}$ of being of genotype $l'$.
+Each genotype has its own vital rate submodels, representing a linked phenotype driving population dynamics for that genotype.
+By analysing the combined transition matrix $\mathbf{A}^*_i$ at each site $i$ with different environmental covariates, the model can consider how selection for each genotype varies across the environment, and predict the genetic makeup of populations.
 
-How to get separate abundances? intrinsic growth rate-derived abundance, multiplied by relative abundance of each of the genotypes.
+This model is compatible with both static and dynamic transition matrices.
+In the static case, the expected abundance $\Lambda_{l,i}$ of each genotype $l$ at each site $i$ can computed by the product of the expected total abundance $\Lambda^*_i$ (computed via an abundance model as in the basic model) and the stable distribution of genotypes, calculated by summing the dominant eigenvector over the stages in each genotype and standardising to sum to one.
+In the dynamic case relative abundances can be calculated in the same way as for the biotic interaction model.
 
@@
 \begin{align}
-a
+\Lambda_{l,i} &= \alpha \Lambda^*_i v_{l,i} / \textstyle\sum_{l=1}^{L}{v_{l,i}} \\ \nonumber
+\Lambda^*_i &= ln(1 + e^{K_i})  \\ \nonumber
+K_i &= (\lambda_i - 1) \alpha \\ \nonumber
+v_{l,i} &= \textstyle\sum_{j=1}^{M}{v^*_{j+(l-1),i}} \\ \nonumber
+v^*_i &= eigvec(\mathbf{A}^*_i) \\ \nonumber
+\lambda_i &= eigval(\mathbf{A}^*_i) \\ \nonumber
+\mathbf{A}^*_i &= \begin{bmatrix} \mathbf{A}_{1,i} & \mathbf{F}_{1,2,i} & \dots & \mathbf{F}_{1,l',i} & \dots & \mathbf{F}_{1,L,i}\\
+\mathbf{F}_{2,1,i} & \mathbf{A}_{2,i} & \dots & \mathbf{F}_{2,l',i} & \dots & \mathbf{F}_{2,L,i} \\
+\vdots & \vdots & \ddots & \vdots & &  \vdots \\
+\mathbf{F}_{l,1,i} & \mathbf{F}_{l,2,i} & \dots & \mathbf{A}_{l,i} & \dots & \mathbf{F}_{l,L,i} \\
+\vdots & \vdots &  & \vdots & \ddots & \vdots \\
+\mathbf{F}_{L,1,i} & \mathbf{F}_{L,2,i} & \dots & \mathbf{F}_{L,l',i} & \dots & \mathbf{A}_{L,i} \end{bmatrix} \\ \nonumber
+\mathbf{F}_{l,l',i} &= \begin{bmatrix} \mathbf{m}_{l,l'} \cdot F_{l',i} \cdot S_{J_{l',i}} & \mathbf{m}_{l,l'} \cdot F_{l',i} \cdot S_{A_{l',i}} \\ 0 & 0 \end{bmatrix} \\ \nonumber
+\mathbf{A}_{l,i} &= \begin{bmatrix} \mathbf{m}_{l,l} \cdot F_{l,i} \cdot S_{J_{l,i}} & \mathbf{m}_{l,l} \cdot F_{l,i} \cdot S_{A_{l,i}} \\ S_{J_{l,i}} & S_{A_{l,i}} \end{bmatrix}  \\ \nonumber
+ln(\mathbf{F}) &= \mathbf{X} \beta_F \\ \nonumber
+logit(\mathbf{S}_{A}) &= \mathbf{X} \beta_S \\ \nonumber
+logit(\mathbf{S}_{J_{l,i}}) &= logit(\mathbf{S}_{A_{l,i}}) - \gamma_l
 \end{align}
+
+where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are $L$x$N$ matrices giving the fecundity, adult and juvenile survival and relative population density of each genotype $l$ at site $i$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ of regression coefficients for the submodels for each of species.
+$\mathbf{m}$ is an $L$x$L$ matrix of positive values with columns summing to one, giving the probabilities of a parent of one genotype producing offspring of each genotype.
+
+An equivalent notation for $\mathbf{A}^*_i$ using Kronecker products of matrices is as follows:
+
+\begin{align}
+\mathbf{A}^*_i &=  (\mathbf{1} \otimes \mathbf{R}_i^T) \odot (\mathbf{m} \otimes \mathbf{B} + \mathbf{I} \otimes \mathbf{C})\\ \nonumber 
+\mathbf{R}_i &= \mathbf{F}_i \otimes \mathbf{B}^T \odot (\mathbf{S}_{J_i} \otimes \mathbf{B} + \mathbf{S}_{A_i} \otimes \mathbf{C})
+\end{align}
+
+where $\mathbf{B}$ is an $M$x$M$ matrix with ones on the first row and zeros elsewhere (ie.
+a mask indicating cells representing fecundity rates), $\mathbf{C}$ is its reverse (zeros on the first row, ones elsewhere) $\mathbf{1}$ is length $L$ vector of ones, $\mathbf{I}$ is an identity matrix of dimension $L$, and $\odot$ and $\otimes$ represent the Hadamard (elementwise) and Kronecker products respectively.
+
+This is the form that we use in the code below, vectorised across all $N$ sites, since it is more computationally efficient to carry out the calculations in this way in greta.
 

### 2019-10-22 manuscript/appendix_a.Rmd
add code for evolutionary model; reorder models; spell check

@@
 
-Each section below details a modelling option for a specific component of the DSDM - the observation model, aundance model, or population dynamic model.
-These components are largely interoperable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the eoclogical question at hand.
-The one exception to this is that abundance models using as input the intrinsic growth rate of the species (the Beverton-Holt model in the manuscript example, and the Ricker model introduced here) cannot be used in combination with the population dynamic models that have either dynamic transition matrices or dispersal, since these models do not yield a separate intrinsic growth rate for each site.
-With these models, the relative abundance is instead calculated directly from the dynamics, so these abundance models are not needed.
+Each section below details a modelling option for a specific component of the DSDM - the observation model, abundance model, or population dynamic model.
+These components are largely inter-operable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the ecological question at hand.
+The one exception to this is that abundance models using as input the intrinsic growth rate of the species (the Beverton-Holt model in the manuscript example, and the Ricker model introduced here) cannot be used in combination with the population dynamic models that have either dynamic transition matrices or dispersal, since those population dynamic models do not yield a separate intrinsic growth rate for each site.
+With those models, the relative abundance is instead calculated directly from the dynamics, so these abundance models are not needed.
+
+This appendix only considers the structures of the proposed models, specification of priors for parameters in the models are not discussed here.
+See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.
+I reiterate here that the aim of DSDMs is to incorporate ecological knowledge via informative priors, and the parameters of the proposed models are unlikely to be identified with uninformative priors.
 
@@
 It presents the model in its three parts, each of which can be modified independently.
-Specification of priors for parameters in the model are not discussed here.
-See [@goldingGretaSimpleScalable2019] and the website [greta-stats.org](https://greta-stats.org) for general guidance on how to build models with greta.
-
 ## Observation model
@@
 
-Given vectors `Lambda` and `y` each with length $N$, the number of sites.
-
 ## Abundance model
@@
 - `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
-- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence
+- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
 
@@
 
-Given a 3D array `A` with dimensions $N$x$M$x$M$, where $N$ is the number of sites, and $M$ the number of life stages in the population model (dimension of the matrices $\mathbf{A}_i$), and `alpha` is a scalar.
-When used with its default settings the function `greta.dynamics::iterate_matrix()` returns, among other quantities, a vector of intrinsic growth rates of the matrices in `A` computed by repeated iteration of the matrix
-
 ## Population dynamic model
 
-The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous timestep (columns) to each stage in the present time step (rows).
+The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous time step (columns) to each stage in the present time step (rows).
 Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult.
@@
 - `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
-- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
+- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
 
@@
 
-Given an $N$x$K$ matrix `X`, length $K$ vectors `beta_S` and `beta_F`, and a scalar `gamma`, where $N$ is the number of sites and $K$ the number of columns (covariates plus an intercept) of the design matrix.
-
 # Observation models
@@
 Below I provide alternative observation models for count, and presence-only data.
-occupancy detection observation models can be implemnted using other greta extension packages, but are byond the scope of this manuscript.
+occupancy detection observation models can be implemented using other greta extension packages, but are beyond the scope of this manuscript.
 
@@
 
-The relative abundance variable $Lambda_i$ is the parameter of the poisson distribution, so a poisson count model is simply given by:
+The relative abundance variable $Lambda_i$ is the parameter of the Poisson distribution, so a Poisson count model is simply given by:
 
@@
 
-Where count data are overdispersed with respect to the expected counts $\Lambda_i$, this overdispersion can be modelled with a negative binomial distribution as:
+Where count data are over-dispersed with respect to the expected counts $\Lambda_i$, this over-dispersion can be modelled with a negative binomial distribution as:
 
@@
 
-- `r`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
+- `size`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
 - `Lambda`: a (positive) vector of the relative abundances for each observation
@@
 ```{r negbin_obs}
-p <- r / (r + Lambda)
-distribution(y) <- negative_binomial(r, p)
+prob <- size / (size + Lambda)
+distribution(y) <- negative_binomial(size, prob)
 ```
-Given a positive scalar `r`, and where y is a vector of counts. `r` and `p` correspond to the `size` and `prob` arguments of `stats::dnbinom` in R.
 
@@
 The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model.
-Here I apply the piecewise constant approximation, defininthe observations $y_i$ to be counts of occurrence records in each cell $i$, having area $a_i$, where the cells represent a partition of the study area [@diggleSpatialSpatioTemporalLogGaussian2013;@simpsonGoingGridComputationally2011].
-That is they tesselate (but do not overlap) and together cover the entire study area.
-Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogenous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting.
-The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\tau_i$ (also assumed homogenous across each cell).
-Values of $\tau_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression submodel against detection covariates.
+Here I apply the piece-wise constant approximation, defining the observations $y_i$ to be counts of occurrence records in each cell $i$, having area $a_i$, where the cells represent a partition of the study area [@diggleSpatialSpatioTemporalLogGaussian2013;@simpsonGoingGridComputationally2011].
+That is they tessellate (but do not overlap) and together cover the entire study area.
+Under this definition, the DSDM relative abundance parameter $\Lambda_i$ represents a density of point observations per unit area (assumed to be homogeneous across cell $i$) that *would* be seen if there were no spatial bias in rates of reporting.
+The expected number of presence-only observations in each cell is therefore the product of the relative abundance $\Lambda_i$, the area $a_i$, and a relative measure of the bias in reporting, $\tau_i$ (also assumed homogeneous across each cell).
+Values of $\tau_i$ for each location can be provided as data, or estimated during model-fitting via an additional regression sub-model against detection covariates.
 Since $\Lambda_i$ is only identified to within a multiplicative constant ($\alpha$ in Equation 2), the absolute value of $\tau_i$ is not identified in this model, so any model for $\tau_i$ should probably not have an intercept term.
@@
 ```
-Where `pa_locations` and `count_locations` are numeric indices to the sites with to presence-absence and count data respectively, and `y_pa` and `y_count` are the corresponding vectors of observed data.
 
@@
 Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $M$ life stages (indexed with $j$) at site $i$, at time $t$.
-$\mathbf{n}_{i,t}$ is evolved through time by multiplication against a timepoint-specific transition matrix $\mathbf{A}_{i,t}$, contructed of timepoint-specific vital rates that depend on the total population density (i.e.
+$\mathbf{n}_{i,t}$ is evolved through time by multiplication against a time-point-specific transition matrix $\mathbf{A}_{i,t}$, constructed of time-point-specific vital rates that depend on the total population density (i.e.
 across all life stages $M$) in the previous time point $\mathbf{N}_{i,t}$.
@@
 - `alpha_S`: a scalar for the effect of population density on logit-survival
-- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
+- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
 
@@
 ```
-Given additional scalar parameters `alpha_F` and `alpha_S`.
 
@@
 Given temporally-indexed response data $y_{i,t}$, temporal variation in environmental conditions and therefore vital rates can be included in a DSDM in one of two ways.
-The applicability of each of these depends on both the temporal scale of the data (the period of time betwen the observations) and the demographics of the species.
+The applicability of each of these depends on both the temporal scale of the data (the period of time between the observations) and the demographics of the species.
 
-If the length of time between observations is on the order of multiple generations opf the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environental conditions.
+If the length of time between observations is on the order of multiple generations of the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environmental conditions.
 In other words, that the population is at, or close to, the long-term carrying capacity of the species if the current environmental conditions were to be held static.
-In this situation, considering temporal change in environmental conditions is simply a matter of redifining the index $i$ in the basic DSDM to indeex unique combinations of time and site, instead of just site.
+In this situation, considering temporal change in environmental conditions is simply a matter of redefining the index $i$ in the basic DSDM to index unique combinations of time and site, instead of just site.
 The relative abundance parameter for each of these observations is therefore computed from the intrinsic growth rate of the transition matrix at the corresponding environmental conditions.
@@
 
-Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive timesteps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above.
-Instead of iterating for some arbitrrary number of iterations $T$ and taking the final state, instead the population is iterated over timesteps $t$ corresponding to time points in the dataset (and all intermediate timesteps).
-Thus the population at each timestep will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.
+Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive time-steps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above.
+Instead of iterating for some arbitrary number of iterations $T$ and taking the final state, instead the population is iterated over time-steps $t$ corresponding to time points in the dataset (and all intermediate time-steps).
+Thus the population at each time-step will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.
 
@@
 - `alpha_S`: a scalar for the effect of population density on logit-survival
-- `gamma`: a (positive) scalar for the difference in logit-survival betwen adults and juveniles
-- `y`: an $N$x$T$ matrix of presence-absence data at each timestep and site
+- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
+- `y`: an $N$x$T$ matrix of presence-absence data at each time-step and site
 
@@
 
+## Dispersal with a static transition matrix
+
+Dispersal of individuals between sites can be incorporated by expanding the matrix population model to simultaneously consider populations in all patches, with dispersal between patches quantified by a dispersal kernel.
+
+In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$.
+To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
+The the $i^{th}$ block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this time-step.
+$d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.
+
+By connecting the populations at all sites with immigration and emigration potential, we now have a single intrinsic growth rate for the entire population rather than for the population at each site.
+Rather than computing the relative abundance at each site $\Lambda_i$ as a function of an intrinsic growth rate, we can instead sum stable stage distributions over the stages in each site to yield a stable population distribution over sites.
+$\Lambda_i$ can therefore be computed as proportional (with constant $\alpha$) to this relative population size.
+
+In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$.
+This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse.
+It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g.
+by iteration using sparse matrix multiplication, as in the biotic interaction model.
+
+#### Notation
+
+\begin{align}
+\Lambda_i &= \alpha v_i / \textstyle\sum_{i=1}^{N}{v_{i}} \\ \nonumber
+v_i &= \textstyle\sum_{j=1}^{M}{v^*_{j+(i-1)}} \\ \nonumber
+v^* &= eigvec(\mathbf{A}^*) \\ \nonumber
+\mathbf{A}^* &= \begin{bmatrix} \mathbf{A}_1 & \mathbf{M}_{1,2} & \dots & \mathbf{M}_{1,i'} & \dots & \mathbf{M}_{1,N}\\
+\mathbf{M}_{2,1} & \mathbf{A}_2 & \dots & \mathbf{M}_{2,i'} & \dots & \mathbf{M}_{2,N} \\
+\vdots & \vdots & \ddots & \vdots & &  \vdots \\
+\mathbf{M}_{i,1} & \mathbf{M}_{i,2} & \dots & \mathbf{A}_i & \dots & \mathbf{M}_{i,N} \\
+\vdots & \vdots &  & \vdots & \ddots & \vdots \\
+\mathbf{M}_{N,1} & \mathbf{M}_{N,2} & \dots & \mathbf{M}_{N,i'} & \dots & \mathbf{A}_N \end{bmatrix} \\ \nonumber
+\mathbf{M}_{i,i'} &= diag(d_{1,i,i'}, \dots, d_{j,i,i'}, \dots, d_{M,i,i'}) \\ \nonumber
+d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
+d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
+\end{align}
+
+##### Code
+
+Input:
+
+- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
+- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
+- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
+- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
+
+```{r static_dispersal}
+dispersal_kernel <- function (distance) {
+  exp(-tau * distance)
+}
+
+analysis <- iterate_matrix(
+  matrix = A,
+  distances = distance_matrix,
+  dispersal_kernel = dispersal_kernel
+)
+
+Lambda <- alpha * analysis$stable_abundance_distribution
+```
+
+## Dispersal with dynamic transition matrix
+
+Where the transition matrix is assumed to be dynamic - i.e..
+to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
+Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
+As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.
+
+I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each life-stage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
+For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
+
+The intermediate population states are computed as in earlier models by matrix multiplication of the state vectors $\mathbf{n}_{i,t}$ with the corresponding transition matrix $\mathbf{A}_i$.
+The ultimate population state for this time point is then given by matrix multiplications - separately for each of the *life stages* - of the intermediate population states against the dispersal matrix, formed from the distance matrix and dispersal function as in the dispersal model for static transition matrices.
+That is, the matrix multiplication is applied to the other dimension of the population state matrix.
+
+##### Notation
+
+\begin{align}
+\mathbf{n}_{j,t+1} &= \mathbf{d}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
+\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
+d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
+d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
+\end{align}
+
+where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilities for life stage $j$, which sums to 1 across each column.
+
+##### Code
+
+Input:
+
+- `A_t_fun`: a function yielding for each time point a $N$x$M$x$M$ array of transition matrices for each site
+- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
+- `timesteps`: the number of time-steps to iterate for
+- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
+- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
+- `...`: additional named arguments to `A_t_fun`, which may be lists with different elements to be used at each time point
+
+```{r dynamic_dispersal}
+dispersal_kernel <- function (distance) {
+  exp(-tau * distance)
+}
+
+iterations <- iterate_dynamic_matrix(
+  matrix_function = A_t_function, 
+  initial_state = n_init,
+  niter = timesteps,
+  distances = distance_matrix,
+  dispersal_kernel = dispersal_kernel,
+  ...
+)
+
+Lambda <- iterations$all_states
+```
+
+Note that the final line could be replaced with: `Lambda <- iterations$stable_population` to yield the long-term stable population in the case that dispersal and density-dependent vital rates are required, but neither the effects of environmental stochasticity nor a time series of population sizes is required.
+
 ## Species with complex demography
@@
 Simultaneous demographic species distribution modelling of multiple species can be achieved by replacing the population transition matrices at each site $\mathbf{A}_i$ with a block-diagonal matrix that has the population transition matrices for each species on the diagonal.
-For model of $L$ species each with $M$ lifestages $\mathbf{A}_i$ therefore has dimension $LM$x$LM$.
+For model of $L$ species each with $M$ life-stages $\mathbf{A}_i$ therefore has dimension $LM$x$LM$.
 The off-diagonal blocks are all zero-matrices, since an individual can neither change species nor produce offspring of another species.
@@
 
-As in the single-species density-dependent vital rate model above the relative population density for each species is computed by iteration; at each timestep recalculating the transition matrix from population densities of all species and matrix-multiplying it by the current population density vector to obtain the subsequent population density vector.
+As in the single-species density-dependent vital rate model above the relative population density for each species is computed by iteration; at each time-step recalculating the transition matrix from population densities of all species and matrix-multiplying it by the current population density vector to obtain the subsequent population density vector.
 The transition matrix $\mathbf{A}_i$ is sparse; with only 1 in $L$ elements being non-zero.
@@
 
-where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are 3D arrays with dimension $L$x$N$x$T$ giving the fecundity, adult and juvenile survival and relative population density of each species $l$ at site $i$ at time $t$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ or regression coefficients for the submodels for each of species, and $L$x$L$ matrices $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$ of density-dependence effects between specie.
+where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are 3D arrays with dimension $L$x$N$x$T$ giving the fecundity, adult and juvenile survival and relative population density of each species $l$ at site $i$ at time $t$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ or regression coefficients for the sub-models for each of species, and $L$x$L$ matrices $\mathbf{\alpha}_F$ and $\mathbf{\alpha}_S$ of density-dependence effects between specie.
 I.e.
@@
 - `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each species
-- `alpha_F`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on log-fecundity
-- `alpha_S`: a $K$x$K$ matrix of the intra- and inter-specific effects of population density on logit-survival
-- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival betwen adults and juveniles for each species
+- `alpha_F`: a $L$x$L$ matrix of the intra- and inter-specific effects of population density on log-fecundity
+- `alpha_S`: a $L$x$L$ matrix of the intra- and inter-specific effects of population density on logit-survival
+- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival between adults and juveniles for each species
 
@@
 In this model, the genotypes are linked via reproduction rather than density-dependent vital rates.
-tThe $L$x$L$ matrix $\mathbf{m}$ quantifies mutation rates between different genotypes.
+The $L$x$L$ matrix $\mathbf{m}$ quantifies mutation rates between different genotypes.
 For an individual of genotype $l$, each of its offspring has probability $\mathbf{m}_{l,l'}$ of being of genotype $l'$.
-Each genotype has its own vital rate submodels, representing a linked phenotype driving population dynamics for that genotype.
+Each genotype has its own vital rate sub-models, representing a linked phenotype driving population dynamics for that genotype.
 By analysing the combined transition matrix $\mathbf{A}^*_i$ at each site $i$ with different environmental covariates, the model can consider how selection for each genotype varies across the environment, and predict the genetic makeup of populations.
@@
 
-where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are $L$x$N$ matrices giving the fecundity, adult and juvenile survival and relative population density of each genotype $l$ at site $i$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ of regression coefficients for the submodels for each of species.
+where $\mathbf{F}$, $\mathbf{S}_A$, $\mathbf{S}_J$, and $\mathbf{N}$ are $L$x$N$ matrices giving the fecundity, adult and juvenile survival and relative population density of each genotype $l$ at site $i$, computed from $K$x$L$ matrices $\beta_F$ and $\beta_S$ of regression coefficients for the sub-models for each of species.
 $\mathbf{m}$ is an $L$x$L$ matrix of positive values with columns summing to one, giving the probabilities of a parent of one genotype producing offspring of each genotype.
@@
 
-where $\mathbf{B}$ is an $M$x$M$ matrix with ones on the first row and zeros elsewhere (ie.
-a mask indicating cells representing fecundity rates), $\mathbf{C}$ is its reverse (zeros on the first row, ones elsewhere) $\mathbf{1}$ is length $L$ vector of ones, $\mathbf{I}$ is an identity matrix of dimension $L$, and $\odot$ and $\otimes$ represent the Hadamard (elementwise) and Kronecker products respectively.
+where $\mathbf{B}$ is an $M$x$M$ matrix with ones on the first row and zeros elsewhere (i.e..
+a mask indicating cells representing fecundity rates), $\mathbf{C}$ is its reverse (zeros on the first row, ones elsewhere) $\mathbf{1}$ is length $L$ vector of ones, $\mathbf{I}$ is an identity matrix of dimension $L$, and $\odot$ and $\otimes$ represent the Hadamard (element-wise) and Kronecker products respectively.
 
@@
 
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a $K$x$L$ matrix of regression coefficients for log-fecundity for each genotype
+- `beta_S`: a $K$x$L$ matrix of regression coefficients for logit-survival for each genotype
+- `m`: a $L$x$L$ matrix of the transition (mutation) rates between genotypes, with columns summing to one
+- `gamma`: a (positive) vector of length $K$ for the differences in logit-survival between adults and juveniles for each species
+
 ```{r evolutionary_dynamics}
+# vital rate models
+log_F<- X %*% beta_F
+logit_S<- X %*% beta_S
+F <- exp(log_F)
+S_A <- plogis(logit_S)
+S_J <- plogis(logit_S - gamma)
 
+L <- length(gamma)
+N <- nrow(X)
+M <- 2
+padding <- zeros(N, L * M)
+B <- zeros(M, M)
+B[1, ] <- 1
+C <- 1 - B
+dd_weights <- diag(L)[rep(1:L, each = M), ]
+
+# array A_star of transition matrices
+mutation <- kronecker(m, B) + kronecker(diag(L), C)
+
+F_rep <- do.call(rbind, replicate(M, F, simplify = FALSE))
+dim(F_rep) <- c(N, L * M)
+F_kron_tB <- abind(F_rep, padding, along = 3)
+
+SJ_rep <- do.call(cbind, replicate(M, S_J, simplify = FALSE))
+SJ_kron_B <- rbind(SJ_rep, padding)
+dim(SJ_kron_B) <- c(N, L * 2,  M)
+
+SA_rep <- do.call(cbind, replicate(M, S_A, simplify = FALSE))
+SA_kron_C <- rbind(padding, SA_rep)
+dim(SA_kron_C) <- c(N, L * 2,  M)
+
+R <- F_kron_tB + SJ_kron_B + SA_kron_C
+one_kron_R <- do.call(abind, c(replicate(L, R, simplify = FALSE), list(along = 3)))
+dim(one_kron_R) <- c(N, L * M, L * M)
+one_kron_R <- aperm(one_kron_R, c(1, 3, 2))
+
+# do multiplication across N as a matrix
+dim(one_kron_R) <- c(N, L ^ 2 * M ^ 2)
+dim(mutation) <- c(L ^ 2 * M ^ 2)
+A_star <- sweep(one_kron_R, 2, mutation, "*")
+dim(A_star) <- c(N, L * M, L * M)
+
+analysis <- iterate_matrix(A_star)
+v_star <- analysis$stable_abundance_distribution
+v_standardised <- analysis$stable_abundance_distribution %*% dd_weights
+lambda <- analysis$lambda
+K <- (lambda - 1) * alpha
+Lambda_star <- log1pe(K)
+Lambda <- Lambda_star * v_standardised
 ```
 
-## Dispersal with a static transition matrix
-
-Dispersal of individuals between sites can be incorporated by expanding the matrix population model to simultaneously consider populations in all patches, with dispersal between patches quantified by a dispersal kernel.
-
-In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$.
-To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
-The the $i$^{th} block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this timestep.
-$d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.
-
-By connecting the populations at all sites with immigration and emigration potential, we now have a single intrinsic growth rate for the entire population rather than for th eopulation at each site.
-Rather than computing the relative abundance at each site $\Lambda_i$ as a function of an intrinsic growth rate, we can instead sum stable stage distributions over the stages in each site to yeild a stable population distribution over sites.
-$\Lambda_i$ can therefore be computed as proportional (with constant $\alpha$) to this relative population size.
-
-In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$.
-This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse.
-It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g.
-by iteration using sparse matrix multiplication, as in the biotic interaction model.
-
-#### Notation
-
-\begin{align}
-\Lambda_i &= \alpha v_i / \textstyle\sum_{i=1}^{N}{v_{i}} \\ \nonumber
-v_i &= \textstyle\sum_{j=1}^{M}{v^*_{j+(i-1)}} \\ \nonumber
-v^* &= eigvec(\mathbf{A}^*) \\ \nonumber
-\mathbf{A}^* &= \begin{bmatrix} \mathbf{A}_1 & \mathbf{M}_{1,2} & \dots & \mathbf{M}_{1,i'} & \dots & \mathbf{M}_{1,N}\\
-\mathbf{M}_{2,1} & \mathbf{A}_2 & \dots & \mathbf{M}_{2,i'} & \dots & \mathbf{M}_{2,N} \\
-\vdots & \vdots & \ddots & \vdots & &  \vdots \\
-\mathbf{M}_{i,1} & \mathbf{M}_{i,2} & \dots & \mathbf{A}_i & \dots & \mathbf{M}_{i,N} \\
-\vdots & \vdots &  & \vdots & \ddots & \vdots \\
-\mathbf{M}_{N,1} & \mathbf{M}_{N,2} & \dots & \mathbf{M}_{N,i'} & \dots & \mathbf{A}_N \end{bmatrix} \\ \nonumber
-\mathbf{M}_{i,i'} &= diag(d_{1,i,i'}, \dots, d_{j,i,i'}, \dots, d_{M,i,i'}) \\ \nonumber
-d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
-d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
-\end{align}
-
-##### Code
-
-Input:
-
-- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
-- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
-- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
-- `alpha`: a (positive) scalar for the combind effect of imperfect detection and density dependence
-
-```{r static_dispersal}
-dispersal_kernel <- function (distance) {
-  exp(-tau * distance)
-}
-
-analysis <- iterate_matrix(
-  matrix = A,
-  distances = distance_matrix,
-  dispersal_kernel = dispersal_kernel
-)
-
-Lambda <- alpha * analysis$stable_abundance_distribution
-```
-
-## Dispersal with dynamic transition matrix
-
-Where the transition matrix is assumed to be dynamic - ie.
-to model the effects of environmental stochasticity, depsnity-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
-Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
-As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.
-
-I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each lifestage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
-For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
-
-The intermediate population states are computed as in earlier models by matrix multiplication of the state vectors $\mathbf{n}_{i,t}$ with the corresponding transition matrix $\mathbf{A}_i$.
-The ultimate population state for this time point is then given by matrix multiplications - separately for each of the *life stages* - of the intermediate population states against the dispersal matrix, formed from the distance matrix and dispersal function as in the dispersal model for static transition matrices.
-That is, the matrix multiplication is applied to the other dimension of the population state matrix.
-
-##### Notation
-
-\begin{align}
-\mathbf{n}_{j,t+1} &= \mathbf{d}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
-\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
-d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
-\end{align}
-
-where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilitiesfor life stage $j$, which sums to 1 across each column.
-
-##### Code
-
-Input:
-
-- `A_t_fun`: a function yielding for each time point a $N$x$M$x$M$ array of transition matrices for each site
-- `n_init`: a (positive) vector of length $M$ or matrix of dimension $N$x$M$ giving the initial population state at each location
-- `timesteps`: the number of timesteps to iterate for
-- `distance_matrix`: a (possibly sparse) $N$x$N$ matrix of distances between sites
-- `tau`: a vector of length $M$ of (positive) dispersal ability parameters for each stage
-- `...`: additional named arguments to `A_t_fun`, which may be lists with different elements to be used at each time point
-
-```{r dynamic_dispersal}
-dispersal_kernel <- function (distance) {
-  exp(-tau * distance)
-}
-
-iterations <- iterate_dynamic_matrix(
-  matrix_function = A_t_function, 
-  initial_state = n_init,
-  niter = timesteps,
-  distances = distance_matrix,
-  dispersal_kernel = dispersal_kernel,
-  ...
-)
-
-Lambda <- iterations$all_states
-```
-
-Note that the final line could be replaced with: `Lambda <- iterations$stable_population` to yield the long-term stable population in the case that dispersal and density-dependent vital rates are required, but neither the effects of environmental stochasticity nor a time series of population sizes is required.
-
 # References

### 2019-10-22 manuscript/appendix_a.Rmd
add MPM with complex demography

@@
 
-Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogenous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015].
+Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogeneous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015].
 The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model.
@@
 In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$.
-To consider interacting population dynamics across sites, we instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
+To consider interacting population dynamics across sites, we can instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
 The the $i^{th}$ block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this time-step.
@@
 Where the transition matrix is assumed to be dynamic - i.e..
-to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
-Because we explicitly iterate the matrix, we can apply dispersal with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
+to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we can use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
+Because the matrix is explicitly iterated dispersal can be incorporated with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
 As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.
@@
 I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each life-stage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
-For simplicity of notation, we assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
+For simplicity of notation, I assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
 
@@
 
-<illustrate that any species for which a transition matrix can be formed can be modelled with a DSDM- at least to illustrate that multiple life stages can be considered.>
+The DSDMs examples covered so far have all used the same two-stage population transition matrix, with adult and juvenile stages.
+This is by no means a limitation of the DSDM approach, the transition matrix can encode as many life-stages and as much complexity as is required to answer the question at hand and is supported by the available data and prior information.
+
+As an illustration of this, the following example implements a matrix model for a plant species with more complex life history; three adult stages $A_{1-3}$ and two seed bank stages $S_{2-3}$, as described in [@griffithPopulationMatrixModels2005].
+As in the original paper, I  simplify the 9 vital rates from the original paper to 4, which I denote, for site $i$: $B_{AA_i}$ (direct recruitment), $B_{AS_i}$ (contribution to the seed bank), $P_{SS_i}$ (persistence in the seed bank), and $P_{SA_i}$ (recruitment from the seed bank).
+I.e.: $B_{AA_i} = B_{A_1-A_1} = B_{A_2-A_1} = B_{A_3-A_1}$, $B_{AS_i} = B_{A_1-S_2} = B_{A_2-S_2} = B_{A_3-S_2}$, $P_{SS_i} = P_{S_2-S_3}$, and $P_{SA_i} = P_{S_2-A_2} = P_{S_3-A_3}$.
+In turn, these rates are formed from one rate: the mean number of seeds per plant $S_i$; and 6 probabilities that I concatenate into a single vector $\mathbf{P}_i$ for each site: the fraction of seeds not dispersing ($\mathbf{P}_{i,1} = \text{No }D$), the probability of staying in the soil without germinating ($\mathbf{P}_{i,2} = P_{bank}$, the probability of a seed germinating once it has survived winter $\mathbf{P}_{i,3} = P_{germ}$, the probability of seed germinating and surviving to the first true leaf stage $\mathbf{P}_{i,4} = P_{estab}$, the probability of a seedling surviving from the first true leaf stage to the time of reproduction $\mathbf{P}_{i,5} = P_{surv}$, and the probability of a seed on the ground surviving overwinter to germination time $\mathbf{P}_{i,6} = P_{winter}$.
+I assume that all seven of these life history parameters vary with environmental covariates, and again model them with linear regression sub-models.
+
+##### Notation
+
+\begin{align}
+\mathbf{A}_i &= \begin{bmatrix} 0 & 0 & B_{AS_i} & B_{AS_i} & B_{AS_i} \\ \nonumber
+P_{SS_i} & 0 & 0 & 0 & 0 \\ \nonumber
+0 & 0 & B_{AA_i} & B_{AA_i} & B_{AA_i} \\ \nonumber
+P_{SA_i} & 0 & 0 & 0 & 0 \\ \nonumber
+0 & P_{SA_i} & 0 & 0 & 0 \end{bmatrix} \\ \nonumber
+B_{AS_i} &= S_i \mathbf{P}_{i,1} \mathbf{P}_{i,2} \\ \nonumber
+B_{AA_i} &= S_i \mathbf{P}_{i,1} P_{SA_i} \\ \nonumber
+P_{SS_i} &= \mathbf{P}_{i,6} \mathbf{P}_{i,2} \\ \nonumber
+P_{SA_i} &= \mathbf{P}_{i,3} \mathbf{P}_{i,4} \mathbf{P}_{i,5} \\ \nonumber
+logit(\mathbf{P}) &= \mathbf{X}\beta_P \\ \nonumber
+ln(S_i) &= \mathbf{X}_i\beta_S
+\end{align}
+
+##### Code
+
+Input:
+
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_P`: a $K$x$6$ matrix of regression coefficients for the 6 logit-probabilities in the vector $\mathbf{P}_i$
+- `beta_S`: a length $K$ vector of regression coefficients for log-seed-production
+
+```{r complex_demography}
+P <- ilogit(X %*% beta_P)
+S <- ilogit(X %*% beta_S)
+
+P_SA <- apply(P[, 3:5], 1, "prod")
+P_SS <- P[, 6] * P[, 2]
+B_AA <- S * P[, 1] * P_SA
+B_AS <- S * apply(P[, 1:2], 1, "prod")
+
+N <- nrow(X)
+zero <- zeros(N)
+col1 <- cbind(zero, P_SS, zero, P_SA, zero)
+col2 <- cbind(zeros(N, 4), P_SA)
+col3 <- cbind(B_AS, zero, B_AA, zeros(N, 2))
+A <- abind(col1, col2, col3, col3, col3, along = 3)
+```
 
@@
 
-<to do, need to package up integration code into greta.dynamics?>
+<to do, don't package up integration code into greta.dynamics yet, show it warts-and-all for now>
 
@@
 
-This is the form that we use in the code below, vectorised across all $N$ sites, since it is more computationally efficient to carry out the calculations in this way in greta.
+This is the form that I use in the code below, vectorised across all $N$ sites, since it is more computationally efficient to carry out the calculations in this way in greta.
 

### 2019-10-22 manuscript/appendix_a.Rmd
tidy up and add reference

@@
 
-- `size`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution
+- `r`: a (positive) scalar giving the dispersion parameter of the negative binomial distribution (equivalent to the `size` argument of `stats::dnbinom()`)
 - `Lambda`: a (positive) vector of the relative abundances for each observation
@@
 ```{r negbin_obs}
-prob <- size / (size + Lambda)
-distribution(y) <- negative_binomial(size, prob)
+p <- r / (r + Lambda)
+distribution(y) <- negative_binomial(r, p)
 ```
@@
 By analysing the combined transition matrix $\mathbf{A}^*_i$ at each site $i$ with different environmental covariates, the model can consider how selection for each genotype varies across the environment, and predict the genetic makeup of populations.
+Note that this is just one approach to considering multiple genotypes and natural selection in a matrix population model, see e.g.
+[@devriesSelectionTwosexStagestructured2019].
 

### 2019-10-25 manuscript/appendix_a.Rmd
tweak text
fix formatting issue
restructure; corresct static dispersal with vec-perumtation matrix

@@
     toc: true
-    toc_depth: 2
+    toc_depth: 3
   word_document:
@@
 
-For ease of reference, I repeat here the equations and interpretation for the basic presence-absence DSDM presented in the main manuscript.
-It presents the model in its three parts, each of which can be modified independently.
-## Observation model
+For ease of reference I repeat here the notation, and add the code, for the basic presence-absence DSDM presented in the main manuscript.
+The model is composed of three of the components detailed below: a static two-stage transition matrix model of population dynamics, a Beverton-Holt model of abundance, and a presence-absence observation model.
 
-The observation model for the basic DSDM relates the species' relative abundance (I.e.
-assumed to be within a multiplicative constant of the true abundance) $\Lambda_i$ at site $i$ to the probability of (observed) presence $p_i$ during a single presence-absence survey, and the presence-absence data are assumed to be independently Bernoulli-distributed following these probabilities as in a presence-absence species distribution model.
+##### Notation
+
+\begin{align}
+y_i &\sim Bernoulli(p_i) \\ \nonumber
+p_i &= 1 - e^{-\Lambda_i} \\ \nonumber
+\Lambda_i &= ln(1 + e^{K_i})  \\ \nonumber
+K_i &= (\lambda_i - 1) \alpha \\ \nonumber
+\lambda_i &= eigval(\mathbf{A}_i) \\ \nonumber
+\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{J_i} & F_i \cdot S_{A_i} \\ S_{J_i} & S_{A_i}  \end{bmatrix} \\ \nonumber
+ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
+logit(S_{A_i}) &= \mathbf{X}_i\beta_S \\ \nonumber
+logit(S_{J_i}) &= logit(S_{A_i}) - \gamma
+\end{align}
+
+where $y$ is a vector of length $N$ giving the presence or absence of the species at site $i$, $eigval(\mathbf{A}_i)$ denoted the dominant eigenvalue of the population transition matrix at site $i$; the intrinsic growth rate of that population, $\mathbf{X}$ a $N$x$K$ design matrix of the values of environmental covariates at each site, an intercept (coded as a column of all ones), and any transformations of those covariates, $\beta_F$ and $\beta_S$ are parameter vectors of length $K$ giving regression coefficients for the log-fecundity and logit-survival submodels respectively, $\gamma$ is a positive scalar parameter for the logit-difference between adult and juvenile survival, and $\alpha$ is a positive scalar parameter for the combined effects of imperfect detection and the relationship between the intrisic growth rate and abundance in the Beverton-Holt model.
+
+##### Code
+
+Input:
+
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
+- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
+- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
+- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
+- `y`: a vector of 1s and 0s encoding presence-absence observations
+
+```{r basic_dsdm}
+log_F <- X %*% beta_F
+logit_S_A <- X %*% beta_S
+F <- exp(log_F)
+S_A <- ilogit(logit_S_A)
+S_J <- ilogit(logit_S_A - gamma)
+top <- cbind(F * S_J, F * S_A)
+bottom <- cbind(S_J, S_A)
+A <- abind(top, bottom, along = 3)
+
+analysis <- iterate_matrix(A)
+lambda <- analysis$lambda
+K <- (lambda - 1) * alpha
+Lambda <- log1pe(K)
+
+p <- 1 - exp(-Lambda)
+distribution(y) <- bernoulli(p)
+```
+
+# Observation models
+
+Below I provide alternative observation models for presence-absence, count, and presence-only species distribution data, linking their parameters to the variable $\Lambda_i$ representing relative abundance at site $i$.
+These observation models can also be extended to distribution data that are disaggregated by time, stage, species, or genotype by iterating over the appropriate index.
+A wider range of observation models for distribution data - such as that for disentangling occupancy and detection - can be implemented using other greta extension packages, but are beyond the scope of this manuscript.
+Similarly, observation models for demographic data (e.g.
+for an integrated population modelling approach) could be included, but are not considered here, since the focus of this paper is the types of data typically used for statistical modelling of species distributions.
+
+## Presence-absence data
+
+Presence-absence data $y_i$ are assumed to be independently Bernoulli-distributed following a probability of presence $p_i$, as in a standard presence-absence species distribution model.
+This probability of presence can be calculated from the relative abundance by considering the presence-absencve data as arising from a truncation of count data: the species recorded as present if one or more individuals were detected, or otherwise a absent.
+By assuming that count follow a Poisson distribution with expected value $\Lambda_i$, the probability of species presence (a count greater than 0) is given by $1-e^{-\Lambda_i}$.
+This is equivalent to the likelihood of a GLM for binary data with complementary log-log link function and linear predictor $ln(\Lambda_i)$.
 
@@
 ```
-
-## Abundance model
-
-The abundance model relates the transition matrix $\mathbf{A}_i$ of the population at site $i$ to the relative abundance $\Lambda_i$.
-The intrinsic growth rate $\lambda_i$ of the population is used to compute the theoretical carrying capacity $K_i$ according to the Beverton-Holt model, with competition effect parameter $\alpha$.
-This theoretical carrying capacity is negative when $\lambda_i < 1$, so it is rectified by the softplus function to always yield a positive abundance $\Lambda_i$, which is close to zero when $\lambda_i < 1$, and approximately equal to $K_i$ when $\lambda_i > 1$.
-
-##### Notation
-
-\begin{align}
-\Lambda_i &= ln(1 + e^{K_i})  \\ \nonumber
-K_i &= (\lambda_i - 1) \alpha \\ \nonumber
-\lambda_i &= eigval(\mathbf{A}_i)
-\end{align}
-
-##### Code
-
-Input:
-
-- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
-- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
-
-```{r basic_abundance}
-analysis <- iterate_matrix(A)
-lambda <- analysis$lambda
-K <- (lambda - 1) * alpha
-Lambda <- log1pe(K)
-```
-
-## Population dynamic model
-
-The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous time step (columns) to each stage in the present time step (rows).
-Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult.
-For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001].
-
-Each of these vital rates at each site $i$ is given by a linear model against a design matrix $\mathbf{X}_i$ representing environmental covariates, an intercept (coded as a column of all ones), and any transformations of those covariates
-
-##### Notation
-
-\begin{align}
-\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{J_i} & F_i \cdot S_{A_i} \\ S_{J_i} & S_{A_i}  \end{bmatrix} \\ \nonumber
-ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
-logit(S_{A_i}) &= \mathbf{X}_i\beta_S \\ \nonumber
-logit(S_{J_i}) &= logit(S_{A_i}) - \gamma
-\end{align}
-
-##### Code
-
-Input:
-
-- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
-- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
-- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
-- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
-
-```{r basic_transition}
-log_F <- X %*% beta_F
-logit_S_A <- X %*% beta_S
-F <- exp(log_F)
-S_A <- ilogit(logit_S_A)
-S_J <- ilogit(logit_S_A - gamma)
-
-top <- cbind(F * S_J, F * S_A)
-bottom <- cbind(S_J, S_A)
-A <- abind(top, bottom, along = 3)
-```
-
-# Observation models
-
-Below I provide alternative observation models for count, and presence-only data.
-occupancy detection observation models can be implemented using other greta extension packages, but are beyond the scope of this manuscript.
 
@@
 
-Specification of integrated models (those with multiple likelihoods for different datatypes) is simple, simply by using multiple `distribution()` statements.
-For example, a a combined presence-absence and abundance model might be defined as follows.
+Specification of integrated models (those with multiple likelihoods for different datatypes) is straightforward.
+For example, a combined observation  presence-absence and abundance model might be defined as:
+
+\begin{align}
+y_i &\sim \begin{cases}
+Bernoulli(p_i), & \text{if } z_i = 1 \\ \nonumber
+Poisson(\alpha^* \Lambda_i) , & \text{otherwise}
+\end{cases} \\ \nonumber
+p_i &= 1 - e^{-\Lambda_i}
+\end{align}
+
+where $z_i$ is an indicator variable for whether the observation $y_i$ represents presence-absence or count data, and $\alpha^*$ is a detection rate parameter - in addition to the $\alpha$ used to define $\Lambda_i$- giving the detectability of individuals in the surveys yielding the count data, relative to that in the surveys yielding the presence-absence data.
 
@@
 - `count_locations`: a vector of (positive) integers with the same length as `y_count` giving the site numbers corresponding to count data
-- `alpha_count`: a (positive) scalar giving the detection rate for the count data, relative to the detection rat of the presence-absence data.
+- `alpha_star`: a (positive) scalar giving the detection rate for the count data, relative to the detection rate of the presence-absence data.
 
@@
 distribution(y_pa) <- bernoulli(p)
-distribution(y_count) <- poisson(alpha_count * Lambda[count_locations])
+distribution(y_count) <- poisson(alpha_star * Lambda[count_locations])
 ```
@@
 
-## Ricker model
+This section introduces a number of options for deriving a site-specific relative abundance from the transition matrix representing population dynamics.
+These can be broadly split into two groups - those computed from a temporally-static population transition matrix and those from a dynamic transition matrix.
 
-The abundance relationship in the DSDM presented in the manuscript derives from the Beverton-Holt model, with the carrying capacity $K_i$ proportional to the intrinsic growth rate $\lambda_i$ minus one.
-This ensures that $K_i$ is zero if $\lambda_i = 1$ (no growth) and increases as lambda increases, though it has the unfortunate consequence that $K_i$ has negative values for $\lambda_i < 1$, which must be rectified in that model.
-An alternative abundance relationship approach, given in [@pagelForecastingSpeciesRanges2012], derives from the Ricker model and has a linear relationship between $K_i$ and $\lambda_i$, yielding a relative abundance expression that does not need rectifying.
-Though with large enough $\alpha$, this approach can yield large values of $\Lambda_i$ even if $\lambda_i$ is well below 1, in which  case the species should not be able to persist.
+## Static transition matrices
+
+Static population transition matrices are the typical form of matrix population models, following the definition of population dynamics over a stage-structured population size vector $\mathbf{n}_{i,t}$ at site $i$:
+
+\begin{align}
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i} \mathbf{n}_{i,t}
+\end{align}
+
+The population transition matrix $\mathbf{A}_i$ at each site is constant between iteration times $t$.
+Because $\mathbf{A}_i$ is constant, it can be analysed to compute an intrinsic growth rate and stable stage distribution, without having to explicitly consider population sizes $\mathbf{n}_{i,t}$.
+This growth rate and/or stable stage distribution can then be related to a relative abundance $\Lambda_i$ under some assumptions.
+This is the approach used in the DSDM examples in the manuscript.
+
+### Beverton-Holt model
+
+The Beverton-Holt model of population dynamics for single (not stage-structured) population $n_{i}$ is:
+
+\begin{align}
+n_{i,t+1} &= \frac{\lambda_i n_{i,t}}{1 + n_{i,t} / M}
+\end{align}
+
+where $\lambda_i$ is the intrinsic growth rate of the population (which can be calculated as the dominant eigenvalue of $\mathbf{A}_i$), and $M$ is a scaling factor implicitly representing density-dependent effects.
+The carrying capacity of the Beverton-Holt model is given by $(\lambda_i - 1) M$, which we assume to be be proportional to the relative abundance that we can observee via species distribution data (assuming the population is at equilibrium), with the constant of proportionality being a per-individual detection rate - the fraction of individuals in the population detected in the survey from which we obtained our distribution data.
+This detection rate is not identified from the density-dependence parameter $M$, so we replace their product with the parameter $\alpha$ to yield the theoretical relative carrying capacty parameter $K_i = (\lambda_i - 1) \alpha$.
+An unfortunate consequence of this simple relationship is that $K_i$ is negative when $\lambda_i < 1$, and thus not a valid population size.
+It is therfore rectified (forced to be positive) by the softplus function to give the relative abundance $\Lambda_i = ln(1 + e^{K_i})$, so that $\Lambda_i$ is close to zero when $\lambda_i < 1$, and approximately equal to $K_i$ when $\lambda_i > 1$.
+
+##### Notation
+
+\begin{align}
+\Lambda_i &= ln(1 + e^{K_i})  \\ \nonumber
+K_i &= (\lambda_i - 1) \alpha \\ \nonumber
+\lambda_i &= eigval(\mathbf{A}_i)
+\end{align}
+
+##### Code
+
+Input:
+
+- `A`: an $N$x$M$x$M$ array containing the $M$x$M$ transition matrices for all sites
+- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
+
+```{r beverton_holt}
+analysis <- iterate_matrix(A)
+lambda <- analysis$lambda
+K <- (lambda - 1) * alpha
+Lambda <- log1pe(K)
+```
+
+### Ricker model
+
+An alternative population dynamic model with which to relate the intrinsic growth rate to relative abundance is the Ricker model, which for single (not stage-structured) population $n_{i}$ is:
+
+\begin{align}
+n_{i,t+1} &= n_{i,t} e^{\lambda_i - n_{i,t} / M}
+\end{align}
+
+where $M$ again denotes a density-dependence parameter relating the intrinsic growth rate to the carrying capacity via a linear function $K_i = \lambda_i M$.
+With positive $M$, $K_i$ does not need rectifying to yield the relative abundance.
+As with the Beverton-Holt model, we define a parameter $\alpha$ which combines the density-dependence and reporting rate, giving us $\Lambda_i = \lambda_i \alpha$.
+This relationship us used in the dynamic range model of @pagelForecastingSpeciesRanges2012.
+Whilst it has the advantage of not requiring rectification of the carrying capacity, it has the disadvantage that with large enough $\alpha$, this approach can yield large values of $\Lambda_i$ even if $\lambda_i$ is well below 1, in which case the species should not be able to persist at this site.
 
@@
 \Lambda_i &= \lambda_i \alpha \\ \nonumber
-\lambda_i &= eigval(\mathbf{A_i})
+\lambda_i &= eigval(\mathbf{A}_i)
 \end{align}
@@
 
-## Density-dependent vital rates
+### Stage-structured abundances
 
-Both the Ricker and Beverton-Holt approaches to the abundance relationship assume that the carrying capacity can be expressed as some function of the intrinsic growth rate.
-This is convenient since the intrinsic growth rate is a well-studied and simple to compute quantity of a temporally-static transition matrix $A_i$.
-However, a more ecologically realistic approach is to explicitly model the vital rates as being density-dependent - that is having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point.
+Where distribution data are disaggregated according to discrete stages considered in the population transition matrix, the corresponded relative abundance $\Lambda_{j,i}$ for each stage $j$ can be calculated by computing the overall relative abundance $\Lambda^*_{j,i}$ via one of the two models above, and then disaggregating it into the $M$ stages using the stable stage distribution vector $\mathbf{v}_i$: the dominant (left) eigenvector $\mathbf{v}^*_i$ normalised to sum to 1.
+
+An approach similar to this can also used be used to compute relative abundances from a static transition matrix for a  matrix metapopulation model - in that model the stable distribution between sites already provides a relative abundance estimate, so there is no need for a model to relate the intrinsic growth rate to relative abundance.
+See [*Dispersal with a static transition matrix*](#static_dispersal) for details.
+
+##### Notation
+
+This notation uses the Beverton-Holt abundance model to compute the overall relative abundance parameter $\Lambda^*_i$, but the Ricker model could be used instead.
+
+\begin{align}
+\Lambda_{j,i} &= \Lambda^*_i \mathbf{v}_{j,i} \\ \nonumber
+\mathbf{v}_{j,i} &= \mathbf{v}^*_{j,i} / \textstyle\sum_{j=1}^{M}{\mathbf{v}^*_{j,i}} \\ \nonumber
+\mathbf{v}^*_i &= eigvec(\mathbf{A}_i) \\ \nonumber
+\Lambda^*_i &= ln(1 + e^{K_i})  \\ \nonumber
+K_i &= (\lambda_i - 1) \alpha \\ \nonumber
+\lambda_i &= eigval(\mathbf{A}_i)
+\end{align}
+
+where $eigvec(\mathbf{A}_i)$ gives the dominant left eigenvector of the population transition matrix $\mathbf{A}_i$.
+
+##### Code
+
+- `A`: a 3D $N$x$M$x$M$ array giving the population transition matrix at each site
+- `alpha`: a (positive) scalar for the combined effect of imperfect detection and density dependence
+
+```{r stage_structured}
+analysis <- iterate_matrix(A)
+lambda <- analysis$lambda
+v <- analysis$stable_abundance_distribution
+K <- (lambda - 1) * alpha
+Lambda_star <- log1pe(K)
+Lambda <- Lambda_star * v
+```
+
+## Dynamic transition matrices
+
+Analysis of matrix population models with static transition matrices assumes both that the population is at equilibrium, and that the dynamics can be adequately by an intrinsic growth rate and stable stage distribution.
+Where those assumptions are incompatible with the study system, data or questions being being asked, an alternative is to specify the matrix population model with a dynamic transition matrix $\mathbf{A}_{i,t}$, which differs between times $t$:
+
+\begin{align}
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t}
+\end{align}
+
+In this model the population transition matrix is recomputed at each iteration $t$, and can change between time points depending on fluctutating environmental conditions or the species' population density.
+Because the dynamic transition matrix is constantly changing, the dynamics cannot be represented by an intrinsic growth rate or stable stage distribution.
+Instead, iteration of the matrix explicitly considers population sizes.
+A separate model relating the transition matrix to a carrying capacity is no longer needed, since the vector of $\mathbf{n}_{i,t}$ of population sizes can be directly converted into a relative abundance.
+
+Matrix models with static population transition matrices implicitly assume density-independent growth (or decline) of population sizes.
+In the static models above, density dependence is introduced by the abundance model.
+This is problematic when explicitly considering the population sizes $\mathbf{n}_{i,t}$, since all populations will tend to either infinity or zero as $t$ increases.
+When considering a dynamic transition matrix it is therefore necessary to consider negative density-dependence in calculation of $\mathbf{A}_{i,t}$, so that the simulated dynamics are not explosive.
+
+### Density-dependent vital rates - equilibrium solution
+
+In static transition models considered above, carrying capacity is modelled as some function of the intrinsic growth rate, an relationship which implicitly represents density dependence.
+Rather than considering that density dependence isa  function of the growth rate, it may be more ecologically realistic to explicitly model the vital rates as being density-dependent - that is having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point.
 This enables incorporation of more detailed ecological knowledge about how vital rates drive abundance, which may be important to consider if the answer to the ecological question of interest depends on density dependence.
 
-Density dependent vital rates can be incorporated in a DSDM by explicitly considering a vector $\mathbf{n}_{i,t}$ of population densities for each of $M$ life stages (indexed with $j$) at site $i$, at time $t$.
-$\mathbf{n}_{i,t}$ is evolved through time by multiplication against a time-point-specific transition matrix $\mathbf{A}_{i,t}$, constructed of time-point-specific vital rates that depend on the total population density (i.e.
-across all life stages $M$) in the previous time point $\mathbf{N}_{i,t}$.
-For example, these density-dependent relationships could be a linear function of population density (controlled by parameters $\alpha_F$ and $\alpha_S$) on the link-scale of the vital rates.
-Since the transition matrix is no longer temporally-static, the population dynamics cannot be summarised by a population growth rate.
-Instead, iteration of the matrix using the density-dependent vital rate functions causes the vector $\mathbf{n}_{i,t}$ of population densities to converge on a static value after a sufficient number of iterations, $T$.
-This terminal density can be used directly as the relative abundance.
-As in the basic DSDM, the parameters $\alpha_F$ and $\alpha_S$ control both density-dependence and the detection rate, since these two processes cannot be differentiated without population growth data.
+Density dependent vital rates can be incorporated in a DSDM by constructing the dynamic population transition matrix $\mathbf{A}_{i,t}$, from dynamic vital rates that are computed as a function of both the environment and the total population (i.e.
+across all life stages $M$) in the previous time point.
+For example, these density-dependent relationships could be a linear function of population density - controlled by parameters $\alpha_F$ and $\alpha_S$ - on the link-scale of the vital rates.
+Where $\alpha_F$ and $\alpha_S$ are negative, density-dependence will lead to self-regulating populations that will ultimately converge on a finite carrying capacity.
+
+Starting with an initial population size vector $\mathbf{n}_{i,0}$, the dynamics can be iterated $T$ times to solve the for the population size at that time $\mathbf{n}_{i,T}$.
+With self-regulating density-dependence (negative feedback), as $T$ increases $\mathbf{n}_{i,T}$ will approach a stable population size.
+If $T$ is selected to be sufficiently large that population sizes have stopped changing between iterations (to within some numerical tolerance), $\mathbf{n}_{i,T}$ can be interpreted as the equilibrium relative abundance.
+As in the static transition matrix models, the parameters $\alpha_F$ and $\alpha_S$ control both the strength of density-dependence and the detection rate, since these parameters are not identified separately.
 
@@
 \Lambda_i &=  \mathbf{n}_{i, T} \\ \nonumber
-\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}_{i,t} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
 ln(F_{i,t}) &= \mathbf{X}_i \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
@@
 \end{align}
+where $\alpha_F$ and $\alpha_S$ are negative scalar parameters controlling the strength of density dependenceadnt he detection rate, and $T$ is positive integer giving a number of iterations chosen such that the population sizes have converged to withing a tolerance, i.e.
+$|\mathbf{n}_{i,T} - \mathbf{n}_{i,T-1}| / \mathbf{n}_{i,T-1} < \epsilon$ for some small positive tolerance $\epsilon$.
+This criterion means that the value of the initial population sizes $\mathbf{n}_{i,0}$ do not impact upon the model.
 
@@
 
-## Long-term temporal variation
+### Density-dependent vital rates - non-equilibrium solution
 
-Given temporally-indexed response data $y_{i,t}$, temporal variation in environmental conditions and therefore vital rates can be included in a DSDM in one of two ways.
-The applicability of each of these depends on both the temporal scale of the data (the period of time between the observations) and the demographics of the species.
-
-If the length of time between observations is on the order of multiple generations of the species population, it may be reasonable to assume that the observed population sixes are reflective of contemporary environmental conditions only, and not past fluctuations in environmental conditions.
-In other words, that the population is at, or close to, the long-term carrying capacity of the species if the current environmental conditions were to be held static.
-In this situation, considering temporal change in environmental conditions is simply a matter of redefining the index $i$ in the basic DSDM to index unique combinations of time and site, instead of just site.
-The relative abundance parameter for each of these observations is therefore computed from the intrinsic growth rate of the transition matrix at the corresponding environmental conditions.
-
-## Short-term temporal variation
-
-Where the time period between observations is sufficiently short that the assumptions above are not met, or when the ecological question being answered requires it, temporal fluctuation in vital rates can instead be modelled by explicitly iterating the population through successive time-steps, at each time step using the corresponding transition matrix, as in the density-dependent vital rate model above.
-Instead of iterating for some arbitrary number of iterations $T$ and taking the final state, instead the population is iterated over time-steps $t$ corresponding to time points in the dataset (and all intermediate time-steps).
-Thus the population at each time-step will be a function not only of the current environmental conditions, but of all previous environmental conditions - capturing the impacts of environmental stochasticity.
+Where the data or questions of the analysis are such that equilibrium population sizes are not representative of the population, the density-dependent vital rate model above can be iterate for predetermined number of timepoints $T$, such that each time $t$ corresponds to an observed time in the distribution data.
+For example, where the impacts of short-term variation in environmental conditions (environmental stochasticity) on population sizes is important, the transition matrix $\mathbf{A}_{i,t}$ at each time can be computed as a function of the design matrix of observed environmental conditions $\mathbf{X}_t$ at that time, and the current population density.
+The resulting population size $\mathbf{n}_{i,t}$ at each time $t$ can then be used in an observation model for the observed distribution data at that time $y_{i,t}$.
 
 Under this formulation, the population size parameters $\mathbf{n}_{i,t}$ will be affected by the initial population state $\mathbf{n}_{i,0}$.
-this can either be introduced as an unknown parameter in the model, or set to an arbitrary value, and adding a sufficient 'burn-in' period prior to the first observation such that its value has no little effect on the estimated model parameters.
-
-This model does not use static transition matrices and therefore cannot describe density dependence as a simple function of an intrinsic growth rate.
-It should therefore be combined with an explicit consideration of density-dependent vital rates as in the model above so that the population sizes $\mathbf{n}_{i,t}$ tend toward a long-term carrying capacity, rather than towards zero or infinity.
+This can either be introduced as an unknown parameter in the model, or set to an arbitrary value but with the considered time period set so that there is a sufficiently long 'burn-in' period prior to the first observation that its value has little effect on the estimated model parameters.
 
@@
 \Lambda_{i,t} &=  \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{A}_{i,t+1} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
+\mathbf{n}_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{A}_{i,t} &= \begin{bmatrix} F_{i,t} \cdot S_{J_{i,t}} & F_{i,t} \cdot S_{A_{i,t}} \\ S_{J_{i,t}} & S_{A_{i,t}}  \end{bmatrix}  \\ \nonumber
 ln(F_{i,t}) &= \mathbf{X}_{i,t} \beta_F + \alpha_F \textstyle\sum_{j=1}^M{\mathbf{n}_{i,j,t}}\\ \nonumber
@@
 
-## Dispersal with a static transition matrix
+## Basic matrix population model
 
-Dispersal of individuals between sites can be incorporated by expanding the matrix population model to simultaneously consider populations in all patches, with dispersal between patches quantified by a dispersal kernel.
+The two-stage static matrix population model considered in the manuscript represents the simplest form of DSDM.
+The transition matrix $\mathbf{A}_i$ describes the fundamental dynamics of the population at location $i$, with cells expressing the rates of transition from each stage at the previous time step (columns) to each stage in the present time step (rows).
+Thus for each adult in the previous time step $F_i \cdot S_{A_i}$ juveniles are contributed to the population in the current time step: each adult survives with probability $S_{A_i}$, and produces an average of $F_i$ offspring, with the product of these giving the expected contribution per adult.
+For more details of matrix population models such as this, see [@caswellMatrixPopulationModels2001], or any number of papers in the matrix population modelling literature.
+
+Each of these vital rates at each site $i$ is given by a linear model against a design matrix $\mathbf{X}_i$ representing environmental covariates, an intercept (coded as a column of all ones), and any transformations of those covariates
+
+##### Notation
+
+\begin{align}
+\mathbf{A}_i &= \begin{bmatrix} F_i \cdot S_{J_i} & F_i \cdot S_{A_i} \\ S_{J_i} & S_{A_i}  \end{bmatrix} \\ \nonumber
+ln(F_i) &= \mathbf{X}_i\beta_F \\ \nonumber
+logit(S_{A_i}) &= \mathbf{X}_i\beta_S \\ \nonumber
+logit(S_{J_i}) &= logit(S_{A_i}) - \gamma
+\end{align}
+
+##### Code
+
+Input:
+
+- `X`: a $N$x$K$ design matrix for vital rate regression sub-models
+- `beta_F`: a vector of length $K$ of regression coefficients for log-fecundity
+- `beta_s`: a vector of length $K$ of regression coefficients for logit-survival
+- `gamma`: a (positive) scalar for the difference in logit-survival between adults and juveniles
+
+```{r basic_transition}
+log_F <- X %*% beta_F
+logit_S_A <- X %*% beta_S
+F <- exp(log_F)
+S_A <- ilogit(logit_S_A)
+S_J <- ilogit(logit_S_A - gamma)
+
+top <- cbind(F * S_J, F * S_A)
+bottom <- cbind(S_J, S_A)
+A <- abind(top, bottom, along = 3)
+```
+
+## Dispersal with a static transition matrix {#static_dispersal}
+
+Dispersal of individuals between sites can be incorporated into a static transition matrix by expanding the matrix population model to a spatially-structured matrix that simultaneously considers populations in all patches, with dispersal between patches quantified by a dispersal kernel.
 
 In previous models, which assumed sites to be independent of one another with respect to their population dynamics, the 3D $N$x$M$x$M$ array $A$ contained the $M$x$M$ transition matrices $\mathbf{A}_i$ for each of the $N$ sites $i$.
-To consider interacting population dynamics across sites, we can instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$, with a symmetric blocked structure.
-The the $i^{th}$ block-diagonal element of $\mathbf{A}^*$ is the site-specific $M$x$M$ population transition matrix $\mathbf{A}_i$ and the off-diagonal blocks are $M$x$M$ diagonal matrices with diagonal entries giving the probability $d_{j,i,i'}$ that an individual of life stage $j$ in site $i$ will move to site $i'$ in this time-step.
-$d_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area and ensure $d_{j,i,i'}$ is a valid probability.
+To consider interacting population dynamics across sites, we can instead consider a single, larger transition matrix $\mathbf{A}^*$ with dimension $NM$x$NM$.
+This matrix combines two $NM$x$NM$ block-diagonal matrices: $\mathbb{A}$ - with each of the $N$ diagonal blocks giving the $M$x$M$ population transition matrix at site $i$ $\mathbf{A}_i$; and $\mathbb{D}$ - with each of the $M$ diagonal blocks giving an $N$x$N$ between-site dispersal matrix $\mathbf{D}_j$ for stage $j$.
+$\mathbf{A}^*$ is formed as $\mathbf{A}^* = \mathbf{P}^T \mathbb{D}\mathbf{P} \mathbb{A}$, where $\mathbf{P}$ is the vec-permutation matrix described in @hunterUseVecpermutationMatrix2005, which composes the two component matrices in the appropriate way.
+Note that this formulation assumes that in each season demographic transitions occur, then dispersal.
+For the reverse order, the matrix would be computed as $\mathbf{A}^* = \mathbf{P} \mathbb{A} \mathbf{P}^T \mathbb{D}$.
+
+Each element of the stage-dispersal matrices $\mathbf{D}_j$ give the probability $D_{j,i,i'}$ that an individual of  stage $j$ in site $i$ will move to site $i'$ in this time-step.
+$D_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area to ensure $D_{j,i,i'}$ is a valid probability.
 
@@
 \begin{align}
-\Lambda_i &= \alpha v_i / \textstyle\sum_{i=1}^{N}{v_{i}} \\ \nonumber
-v_i &= \textstyle\sum_{j=1}^{M}{v^*_{j+(i-1)}} \\ \nonumber
-v^* &= eigvec(\mathbf{A}^*) \\ \nonumber
-\mathbf{A}^* &= \begin{bmatrix} \mathbf{A}_1 & \mathbf{M}_{1,2} & \dots & \mathbf{M}_{1,i'} & \dots & \mathbf{M}_{1,N}\\
-\mathbf{M}_{2,1} & \mathbf{A}_2 & \dots & \mathbf{M}_{2,i'} & \dots & \mathbf{M}_{2,N} \\
-\vdots & \vdots & \ddots & \vdots & &  \vdots \\
-\mathbf{M}_{i,1} & \mathbf{M}_{i,2} & \dots & \mathbf{A}_i & \dots & \mathbf{M}_{i,N} \\
-\vdots & \vdots &  & \vdots & \ddots & \vdots \\
-\mathbf{M}_{N,1} & \mathbf{M}_{N,2} & \dots & \mathbf{M}_{N,i'} & \dots & \mathbf{A}_N \end{bmatrix} \\ \nonumber
-\mathbf{M}_{i,i'} &= diag(d_{1,i,i'}, \dots, d_{j,i,i'}, \dots, d_{M,i,i'}) \\ \nonumber
-d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
-d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
+\Lambda_i &= \alpha \mathbf{v}_i / \textstyle\sum_{i=1}^{N}{\mathbf{v}_{i}} \\ \nonumber
+\mathbf{v}_i &= \textstyle\sum_{j=1}^{M}{\mathbf{v}^*_{j+(i-1)}} \\ \nonumber
+\mathbf{v}^* &= eigvec(\mathbf{A}^*) \\ \nonumber
+\mathbf{A}^* &= \mathbf{P} \mathbb{A} \mathbf{P}^T \mathbb{D} \\ \nonumber
+\mathbb{A} = \begin{bmatrix} \mathbf{A}_1 & & \dots & & 0 \\
+& \ddots & & & \\
+\vdots & & \mathbf{A}_i & & \vdots \\
+& & & \ddots & \\
+0 & & \dots & & \mathbf{A}_N \end{bmatrix}
+&\,\, \mathbb{D} = \begin{bmatrix} \mathbf{D}_1 & & \dots & & 0 \\
+& \ddots & & & \\
+\vdots & & \mathbf{D}_j & & \vdots \\
+& & & \ddots & \\
+0 & & \dots & & \mathbf{D}_M \end{bmatrix} \\ \nonumber
+\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
+d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
 \end{align}
+where $D$ is an $N$x$N$  dense, symmetric matrix of Euclidean distances between all sites.
 
@@
 Where the transition matrix is assumed to be dynamic - i.e..
-to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we can use the same model structure as for a static transition mat, but combine it with the iterative approach used for dynamic transition matrix models above.
-Because the matrix is explicitly iterated dispersal can be incorporated with a second (dense) multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
+to model the effects of environmental stochasticity, density-dependent vital rates, or population time series - we can use the same model structure as for a static transition matrix, but combine it with the iterative approach used for dynamic transition matrix models above.
+Because the matrix is explicitly iterated dispersal can be incorporated with a second matrix multiplication rather than needing to combine the population transition and dispersal matrices into a single block matrix.
 As for other dynamic transition matrix models, density-dependent vital rates are required to compute the relative abundances, since neither an intrinsic growth rate nor a stable population distribution are available in the dynamic matrix model.
 
-I n the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each life-stage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
+In the notation below, I introduce an intermediate $N$x$M$ population state matrix $\mathbf{n}^*_{t+1}$ for the sizes of the populations in each life-stage at each site, between iteration of the population transition matrix (breeding) and iteration of the dispersal matrix (dispersal), which both occur between each successive time steps.
 For simplicity of notation, I assume here that breeding happens before dispersal, but the order of the operations can be reversed as required.
@@
 \begin{align}
-\mathbf{n}_{j,t+1} &= \mathbf{d}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
-\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t+1} \mathbf{n}_{i,t} \\ \nonumber
-d_{j,i,i'} &= d^*_{j,i,i'} / \textstyle\sum_{i=1}^Nd^*_{j,i,i'} \\ \nonumber
-d^*_{j,i,i'} &=  e^{-\tau_j \mathbf{D}_{i,i'}}
+\mathbf{n}_{j,t+1} &= \mathbf{D}_j \mathbf{n}^*_{j,t+1} \\ \nonumber
+\mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
+\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
+d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
 \end{align}
-
-where $\mathbf{D}$ is an $N$x$N$ matrix of distances between sites, used to define $N$x$N$ matrix $\mathbf{d}_j$ of dispersal probabilities for life stage $j$, which sums to 1 across each column.
 
@@
 Note that this is just one approach to considering multiple genotypes and natural selection in a matrix population model, see e.g.
-[@devriesSelectionTwosexStagestructured2019].
+@devriesSelectionTwosexStagestructured2019.
 
@@
 \begin{align}
-\Lambda_{l,i} &= \alpha \Lambda^*_i v_{l,i} / \textstyle\sum_{l=1}^{L}{v_{l,i}} \\ \nonumber
+\Lambda_{l,i} &= \Lambda^*_i \mathbf{v}_{l,i} / \textstyle\sum_{l=1}^{L}{\mathbf{v}_{l,i}} \\ \nonumber
 \Lambda^*_i &= ln(1 + e^{K_i})  \\ \nonumber
 K_i &= (\lambda_i - 1) \alpha \\ \nonumber
-v_{l,i} &= \textstyle\sum_{j=1}^{M}{v^*_{j+(l-1),i}} \\ \nonumber
-v^*_i &= eigvec(\mathbf{A}^*_i) \\ \nonumber
+\mathbf{v}_{l,i} &= \textstyle\sum_{j=1}^{M}{\mathbf{v}^*_{j+(l-1),i}} \\ \nonumber
+\mathbf{v}^*_i &= eigvec(\mathbf{A}^*_i) \\ \nonumber
 \lambda_i &= eigval(\mathbf{A}^*_i) \\ \nonumber

### 2019-10-25 manuscript/appendix_a.Rmd
starrt adding IPM code

@@
 
+##### Notation
+
+\begin{align}
+a
+\end{align}
+
+##### Code
+
+Input:
+
+- `matrix_dim`: a positive integer giving the dimension of the matrix used to integrate the IPM kernel
+- `X_off_size`: a $N$x$K_{off_size}$ design matrix for the offspring size environmental sub-model
+- `X_surv`: a $N$x$K_{surv}$ design matrix for the survival environmental sub-models
+- `X_growth`: a $N$x$K_{grow}$ design matrix for the growth environmental sub-model
+- `X_seedheads`: a $N$x$K_{seedhead}$ design matrix for the seedhead environmental sub-model
+- `beta_env_off_size`: a length $K_{off_size}$ vector of environmental regression coefficients for offspring size
+- `beta_env_surv_seedling`: a length $K_{surv}$ vector of environmental regression coefficients for logit-seedling-survival
+- `beta_env_surv_adult`: a length $K_{surv}$ vector of environmental regression coefficients for logit-adult-survival
+- `beta_env_growth`: a length $K_{grow}$ vector of environmental regression coefficients for growth
+- `beta_env_seedheads`: a length $K_{seedhead}$ vector of environmental regression coefficients for log-seadhead production
+- `beta_size_surv_seedling`: ?
+- `beta_size_surv_adult`: ?
+- `beta_size_grow`: ?
+- `beta_size_flower`: ?
+- `beta_germ`: ?
+- `beta_seeds`: ?
+
+```{r ipms}
+min_size <- 0
+max_size <- 5
+bounds <- min_size + (0:matrix_dim) * (max_size - min_size) / matrix_dim
+sizes <- 0.5 * (bounds[-(matrix_dim + 1)] + bounds[-1])
+bin_width <- sizes[2] - sizes[1]
+
+# environmental components of sub models
+off_size <- X_off_size %*% beta_env_off_size
+growth <- X_growth %*% beta_env_growth
+env_surv_seedling <- X_surv %*% beta_env_surv_seedling
+env_surv_adult <- X_surv %*% beta_env_surv_adult
+env_seedheads <- X_seedheads %*% beta_env_seedheads
+
+# size components of submodels
+flower <- ilogit(sizes * beta_size_flower)
+size_surv_seedling <- sizes * beta_size_surv_seedling
+size_surv_adult <- sizes * beta_size_surv_adult
+size_seedheads <- sizes * beta_size_seedheads
+
+# combined size and environmental effects
+seedheads <- exp(kronecker(size_seedheads, env_seedheads))
+surv_seedling <- ilogit(kronecker(size_surv_seedling, env_surv_seedling))
+surv_adult <- ilogit(kronecker(size_surv_adult, env_surv_adult))
+
+germination <- ilogit(beta_germ)
+seeds <- exp(beta_seeds)
+
+# kernelise growth and offspring size
+growth_density <- distribution_kernel(grow, grow_sd, sizes, bin_width)
+off_size_density <- distribution_kernel(off_size, off_size_sd, sizes, bin_width)
+
+# combine adult and seedling survival rates into one
+survival <- interpolate_survival(surv_adult, surv_seedling, sizes)
+
+# apply survival to growth density
+P <- sweep_3(growth_density, survival, FUN = "*")
+
+# apply recruitment rate to offspring size density
+recruits <- flower * seedheads * seeds * germ
+F <- sweep_3(offspring_size_density, recruits, FUN = "*")
+
+A <- P + F
+```
+
 ## Biotic interactions

### 2019-10-30 manuscript/appendix_a.Rmd
complete IPM code

--- 2019-10-25 manuscript/appendix_a.Rmd
@@
 
+##### Notation
+
+\begin{align}
+a
+\end{align}
+
+##### Code
+
+The below code requires some custom functions and a custom greta operation to perform the integration steps and build the kernels efficiently across 3D arrays representing kernels across multiple sites, and to carry out the interpolation of survival rates between adults and seedlings that is implemented in @merowUsingIntegralProjection2014.
+
+Input:
+
+- `matrix_dim`: a positive integer giving the dimension of the matrix used to integrate the IPM kernel
+- `X_offsize`: a $N$x$K_{\text{offsize}}$ design matrix for the offspring size environmental sub-model
+- `X_surv`: a $N$x$K_{\text{surv}}$ design matrix for the survival environmental sub-models
+- `X_grow`: a $N$x$K_{\text{grow}}$ design matrix for the growth environmental sub-model
+- `X_seedheads`: a $N$x$K_{\text{seedhead}}$ design matrix for the seedhead environmental sub-model
+- `beta_env_offsize`: a length $K_{\text{offsize}}$ vector of environmental regression coefficients for offspring size
+- `beta_env_surv_seedling`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-seedling-survival
+- `beta_env_surv_adult`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-adult-survival
+- `beta_env_grow`: a length $K_{\text{grow}}$ vector of environmental regression coefficients for growth
+- `beta_env_seedheads`: a length $K_{\text{seedhead}}$ vector of environmental regression coefficients for log-seadhead production
+- `beta_size_surv_seedling`: a scalar coefficient for the effect of size on logit-seedling-survival
+- `beta_size_surv_adult`: a scalar coefficient for the effect of size on logit-adult-survival
+- `beta_size_grow`: a scalar coefficient for the effect of size on growth
+- `beta_size_flower`: a scalar coefficient for the effect of size on logit-flowering-probability
+- `alpha_germ`: a scalar parameter for logit-germination-probability
+- `alpha_seeds`: a scalar parameter for log-seed-production
+- `alpha_flower`: a scalar intercept parameter for on logit-flowering-probability
+- `offsize_sd`: a positive scalar parameter for the standard deviation of the offspring size distribution
+- `grow_sd`: a positive scalar parameter for the standard deviation of the growth distribution
+
+```{r ipms}
+# normalise a matrix to sum to one rowwise *in exponential space*. Ie. return y
+# where: exp(y[i, ]) = exp(x[i, ]) / sum(exp(x[i, ]))
+log_normalise <- function (x) {
+  op <- greta::.internals$nodes$constructors$op
+  op("log_normalise", x, tf_operation = "tf_log_normalise")
+}
+
+# tensorflow code for log normalisation
+tf_log_normalise <- function (x) {
+  sums <- tensorflow::tf$reduce_logsumexp(x, axis = 2L, keepdims = TRUE)
+  x - sums
+}
+
+# multiply matrix B by each slice of the 3D array A in 3rd dimension
+sweep_prod_3d <- function(A, B) {
+  n <- dim(A)[1]
+  m <- dim(A)[2]
+  Al <- aperm(A, c(2, 3, 1))
+  dim(Al) <- c(m, n * m)
+  Bl <- t(B)
+  dim(Bl) <- n * m
+  C <- sweep(Al, 2, Bl, "*")
+  dim(C) <- c(m, m, n)
+  aperm(C, c(3, 1, 2))
+}
+
+# create 3D kernel array for Gaussian transitions (working with the log density
+# for computational stability)
+distribution_kernel <- function (mean_mat, sd, sizes, bin_width) {
+  var <- sd ^ 2
+  log_constant <- -0.5 * log(2 * pi) - log(sd)
+  diff <- kronecker(sizes, mean_mat, FUN = "-")
+  log_density <- log(bin_width) + log_constant - (diff ^ 2) / (2 * var)
+  log_density <- log_normalise(log_density)
+  dim(log_density) <- c(dim(mean_mat), ncol(mean_mat))
+  log_density <- aperm(log_density, c(1, 3, 2))
+  exp(log_density)
+}
+
+interpolate_survival <- function (adult_surv, seedling_surv, sizes) {
+  range <- 0.728 - 0.182
+  weights <- 0.728 / range  - (sizes / range)
+  weights[sizes > 0.728] <- 0
+  weights[sizes <= 0.182] <- 1
+  weighted_seedling_surv <- sweep(seedling_surv, 2, weights, FUN = "*")
+  weighted_adult_surv <- sweep(adult_surv, 2, 1 - weights, FUN = "*")
+  weighted_seedling_surv + weighted_adult_surv
+}
+
+min_size <- 0
+max_size <- 5
+bounds <- min_size + (0:matrix_dim) * (max_size - min_size) / matrix_dim
+sizes <- 0.5 * (bounds[-(matrix_dim + 1)] + bounds[-1])
+bin_width <- sizes[2] - sizes[1]
+
+# environmental components of sub models
+offsize <- X_offsize %*% beta_env_offsize
+grow <- X_grow %*% beta_env_grow
+env_surv_seedling <- X_surv %*% beta_env_surv_seedling
+env_surv_adult <- X_surv %*% beta_env_surv_adult
+env_seedheads <- X_seedheads %*% beta_env_seedheads
+
+# size components of submodels
+flower <- ilogit(alpha_flower + sizes * beta_size_flower)
+size_surv_seedling <- sizes * beta_size_surv_seedling
+size_surv_adult <- sizes * beta_size_surv_adult
+size_seedheads <- sizes * beta_size_seedheads
+
+# vital rates at all sites
+seedheads <- exp(kronecker(size_seedheads, env_seedheads))
+surv_seedling <- ilogit(kronecker(size_surv_seedling, env_surv_seedling))
+surv_adult <- ilogit(kronecker(size_surv_adult, env_surv_adult))
+germination <- ilogit(alpha_germ)
+seeds <- exp(alpha_seeds)
+
+# kernelise growth * survival and recruitment * offspring size, and combine
+survival <- interpolate_survival(surv_adult, surv_seedling, sizes)
+next_size_density <- distribution_kernel(sizes + grow, grow_sd, sizes, bin_width)
+P <- sweep_prod_3d(next_size_density, survival)
+recruits <- flower * seedheads * seeds * germ
+offsize_density <- distribution_kernel(offsize, offsize_sd, sizes, bin_width)
+F <- sweep_prod_3d(offsize_density, recruits)
+A <- P + F
+```
+
 ## Biotic interactions

### 2019-10-30 manuscript/appendix_a.Rmd
add notation and description for IPMs

--- 2019-10-25 manuscript/appendix_a.Rmd
@@
 
+\newcommand{\smallunderscore}{\textscale{.5}{\textunderscore}}
+
 ```{r knitrOpts, echo = FALSE, cache = FALSE, eval = TRUE}
@@
 
-Each element of the stage-dispersal matrices $\mathbf{D}_j$ give the probability $D_{j,i,i'}$ that an individual of  stage $j$ in site $i$ will move to site $i'$ in this time-step.
-$D_{j,i,i'}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area to ensure $D_{j,i,i'}$ is a valid probability.
+Each element of the stage-dispersal matrices $\mathbf{D}_j$ give the probability $D_{j,i',i}$ that an individual of  stage $j$ in site $i$ will move to site $i'$ in this time-step.
+$D_{j,i',i}$ is computed using a dispersal kernel (an exponential kernel in this example) with stage-specific dispersal ability parameter $\tau_j$, standardised to sum to 1 over the study area to ensure $D_{j,i',i}$ is a valid probability.
 
@@
 In practice, the matrix $\mathbf{A}^*$ can be very sparse: the proportion of non-zero elements is $(M + N - 1) / (MN)$; as N increases, the proportion of non-zero approaches $1/M$.
-This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i,i'}$ is set to zero, or by limiting the number of stages that can disperse.
+This sparsity can be increased further by setting an upper dispersal limit, beyond which $d_{j,i',i}$ is set to zero, or by limiting the number of stages that can disperse.
 It is therefore computationally advantageous to compute the dominant eigenvector using sparse matrix methods, e.g.
@@
 0 & & \dots & & \mathbf{D}_M \end{bmatrix} \\ \nonumber
-\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
-d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
+\mathbf{D}_{j,i',i} &= d_{j,i',i} / \textstyle\sum_{i'=1}^Nd_{j,i',i} \\ \nonumber
+d_{j,i',i} &=  e^{-\tau_j D_{i',i}}
 \end{align}
@@
 \mathbf{n}^*_{i,t+1} &= \mathbf{A}_{i,t} \mathbf{n}_{i,t} \\ \nonumber
-\mathbf{D}_{j,i,i'} &= d_{j,i,i'} / \textstyle\sum_{i=1}^Nd_{j,i,i'} \\ \nonumber
-d_{j,i,i'} &=  e^{-\tau_j D_{i,i'}}
+\mathbf{D}_{j,i',i} &= d_{j,i',i} / \textstyle\sum_{i=1}^Nd_{j,i',i} \\ \nonumber
+d_{j,i',i} &=  e^{-\tau_j D_{i',i}}
 \end{align}
@@
 
-<to do, don't package up integration code into greta.dynamics yet, show it warts-and-all for now>
+For species with complex demography and populations that are better represented as size-structureed rather than stage-structured DSDMs can be formed by replacing the MPM with an integral projection model (IPM).
+As an example, this section presents the equivalent code and notation for the complex, environmentally-explicit spatial IPM for *Protea repens* presented in @merowUsingIntegralProjection2014 and studied as a DSDM in the manuscript.
+This model is particularly complex - having 8 vital rate submodels, and requiring interpolation of survival rates betwen seedlings and adults.
+
+Whereas matrix population models consider populations with a discrete population distribution over different stages in the model, integral projection models consider populations via a continuous population distribution over individual sizes.
+The dynamics of a population at site $i$ are therefore modelled as $\mathbf{n}_{i,t+1}(z') = \int \mathbf{K}_{i}(z'|z)\mathbf{n}_{i,t}(z) \, dz$, where $\mathbf{n}_{i,t}(z)$ is an abundance *function* (at site $i$ and time $t$) of sizes $z$, giving the number of individuals of with size around $z$, and $\mathbf{K}_{i}(z'|z)$ is a continuous transition *kernel* at site $i$ mapping these abundance functions between time points.
+Whereas a matrix population model is described by a matrix-vector multiplication (which sums products between elements of the matrix and vector), the analagous operaiton in this continuous representation is an integration over a convolution.
+In practice though, this integration and convolution is commonly approximated by discretizing and normalising the kernel $\mathbf{K}_{i}(z'|z)$ to yield a matrix $\mathbf{A}_i$ with a sufficiently large number of size bins $j$ which can be analysed in the same way as a matrix population model.
+We can therefore compute such a matrix and include it in the DSDM framework.
+
+Whereas MPM-based DSDMs have a single environmental regression submodel for each vital rate, in the IPM these are replaced with submodels that combine the effects of both the environment and individual size.
+These functions must therefore be manipulated to create arrays (over sites and size bins) that are then combined and normalised to yield an approximation to the IPM.
+The discretized notation and code to implement these models is therefore somewhat more involved than for MPMs.
+
+##### Notation
+
+\begin{align}
+\mathbf{A}_i &= \mathbf{F}_i + \mathbf{P}_i \\ \nonumber
+\mathbf{F}_{i,j',j} &= p_{\text{flower}_{j}} f_{\text{seedheads}_i} f_{\text{seeds}} p_{\text{germ}} f^*_{\text{offsize}_{i,j'}} \\ \nonumber
+\mathbf{P}_{i,j',j} &= p_{\text{surv}_{i,j}} f^*_{\text{grow}_{i,j',j}} \\ \nonumber
+f^*_{\text{offsize}_{i,j'}} &= f_{\text{offsize}_{i,j'}} / \textstyle\sum_{j'=1}^{M}{f_{\text{offsize}_{i,j'}}} \\ \nonumber
+f^*_{\text{nextsize}_{i,j',j}} &= f_{\text{nextsize}_{i,j',j}} / \textstyle\sum_{j'=1}^{M}{f_{\text{nextsize}_{i,j',j}}} \\ \nonumber
+logit(p_{\text{flower}}) &= \alpha_{\text{flower}} + z \gamma_{\text{flower}} \\ \nonumber
+log(f_{\text{seedheads}}) &= \mathbf{X}_{\text{seedheads}} \beta_{\text{seedheads}} \\ \nonumber
+p_{\text{surv}_{i,j}} &= w_j p_{\text{surv-adult}_{i,j}} + (1 - w_j) p_{\text{surv-seedling}_i}  \\ \nonumber
+logit(p_{\text{surv-adult}_{i,j}}) &= \mathbf{X}_{\text{surv}_i} \beta_{\text{surv-adult}} + z_j \gamma_{\text{surv-adult}} \\ \nonumber
+logit(p_{\text{surv-seedling}_i}) &= \mathbf{X}_{\text{surv}_i} \beta_{\text{surv-seedling}} + z_j \gamma_{\text{surv-seedling}} \\ \nonumber
+f_{\text{offsize}_{i,j'}} &= \pi (z_{j'} | \mu_{\text{offsize}_{i,j'}}, \sigma_{\text{offsize}}^2) \\ \nonumber
+\mu_{\text{offsize}_{i,j'}} &= \mathbf{X}_{\text{offsize}_i} \beta_{\text{offsize}} \\ \nonumber
+f_{\text{nextsize}_{i,j',j}} &= \pi (z_{j'} | z_{j} + \mu_{\text{grow}_{i,j}}, \sigma_{\text{grow}}^2) \\ \nonumber
+\mu_{\text{grow}_{i,j}} &= \mathbf{X}_{\text{grow}_i} \beta_{\text{grow}} + z_{j} \gamma_{\text{grow}} \\ \nonumber
+\end{align}
+
+where $z$ is a vector of sizes of length $M$; $\mathbf{X}_{\text{seedheads}}$, $\mathbf{X}_{\text{surv}}$, $\mathbf{X}_{\text{offsize}}$, $\mathbf{X}_{\text{grow}}$, $\beta_{\text{seedheads}}$, $\beta_{\text{surv-adult}}$, $\beta_{\text{surv-seedling}}$, $\beta_{\text{offsize}}$, and $\beta_{\text{grow}}$ are design matrices of environmental covariates (and intercept columns) and vectors of regression coefficients for the environmental components of log-seedhead production, logit- adult and seedling survival probabilities, mean offspring size and mean growth rates; $\alpha_{\text{flower}}$, $\gamma_{\text{flower}}$, $\gamma_{\text{surv-adult}}$, $\gamma_{\text{surv-seedling}}$, and $\gamma_{\text{grow}}$ are scalar intercepts and regression coefficients for the effect of size on the logit-probabilities of flowering, adult survival, seedling survival, and on the growth rate; $f_{\text{seeds}}$ and $p_{\text{germ}}$ are scalar parameters for the average number of seeds per seedhead and the probability of germination; $w$ is a vector of weights of length $M$ linearly interpolating size-specific survival between adult and seedling survival probabilities; $\sigma_{\text{offsize}}$ and $\sigma_{\text{grow}}$ are positive scalar parameters giving the standard deviations of offspring sizes and growth rates; and $\pi (x | \mu, \sigma ^ 2)$ denotes the density of univariate normal distribution with parameters $\mu$ and $\sigma^2$ evaluated at $x$.
+
+##### Code
+
+The below code requires some custom functions and a custom greta operation to perform the integration steps and build the kernels efficiently across 3D arrays representing kernels across multiple sites, and to carry out the interpolation of survival rates between adults and seedlings that is implemented in @merowUsingIntegralProjection2014.
+
+Input:
+
+- `matrix_dim`: a positive integer giving $M$ - the dimension of the matrix used to integrate the IPM kernel
+- `X_offsize`: a $N$x$K_{\text{offsize}}$ design matrix for the offspring size environmental sub-model
+- `X_surv`: a $N$x$K_{\text{surv}}$ design matrix for the survival environmental sub-models
+- `X_grow`: a $N$x$K_{\text{grow}}$ design matrix for the growth environmental sub-model
+- `X_seedheads`: a $N$x$K_{\text{seedhead}}$ design matrix for the seedhead environmental sub-model
+- `beta_offsize`: a length $K_{\text{offsize}}$ vector of environmental regression coefficients for offspring size
+- `beta_surv_seedling`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-seedling-survival
+- `beta_surv_adult`: a length $K_{\text{surv}}$ vector of environmental regression coefficients for logit-adult-survival
+- `beta_grow`: a length $K_{\text{grow}}$ vector of environmental regression coefficients for growth
+- `beta_seedheads`: a length $K_{\text{seedhead}}$ vector of environmental regression coefficients for log-seadhead production
+- `gamma_surv_seedling`: a scalar coefficient for the effect of size on logit-seedling-survival
+- `gamma_surv_adult`: a scalar coefficient for the effect of size on logit-adult-survival
+- `gamma_grow`: a scalar coefficient for the effect of size on growth
+- `gamma_flower`: a scalar coefficient for the effect of size on logit-flowering-probability
+- `germination`: a scalar parameter in $(0,1)$ for germination-probability
+- `seeds`: a positive scalar parameter for seed-production
+- `alpha_flower`: a scalar intercept parameter for on logit-flowering-probability
+- `offsize_sd`: a positive scalar parameter for the standard deviation of the offspring size distribution
+- `grow_sd`: a positive scalar parameter for the standard deviation of the growth distribution
+
+```{r ipms}
+# normalise a matrix to sum to one rowwise *in exponential space*. Ie. return y
+# where: exp(y[i, ]) = exp(x[i, ]) / sum(exp(x[i, ]))
+log_normalise <- function (x) {
+  op <- greta::.internals$nodes$constructors$op
+  op("log_normalise", x, tf_operation = "tf_log_normalise")
+}
+
+# tensorflow code for log normalisation
+tf_log_normalise <- function (x) {
+  sums <- tensorflow::tf$reduce_logsumexp(x, axis = 2L, keepdims = TRUE)
+  x - sums
+}
+
+# multiply matrix B by each slice of the 3D array A in 3rd dimension
+sweep_prod_3d <- function(A, B) {
+  n <- dim(A)[1]
+  m <- dim(A)[2]
+  Al <- aperm(A, c(2, 3, 1))
+  dim(Al) <- c(m, n * m)
+  Bl <- t(B)
+  dim(Bl) <- n * m
+  C <- sweep(Al, 2, Bl, "*")
+  dim(C) <- c(m, m, n)
+  aperm(C, c(3, 1, 2))
+}
+
+# create 3D kernel array for Gaussian transitions (working with the log density
+# for computational stability)
+distribution_kernel <- function (mean_mat, sd, sizes, bin_width) {
+  var <- sd ^ 2
+  log_constant <- -0.5 * log(2 * pi) - log(sd)
+  diff <- kronecker(sizes, mean_mat, FUN = "-")
+  log_density <- log(bin_width) + log_constant - (diff ^ 2) / (2 * var)
+  log_density <- log_normalise(log_density)
+  dim(log_density) <- c(dim(mean_mat), ncol(mean_mat))
+  log_density <- aperm(log_density, c(1, 3, 2))
+  exp(log_density)
+}
+
+interpolate_survival <- function (adult_surv, seedling_surv, sizes) {
+  range <- 0.728 - 0.182
+  weights <- 0.728 / range  - (sizes / range)
+  weights[sizes > 0.728] <- 0
+  weights[sizes <= 0.182] <- 1
+  weighted_seedling_surv <- sweep(seedling_surv, 2, weights, FUN = "*")
+  weighted_adult_surv <- sweep(adult_surv, 2, 1 - weights, FUN = "*")
+  weighted_seedling_surv + weighted_adult_surv
+}
+
+min_size <- 0
+max_size <- 5
+bounds <- min_size + (0:matrix_dim) * (max_size - min_size) / matrix_dim
+sizes <- 0.5 * (bounds[-(matrix_dim + 1)] + bounds[-1])
+bin_width <- sizes[2] - sizes[1]
+
+# environmental components of sub models
+offsize <- X_offsize %*% beta_offsize
+grow <- X_grow %*% beta_grow
+env_surv_seedling <- X_surv %*% beta_surv_seedling
+env_surv_adult <- X_surv %*% beta_surv_adult
+env_seedheads <- X_seedheads %*% beta_seedheads
+
+# size components of submodels
+flower <- ilogit(alpha_flower + sizes * gamma_flower)
+size_surv_seedling <- sizes * gamma_surv_seedling
+size_surv_adult <- sizes * gamma_surv_adult
+size_seedheads <- sizes * gamma_seedheads
+
+# vital rates at all sites
+seedheads <- exp(kronecker(size_seedheads, env_seedheads))
+surv_seedling <- ilogit(kronecker(size_surv_seedling, env_surv_seedling))
+surv_adult <- ilogit(kronecker(size_surv_adult, env_surv_adult))
+
+# kernelise growth * survival and recruitment * offspring size, and combine
+survival <- interpolate_survival(surv_adult, surv_seedling, sizes)
+next_size_density <- distribution_kernel(sizes + grow, grow_sd, sizes, bin_width)
+P <- sweep_prod_3d(next_size_density, survival)
+recruits <- flower * seedheads * seeds * germ
+offsize_density <- distribution_kernel(offsize, offsize_sd, sizes, bin_width)
+F <- sweep_prod_3d(offsize_density, recruits)
+A <- P + F
+```
 

### 2019-10-30 R/fetch_data.R
try to cache downloads

@@ -1,5 +1,5 @@
 # download and unzip the RData object from Merow et al. (hosted on the Ecography data portal)
-download_merow <- function () {
+download_protea <- function (file) {
 
   # download and unzip
   url <- "http://www.ecography.org/sites/ecography.org/files/appendix/ecog-00839_appendix.zip"
@@ -45,13 +45,15 @@ download_merow <- function () {
       summer_smd = smdsum,
       winter_smd = smdwin,
       mean_annual_precip = map
-    )
+    ) -> protea_data
+
+  saveRDS(protea_data, file)
 
 }
 
 # get landcover and bioclim covariates for the BBS example and
 # remove extranous areas
-download_bbs_rasters <- function () {
+download_bbs_rasters <- function (file) {
 
   # get CarolineWren rasters from maxlike package and convert to a raster
   data(carw, envir = environment())
@@ -75,13 +77,14 @@ download_bbs_rasters <- function () {
   covs2 <- mask(covs2, covs1[[1]])
   covs <- stack(covs1, covs2)
 
+  writeRaster(covs, file = file, overwrite = TRUE)
   covs
-  # writeRaster(covs, file = "data/clean/bbs_covs")
+
 }
 
 # download the station names where this species was collected to derive vital
 # rate estimates, and the coordinates of those stations
-download_maps_locations <- function () {
+download_maps_locations <- function (file) {
 
   # get the data used to fit vital rates for the Carolina Chickadee
   cach_url <- "https://raw.githubusercontent.com/Akcakaya/MAPS-to-Models/master/Public%20dataset/CACH_BandDataAllMonths.csv"
@@ -98,7 +101,7 @@ download_maps_locations <- function () {
   stations_list <- xml2::as_list(stations_html)
 
   # pull out the section of the JS code with the station data
-  js_functions <- stations_list$html$head[[11]][[1]]
+  js_functions <- stations_list$html$head[[14]][[1]]
   js_function_split <- strsplit(js_functions, "\n")[[1]]
   circles <- js_function_split[32:1317]
 
@@ -119,6 +122,8 @@ download_maps_locations <- function () {
   idx <- match(names(station_weights), stations$id)
   stations <- stations[idx, ]
   stations$weight <- station_weights
+
+  saveRDS(stations, file)
   stations
 
 }
@@ -171,7 +176,7 @@ clean_bbs <- function(file, AOU, years = 2015) {
 }
 
[…]

### 2019-10-30 main.R

@@ -27,17 +27,10 @@ source("R/fetch_data.R")
 
 read_all_data <- drake_plan(
 
-  # bbs analysis
-  bbs_covs = download_bbs_rasters(),
-  tidy_covs = writeRaster(bbs_covs, file_out("data/clean/bbs_covs.grd"), overwrite = TRUE),
-  bbs_occurrence = download_bbs(),
-  tidy_occ = saveRDS(bbs_occurrence, file_out("data/clean/bbs_occ.RDS")),
-  maps_coords = download_maps_locations(),
-  tidy_maps = saveRDS(maps_coords, file_out("data/clean/maps_stations.RDS")),
-
-  # protea analysis
-  protea_occurrence = download_merow(),
-  tidy_file = saveRDS(protea_occurrence, file_out("data/clean/protea_occurrence.RDS"))
+  download_bbs_rasters(file_out("data/clean/bbs_covs.grd")),
+  download_bbs(file_out("data/clean/bbs_occ.RDS")),
+  download_maps_locations(file_out("data/clean/maps_stations.RDS")),
+  download_protea(file_out("data/clean/protea_occurrence.RDS")),
 
 )
 
@@ -84,12 +77,23 @@ fit_bbs_model <- drake_plan(
 make(fit_bbs_model)
 
 fit_protea_model <- drake_plan(
+
   protea_occurrence = readRDS(file_in("data/clean/protea_occurrence.RDS")),
-  model_list = build_protea_model(protea_occurrence),
-  draws_list = run_mcmc(model_list),
-  draws_summary = check_draws(draws_list, "protea"),
-  relationships_fig = plot_protea_relationships(model_list, draws_list, protea_occurrence),
-  predictions = make_protea_predictions(model_list, draws_list, protea_occurrence)
+
+  informative_model_list = build_protea_model(
+    protea_occurrence,
+    informative_priors = TRUE
+  ),
+  informative_draws_list = run_mcmc(informative_model_list, verbose = TRUE, sample = hmc(Lmin = 10, Lmax = 15)),
+  informative_draws_summary = check_draws(informative_draws_list, "protea_informative"),
+
+  naive_model_list = build_protea_model(
+    protea_occurrence,
+    informative_priors = FALSE
+  ),
+  naive_draws_list = run_mcmc(naive_model_list),
+  naive_draws_summary = check_draws(naive_draws_list, "protea_naive"),
+
 )
 
 make(fit_protea_model)
[…]

### 2019-10-30 R/plotting.R
add basic forest plot code

@@ -118,3 +118,40 @@ plot_occ <- function (occurrence, filename) {
   )
 
 }
+
+# background for a forest plot
+forest_plot_base <- function (params) {
+
+}
+
+# ad points and bars to a forest plot
+forest_plot_add <- function (params) {
+
+  lines()
+  points()
+
+}
+
+# Produce forest plots for parameters in the protea DSDM, with a panerl each for the
+# informative and noninformative priors.  Priors in grey in the backgound,
+# posteriors in the foreground.
+plot_estimates <- function (protea_estimates, filename) {
+
+  on.exit(dev.off())
+  png(filename,
+      width = 1000,
+      height = 700,
+      pointsize = 30)
+
+  # non-informative prior plot
+  par(mfrow = c(1, 2))
+
+  forest_plot_base()
+  forest_plot_add(noninformative_params)
+
+  forest_plot_base()
+  forest_plot_add(informative_params)
+
+  dev.off()
+
+}

### 2019-10-30 R/modelling.R
change IPM to use new observation model

@@ -738,7 +738,7 @@ build_protea_model <- function (occurrence, informative_priors = FALSE) {
   # a few observations (for now)
   occurrence %>%
     filter(!is.na(presence)) %>%
-    sample_n(150) -> occ_train
+    sample_n(500) -> occ_train
 
   # formulae for the *environmental components* of the vital rate regressions
   formulae <- list(
@@ -901,28 +901,40 @@ build_protea_model <- function (occurrence, informative_priors = FALSE) {
   # observation model parameter
   likelihood_intercept <- normal(0, 10)
 
+  # probability of detection under Poisson sampling
+
   # form the likelihood
-  p_present <- icloglog(likelihood_intercept + log(lambda))
+  K <- (lambda - 1) * likelihood_intercept
+  Lambda <- log1pe(K)
+  p_present <- 1 - exp(-Lambda)
   distribution(occ_train$presence) <- bernoulli(p_present)
 
+  # set up parameters to trace
+  names(betas) <- paste0("beta_", names(betas))
+  names(size_coefs) <- paste0("gamma_", names(size_coefs))
+  params <- c(betas,
+              size_coefs,
+              list(growth_sd = growth_sd,
+                   offspring_size_sd = offspring_size_sd,
+                   likelihood_intercept = likelihood_intercept))
+
   # build the model
-  m <- model(lambda)
+  m <- with(params, model())
+
   # return the model and greta arrays (used for plotting and prediction)
   list(
     model = m,
-    environment_coefs = betas,
-    size_coefs = size_coefs,
-    growth_sd = growth_sd,
-    offspring_size_sd = offspring_size_sd
+    parameters = params,
+    lambda = lambda
   )
 
 }
 
 # do MCMC on the model
-run_mcmc <- function (model_list) {
+run_mcmc <- function (model_list, verbose = FALSE, ...) {
 
   timing <- system.time(
-    draws <- mcmc(model_list$model, verbose = FALSE)
+    draws <- mcmc(model_list$model, verbose = verbose, ...)
   )
 
   list(
@@ -931,19 +943,6 @@ run_mcmc <- function (model_list) {
   )
 
 }
-
-# use the model and MCMC parameter samples to visualise the fitted vital rate
-# relationships (using calculate and relevant design matrices)
-plot_protea_relationships <- function(draws_list, model_list, occurrence) {
-
-}
-
-# use the model and MCMC parameter samples to make posterior predictions of
[…]

### 2019-10-30 R/plotting.R
add forest plotting code

@@ -120,22 +120,49 @@ plot_occ <- function (occurrence, filename) {
 }
 
 # background for a forest plot
-forest_plot_base <- function (params) {
-
+forest_base <- function (params) {
+  locs <- seq(1, 0, length.out = nrow(params))
+  plot(locs ~ mean,
+       data = params,
+       type = "n",
+       ylab = "",
+       xlab = "",
+       axes = FALSE,
+       bty = "n",
+       ylim = c(0, 1),
+       xlim = range(c(params$lower, params$upper)))
+  abline(v = 0, lwd = 1, col = grey(0.5), lty = 3)
+  axis(side = 1, tcl = -0.3)
+  axis(side = 2, lty = 0, at = locs, labels = params$param, las = 2)
+  title(xlab = "estimate", col.lab = grey(0.3))
 }
 
-# ad points and bars to a forest plot
-forest_plot_add <- function (params) {
+# add points and bars to existing forest plot
+forest_points <- function(params, col = grey(0.8), scaling = 1) {
 
-  lines()
-  points()
+  locs <- seq(1, 0, length.out = nrow(params))
+  for (i in seq_len(nrow(params))) {
+    lines(x = c(params$lower[i], params$upper[i]),
+          y = cbind(locs[i], locs[i]),
+          lwd = 1.5 * scaling ^ 2,
+          col = col)
+  }
+
+  points(locs ~ params$mean,
+         pch = 16,
+         cex  = 1 * sqrt(scaling),
+         col = col)
 
 }
 
 # Produce forest plots for parameters in the protea DSDM, with a panerl each for the
 # informative and noninformative priors.  Priors in grey in the backgound,
 # posteriors in the foreground.
-plot_estimates <- function (protea_estimates, filename) {
+plot_estimates <- function(
+  naive_draws_summary,
+  informative_draws_summary,
+  filename
+) {
 
   on.exit(dev.off())
   png(filename,
@@ -143,15 +170,18 @@ plot_estimates <- function (protea_estimates, filename) {
       height = 700,
       pointsize = 30)
 
-  # non-informative prior plot
   par(mfrow = c(1, 2))
 
-  forest_plot_base()
-  forest_plot_add(noninformative_params)
+  # naive prior plot
+  forest_base(naive_params)
+  title(main = "with naive priors", col.main = grey(0.3))
+  forest_points(naive_priors, col = grey(0.9), scaling = 2)
[…]

### 2019-10-30 main.R
don't depend on truncdist any more

@@ -16,7 +16,6 @@ library(RColorBrewer)
 library(dplyr)
 library(maxlike)
 library(pROC)
-library(truncdist)
 
 # modelling
 library(greta.dynamics)

### 2019-10-30 R/modelling.R
run protea sampler in parallel on a server

@@ -1,4 +1,4 @@
-# modelling functions for boht the BBS and protea analyses
+# modelling functions for both the BBS and protea analyses
 
 # given a vector greta array, where each element corresponds to a cell in the
 # raster, an mcmc.list object of posterior samples, return a raster giving the
@@ -522,9 +522,7 @@ compare_bbs_predictions <- function (predictions, data_list) {
 
 }
 
-
 op <- greta::.internals$nodes$constructors$op
-tf <- tensorflow::tf
 
 # normalise a matrix rowwise in exponential space. Ie. Given a matrix x, return a
 # matrix y where:
@@ -534,13 +532,15 @@ tf <- tensorflow::tf
 # or:
 # y[i, ] <- x[i, ] -  log(sum(exp(x[i, ])))
 log_normalise <- function (x) {
-  op("log_normalise", x, tf_operation = "tf_log_normalise")
-}
 
-tf_log_normalise <- function (x) {
-  # skipping batch dimension, do log_sum_exp on rows, then broadcast subtract it
-  sums <- tf$reduce_logsumexp(x, axis = 2L, keepdims = TRUE)
-  x - sums
+  tf_log_normalise <- function (x) {
+    # skipping batch dimension, do log_sum_exp on rows, then broadcast subtract it
+    sums <- tensorflow::tf$reduce_logsumexp(x, axis = 2L, keepdims = TRUE)
+    x - sums
+  }
+
+  op("log_normalise", x, tf_operation = "tf_log_normalise")
+
 }
 
 # load the parameter estimates from Merow et al. and format them as the

### 2019-10-30 main.R

@@ -83,6 +83,7 @@ fit_protea_model <- drake_plan(
     protea_occurrence,
     informative_priors = TRUE
   ),
+
   informative_draws_list = run_mcmc(informative_model_list, verbose = TRUE, sample = hmc(Lmin = 10, Lmax = 15)),
   informative_draws_summary = check_draws(informative_draws_list, "protea_informative"),
 
@@ -95,7 +96,10 @@ fit_protea_model <- drake_plan(
 
 )
 
+# run each sampler independently, each using multiple cores
+future::plan(multisession, workers = 4)
 make(fit_protea_model)
+future::plan(old_plan)
 
 # plot data  --------------------------------------------------------
 

### 2019-12-02 manuscript/ms.Rmd
fix equations

@@
 \gamma &\sim N(1.17, 2.10) \\ \nonumber
-alpha &\sim U(0, \infty)
+\alpha &\sim U(0, \infty)
 \end{align}.
@@
 These normally-distributed implied priors correspond exactly to multivariate normal priors on the vectors of regression coefficients, since the matrix-multiplication that maps between regression coefficients and link-scale response is an affine transformation, under which the multivariate normal distribution is closed.
-The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma} = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$.
+The parameters of the priors over $\beta_F$ are therefore given by: $\mu_F = \mathbf{X}_*^{-1}(\text{-}0.63\, \mathbf{I})$ and $\mathbf{\Sigma}_F = \mathbf{X}_*^{-1} (0.18\,\mathbf{I}) \mathbf{X}_*^{-T}$, where $\mathbf{X}_*$ denotes the design matrix over the $J$ survey stations, $\mathbf{X}_*^{-1}$ its Moore-Penrose pseudoinverse, $\mathbf{X}_*^{-T}$ the transpose of that pseudoinverse, and $\mathbf{I}$ an identity matrix of dimension $J$.
 The prior parameters $\mu_S$ and $\mathbf{\Sigma}_S$ were defined in the same way using the implied prior parameters over $logit(S_{A_j})$: 1.12, and 1.28.

### 2019-12-02 manuscript/ms.Rmd
drop confounded min temperature covariate

@@
 I fitted the model to a random subset of `r bbs_n_train` presence-absence records from the North American Breeding Bird Survey in 2015 [@sauerNorthAmericanBreeding2011].
-I used six environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species), minimum temperature (to capture thermal limits on survival of the species), annual precipitation and mean temperature (as proxies for insect abundance and plant growth, food sources for this species).
-The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the climatic covariates were bioclim layers 6, 1, and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017].
+I used five environmental covariates in $\mathbf{X}$: percentage cover of deciduous forest, coniferous forest, and mixed forest (to represent breeding habitat for this species) mean temperature (to capture thermal limits on survival of the species), and annual precipitation (as a proxy for insect abundance and plant growth, food sources for this species).
+The forest cover layers were obtained from the maxlike R package [@royleLikelihoodAnalysisSpecies2012], and the climatic covariates were bioclim layers 1 and 12 from the WorldClim 2.0 dataset [@fickWorldClimNew1km2017].
 In general, ecological knowledge could be used to select separate covariates to use for each of the submodels.

### 2019-12-02 R/modelling.R

@@ -13,7 +13,7 @@ map_variable <- function (greta_array, draws, raster_template) {
 }
 
 # use all covariates in both models
-surv_covs <- fec_covs <- all_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio6", "bio12")
+surv_covs <- fec_covs <- all_covs <- c("pcMix", "pcDec", "pcCon", "bio1", "bio12")
 
 # calculate survival, given covariates
 get_survival <- function (X, params) {

### 2019-12-02 main.R

@@ -51,7 +51,7 @@ fit_bbs_model <- drake_plan(
   maps_coords = readRDS(file_in("data/clean/maps_stations.RDS")),
 
   # prepare the train/test split and design matrices
-  bbs_formula = ~ 1 + pcMix + pcDec + pcCon + bio1 + bio6 + bio12,
+  bbs_formula = ~ 1 + pcMix + pcDec + pcCon + bio1 + bio12,
   bbs_data_list = prep_data(bbs_formula, bbs_occurrence, bbs_covs, maps_coords),
 
   # model fitting

### 2020-11-25 manuscript/appendix_a.Rmd
minor edits up to Bev Holt

@@
 
-Each section below details a modelling option for a specific component of the DSDM - the observation model, abundance model, or population dynamic model.
+Each section below details a modelling option for a specific component of the DSDM: the observation model, abundance model, or population dynamic model.
 These components are largely inter-operable; a user can pick one modelling option from each category that best fits the ecology of the species of interest and the ecological question at hand.
@@
 
+**$K$ is used both for site carrying capacity and number of covariates?**
+
 ##### Code
@@
 These observation models can also be extended to distribution data that are disaggregated by time, stage, species, or genotype by iterating over the appropriate index.
-A wider range of observation models for distribution data - such as that for disentangling occupancy and detection - can be implemented using other greta extension packages, but are beyond the scope of this manuscript.
+A wider range of observation models for distribution data &mdash; such as that for disentangling occupancy and detection &mdash; can be implemented using other greta extension packages, but are beyond the scope of this manuscript.
 Similarly, observation models for demographic data (e.g.
@@
 Presence-absence data $y_i$ are assumed to be independently Bernoulli-distributed following a probability of presence $p_i$, as in a standard presence-absence species distribution model.
-This probability of presence can be calculated from the relative abundance by considering the presence-absencve data as arising from a truncation of count data: the species recorded as present if one or more individuals were detected, or otherwise a absent.
-By assuming that count follow a Poisson distribution with expected value $\Lambda_i$, the probability of species presence (a count greater than 0) is given by $1-e^{-\Lambda_i}$.
+This probability of presence can be calculated from the relative abundance by considering the presence-absence data as arising from a truncation of count data: the species recorded as present if one or more individuals were detected, or otherwise a absent.
+By assuming that counts follow a Poisson distribution with expected value $\Lambda_i$, the probability of species presence (a count greater than 0) is given by $1-e^{-\Lambda_i}$.
 This is equivalent to the likelihood of a GLM for binary data with complementary log-log link function and linear predictor $ln(\Lambda_i)$.
@@
 
-The relative abundance variable $Lambda_i$ is the parameter of the Poisson distribution, so a Poisson count model is simply given by:
+The relative abundance variable $\Lambda_i$ is the parameter of the Poisson distribution, so a Poisson count model is simply given by:
 
@@
 
-Presence-only data - where records consist of locations where the species has been reported as being present - can be considered as a thinned, inhomogeneous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015].
+Presence-only data &mdash; where records consist of locations where the species has been reported as being present &mdash; can be considered as a thinned, inhomogeneous Poisson point process [@wartonPoissonPointProcess2010;@rennerPointProcessModels2015].
 The likelihood for this model comprises an intractable integral over the study area, which can be approximated in a number of ways that reduce the model to having a Poisson observation model.
@@
 
-where $z_i$ is an indicator variable for whether the observation $y_i$ represents presence-absence or count data, and $\alpha^*$ is a detection rate parameter - in addition to the $\alpha$ used to define $\Lambda_i$- giving the detectability of individuals in the surveys yielding the count data, relative to that in the surveys yielding the presence-absence data.
+where $z_i$ is an indicator variable for whether the observation $y_i$ represents presence-absence or count data, and $\alpha^*$ is a detection rate parameter &mdash; in addition to the $\alpha$ used to define $\Lambda_i$ &mdash; giving the detectability of individuals in the surveys yielding the count data, relative to that in the surveys yielding the presence-absence data.
 
@@
 where $\lambda_i$ is the intrinsic growth rate of the population (which can be calculated as the dominant eigenvalue of $\mathbf{A}_i$), and $M$ is a scaling factor implicitly representing density-dependent effects.
-The carrying capacity of the Beverton-Holt model is given by $(\lambda_i - 1) M$, which we assume to be be proportional to the relative abundance that we can observee via species distribution data (assuming the population is at equilibrium), with the constant of proportionality being a per-individual detection rate - the fraction of individuals in the population detected in the survey from which we obtained our distribution data.
+The carrying capacity of the Beverton-Holt model is given by $(\lambda_i - 1) M$, which we assume to be be proportional to the relative abundance that we can observee via species distribution data (assuming the population is at equilibrium), with the constant of proportionality being a per-individual detection rate; the fraction of individuals in the population detected in the survey from which we obtained our distribution data.
 This detection rate is not identified from the density-dependence parameter $M$, so we replace their product with the parameter $\alpha$ to yield the theoretical relative carrying capacty parameter $K_i = (\lambda_i - 1) \alpha$.

### 2021-05-12 manuscript/appendix_a.Rmd

@@
 
+**BP: Here and Ricker: should A be an NxK matrix, as in first section, above?; Further below, M seems to now be used for number of stages rather than densitiy dependence parameter**
+
 ```{r beverton_holt}
@@
 
-An approach similar to this can also used be used to compute relative abundances from a static transition matrix for a  matrix metapopulation model - in that model the stable distribution between sites already provides a relative abundance estimate, so there is no need for a model to relate the intrinsic growth rate to relative abundance.
+An approach similar to this can also used be used to compute relative abundances from a static transition matrix for a  matrix metapopulation model: in that model the stable distribution between sites already provides a relative abundance estimate, so there is no need for a model to relate the intrinsic growth rate to relative abundance.
 See [*Dispersal with a static transition matrix*](#static_dispersal) for details.
@@
 
-In static transition models considered above, carrying capacity is modelled as some function of the intrinsic growth rate, an relationship which implicitly represents density dependence.
-Rather than considering that density dependence isa  function of the growth rate, it may be more ecologically realistic to explicitly model the vital rates as being density-dependent - that is having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point.
+In static transition models considered above, carrying capacity is modelled as some function of the intrinsic growth rate; a relationship which implicitly represents density dependence.
+Rather than considering that density dependence is a function of the growth rate, it may be more ecologically realistic to explicitly model the vital rates as being density-dependent: that is, having the survival probabilities or fecundity rates at a given time point depend on the population density at the previous time point.
 This enables incorporation of more detailed ecological knowledge about how vital rates drive abundance, which may be important to consider if the answer to the ecological question of interest depends on density dependence.
@@
 across all life stages $M$) in the previous time point.
-For example, these density-dependent relationships could be a linear function of population density - controlled by parameters $\alpha_F$ and $\alpha_S$ - on the link-scale of the vital rates.
+For example, these density-dependent relationships could be a linear function of population density, controlled by parameters $\alpha_F$ and $\alpha_S$ on the link-scale of the vital rates.
 Where $\alpha_F$ and $\alpha_S$ are negative, density-dependence will lead to self-regulating populations that will ultimately converge on a finite carrying capacity.
@@
 \end{align}
-where $\alpha_F$ and $\alpha_S$ are negative scalar parameters controlling the strength of density dependenceadnt he detection rate, and $T$ is positive integer giving a number of iterations chosen such that the population sizes have converged to withing a tolerance, i.e.
+where $\alpha_F$ and $\alpha_S$ are negative scalar parameters controlling the strength of density dependence and the detection rate, and $T$ is positive integer giving a number of iterations chosen such that the population sizes have converged to within a tolerance, i.e.
 $|\mathbf{n}_{i,T} - \mathbf{n}_{i,T-1}| / \mathbf{n}_{i,T-1} < \epsilon$ for some small positive tolerance $\epsilon$.
@@
 
-Where the data or questions of the analysis are such that equilibrium population sizes are not representative of the population, the density-dependent vital rate model above can be iterate for predetermined number of timepoints $T$, such that each time $t$ corresponds to an observed time in the distribution data.
+Where the data or questions of the analysis are such that equilibrium population sizes are not representative of the population, the density-dependent vital rate model above can be iterated for predetermined number of timepoints $T$, such that each time $t$ corresponds to an observed time in the distribution data.
 For example, where the impacts of short-term variation in environmental conditions (environmental stochasticity) on population sizes is important, the transition matrix $\mathbf{A}_{i,t}$ at each time can be computed as a function of the design matrix of observed environmental conditions $\mathbf{X}_t$ at that time, and the current population density.

### 2021-05-12 manuscript/ms.Rmd

@@
 3.
-Example applications mapping the Carolina Chickadee in North Ameerica, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates.
+Example applications mapping the Carolina Chickadee in North America, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates.
 Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.
@@
 
-Whilst users typically have some knowledge of the processes driving these aspects of species distributions - such as the species' life history, approximate survival and fecundity rates or dispersal ability - incorporating this knowledge into correlative statistical models is problematic.
-On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed pyhsiological modelling [@kearneyModellingSpeciesDistributions2008].
+Whilst users typically have some knowledge of the processes driving these aspects of species distributions &mdash; such as the species' life history, approximate survival and fecundity rates or dispersal ability &mdash; incorporating this knowledge into correlative statistical models is problematic.
+On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed physiological modelling [@kearneyModellingSpeciesDistributions2008].
 
@@
 DRMs model population growth rates using a correlative model and environmental covariates, just as a correlative SDM models the abundance or probability of presence of a species. 
-DRMs can also be fitted to distribution data, though they are intended for use with timeseries of abundance or presence-only data, since repeat observations are needed to accurately quantify population growth rates and dispersal.
+DRMs can also be fitted to distribution data, though they are intended for use with time series of abundance or presence-only data, since repeat observations are needed to accurately quantify population growth rates and dispersal.
 
-Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change - for example a conservation intervention like predator control - that affects a single vital rate.
+Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change &mdash; for example a conservation intervention like predator control &mdash; that affects a single vital rate.
 
@@
 By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
-Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the species’ distribution [@merowUsingIntegralProjection2014].
+Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the distribution of a species [@merowUsingIntegralProjection2014].
 
@@
 where $\mathbf{X}$ is a design matrix of environmental covariates and an intercept dummy variable for the fecundity and survival submodels and $\beta_F$ and $\beta_S$ are vectors of regression coefficients for the two submodels.
-Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale - to maintain the (0,1) constraint on survival probabilities.
+Rather than separately modelling spatial variation in adult and juvenile fecundity, I assume that the environmental predictors of survival are the same in both stages, but that juvenile survival differs from adult survival by a constant amount on the logit scale, this maintains the (0,1) constraint on survival probabilities.
 
@@
 
-where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the populations at location $i$.
+where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the population at location $i$.
 The theoretical carrying capacity at each site $K_i$ is computed from the carrying capacity relationship of the Beverton-Holt model.
@@
 The softplus transformation is a smooth version of the Heaviside function (which sets negative values to zero), so $\Lambda_i$ closely approximates $K_i$ for most $lambda_i > 1$, and rapidly approaches zero below $lambda_i = 1$.
-This formulation is statistically convenient, but as discussed in the text below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
+This formulation is statistically convenient, but as discussed below, DSDMs can incorporate more complex and realistic forms of density-dependence if required.
 The parameter $\alpha$ controls the relationship between the intrinsic population growth rate and relative abundance.
-It is a compound of two parameters that are only identified together: the Beverton-Holt density-dependence parameter relating growth rates to carrying capacities; and the detection rate - the fraction of individuals in the total population that are detected during a presence-absence survey at site $i$.
+It is a compound of two parameters that are only identified together: the Beverton-Holt density-dependence parameter relating growth rates to carrying capacities; and the detection rate &mdash; the fraction of individuals in the total population that are detected during a presence-absence survey at site $i$.
 From $\Lambda_i$, we obtain the probability $p_i$ of detecting (one or more individuals of) the species $y_i$, under a poisson sampling distribution [@pagelQuantifyingRangewideVariation2014].
@@
 The spatially-averaged fecundity and adult survival estimates from @ryuDevelopingPopulationModels2016 were used to define priors on regression coefficients for the vital rate submodels.
-These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) - locations where the species occurs.
+These estimated vital rates correspond to populations at the $J$ survey stations used in the recapture analysis (station names provided in the supplemental data to @ryuDevelopingPopulationModels2016) &mdash; locations where the species occurs.
 I defined a prior on $\beta_F$ and $\beta_S$ such that values were *a priori* more likely if they resulted in estimates of $F_j$ and $S_{A_j}$, for each survey station location $j$, that were more consistent with the estimates in @ryuDevelopingPopulationModels2016.
@@
 However, in the absence of a strong belief that any of these covariates would affect only one vital rate, all six covariates were used in both submodels.
-These covariate values were centred and scaled (before calculation of the priors on $\beta_F$ and $\beta_S$) to enable comparison of regreesion coefficients as effect sizes.
+These covariate values were centred and scaled (before calculation of the priors on $\beta_F$ and $\beta_S$) to enable comparison of regression coefficients as effect sizes.
 
@@
 
-Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates - for example constraining spatial estimates of fecundity and survival to within a range of plausible values - via prior distributions or changes to model structure.
+Note that this model could undoubtedly be improved by including more information of the species ecology and vital rates &mdash; for example constraining spatial estimates of fecundity and survival to within a range of plausible values &mdash; via prior distributions or changes to model structure.
 As with any model, the reliability of these vital rate and distribution predictions will depend on the assumptions made and information provided by the modeller.
@@
 Specifically: models with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values.
-This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], thought it necessarily restricts the scope of these models to those with only continuous parameters.
-The model variations I propose below encompass a wide range of features whist remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.
+This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], though it necessarily restricts the scope of these models to those with only continuous parameters.
+The model variations I propose below encompass a wide range of features whilst remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.
 
@@
 The matrix iterations will then converge to a stable state; the population carrying capacity or the expected abundance.
-this cna be achieved by iterating populations a sufficient number of time $K$ for the populations to converge on stable population sizes.
+this can be achieved by iterating populations a sufficient number of times $K$ for the populations to converge on stable population sizes.
 
@@
 for iteration $k$ in $K$.
- note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps (T) to get a temporally-static density-dependent model.
-In practice, T can be determined adaptively, iterating a sufficient number of times for N_i to converge and then terminating.
+ note that this can be combined with Eq 5, or can be iterated for a sufficient number of timesteps ($T$) to get a temporally-static density-dependent model.
+In practice, $T$ can be determined adaptively, iterating a sufficient number of times for $N_i$ to converge, and then terminating.
 
@@
 
-For example, theese linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008].
+For example, these linear models could be applied to very large numbers of covariates and transformations thereof in combination with regularising priors to perform automatic variable selection, similarly to the popular Maxent SDM software[@phillipsModelingSpeciesDistributions2008].
 Or to specify mixed-effects models to account for clustered sampling designs, or use random slopes to share environmental reponse information between species in a biotic interaction DSDM.
 Instead of linear models, non-linear functions could be introduced with generalisd additive models or Gaussian process models, for which there are existing greta extension packages: greta.gam and greta.gp [@goldingGretaGamGeneralised2019; @goldingGretaGpGaussian2019].
-These can also be used to model spatial and teemporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables.
+These can also be used to model spatial and temporal random effects and account for drivers of species vital rates that are not captured in the measured environmental variables.
 Prespecified mathematical relationships or covariate values can also be included, enabling incorporation of  spatial and temporal predictions from physiological models.
@@
 
-Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty those predictions, there is no spatially-varying information on vital rates with which to test those assumptions for this species.
-The time and cost required to robustly estimate how a species' vital rates vary with environmental covariates means that such estimates have been produced for relatively few species - one of the motivations for DSDMs.
+Whilst the model presented in Example 1 makes predictions of spatial variation in vital rates, and provides estimates of uncertainty in those predictions, there is no spatially-varying information on vital rates with which to test those predictions for this species.
+The time and cost required to robustly estimate how a species' vital rates vary with environmental covariates means that such estimates have been produced for relatively few species; one of the motivations for DSDMs.
 
@@
 
-Example 1 indicates that identifiability issues do not affect predictions of the probability of presence - even if the data and priors do not provide sufficient information to disentangle vital rates.
+Example 1 indicates that identifiability issues do not affect predictions of the probability of presence, even if the data and priors do not provide sufficient information to disentangle vital rates.
 This example also suggests that a minimal amount of information is required to make the vital rates identifiable.
@@
 Fitting the analytic version of the model with the same Bayesian inference routine took `r round(bbs_analytic_time)` seconds on the same machine.
-Such convenient and computationally efficient analytic solutions will not be available for all DSDM models however - particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal.
+Such convenient and computationally efficient analytic solutions will not be available for all DSDM models, however,particularly where matrices are large or encode more complex demography, such as trait-dependent or temporally stochastic vital rates, density dependence, or dispersal.
 Moreover, deriving analytic solutions for quantities relating to matrix population models is likely to be outside the skill-set of most ecologists, and in any case would force users to focus on model implementation details rather than the ecology of their study species.
@@
 The greta extension package greta.dynamics provides methods for efficiently analysing multiple population transition matrices to determine their intrinsic growth rates and stable stage distributions.
-The package also enables efficient analysis of dynamic transition matrices, where vital rates and therefor the matrix at each iteration depend on the previous population size or temporally-varying covariates - enabling modelling of density-dependent or temporally stochastic demographics in DSDMs.
-Whilst dispersal can already be implemented by convolving population transition matrices with a dispersal matrix, future versions of the R package will facilitate more computationally efficient algorithms for accounting for dispersal - leveraging the inherent sparsity in dispersal matrices.
+The package also enables efficient analysis of dynamic transition matrices, where vital rates and therefore the matrix at each iteration depend on the previous population size or temporally-varying covariates,enabling modelling of density-dependent or temporally stochastic demographics in DSDMs.
+Whilst dispersal can already be implemented by convolving population transition matrices with a dispersal matrix, future versions of the R package will facilitate more computationally efficient algorithms for accounting for dispersal, leveraging the inherent sparsity in dispersal matrices.
 

### 2021-05-12 manuscript/ms.Rmd
merge Ben's edits and some of my own

@@
 Quantifying the distributions and abundances of species is a problem central to ecology.
-Correlative species distribution models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they have questionable ability to predict how distributions will change in response to environmental change, habitat loss, or human interventions.
+Correlative species distribution models (SDMs) are are widely used, and whilst they perform well at predicting current distributions, their lack of ecological mechanism means they often perform poorly at predicting how distributions will change in response to environmental change, habitat loss, or human interventions.
 
 2.
-I propose demographic species distribution models (DSDMs): matrix population models that are statistically fitted to species distribution data, to form a mechanistically-based statistical model for species distributions.
+I propose demographic species distribution models (DSDMs): matrix population models that are statistically fitted to species distribution data to form a mechanistic statistical model for species distributions.
 This approach differs from existing mechanistic SDMs in that it makes use of commonly available species distribution data (abundance, presence-only, or presence-absence), explicitly models how vital rates such as fecundity and survival vary over space, and enables - but does not require - the incorporation of prior knowledge of a species' ecology.
@@
 3.
-Example applications mapping the Carolina Chickadee in North America, and a Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs, and produce plausible spatial estimates of vital rates.
-Even when DSDMs are provided little information on the ecology of the target species, predictions of distributions are well-identified, and increased uncertainty about population dynamics is reflected in more uncertain estimates of vital rates.
+Example applications mapping the Carolina Chickadee in North Ameerica, and a species of Protea in southern Africa indicate that DSDMs have similar predictive performance to SDMs and produce plausible spatial estimates of vital rates.
+Even when DSDMs are provided with little information on the ecology of the target species, predictions of distributions are well-identified.
+Where there is uncertainty about a species' population dynamics, this uncertainty is reflected in more uncertain estimates of vital rates.
 
@@
 DSDMs can be extended to consider features of population dynamics such as dispersal, density dependence, and biotic interactions.
-Being specifically designed for efficient statistical inference, they are a promising replacement for correlative SDMs where study aims require the incorporating of more ecological mechanism.
+Being specifically designed for efficient statistical inference, they are a promising replacement for correlative SDMs where study aims require the incorporation of more ecological mechanism.
 

### 2021-05-12 manuscript/ms.Rmd
edits to intro and DSDM section

@@
 However the arbitrary, flexible functional forms of these correlative species distribution models (SDMs) make them poor at predicting to novel conditions [@dormannCorrelationProcessSpecies2012].
-Moreover, models of species distributions are often needed to predict new distributions in response to changes other than the environmental predictors that correlative SDMs rely upon. For instance
+Moreover, models of species distributions are often needed to predict new distributions in response to changes other than the environmental predictors that correlative SDMs rely upon.
+For instance <the impact of a novel pathogen or the introduction of a competitor, on a species' distribution and abundance.>
 
-Whilst users typically have some knowledge of the processes driving these aspects of species distributions &mdash; such as the species' life history, approximate survival and fecundity rates or dispersal ability &mdash; incorporating this knowledge into correlative statistical models is problematic.
+Whilst users typically have some knowledge of the processes driving these aspects of species distributions &mdash; such as the species' life history, approximate survival and fecundity rates or dispersal ability &mdash; it is not generally possible to incorporate this knowledge into correlative statistical models in a principled way.
 On the other hand, ecologists rarely have enough information to accurately predict species distributions from their knowledge of the species life history and demographics alone, since quantifying how vital rates vary over space typically requires large-scale long-term field studies [@merowUsingIntegralProjection2014] or detailed physiological modelling [@kearneyModellingSpeciesDistributions2008].
 
-When no single modelling approach is available that fits their needs, species distribution modellers needing to incorporate more ecological mechanism into their predictions must employ a number of tricks to shoehorn their ecological knowledge into correlative SDMs.
+When no single modelling approach is available that fits their needs, species distribution modellers that require more ecological mechanism in their predictions may rely on a number of tricks to shoehorn their ecological knowledge into correlative SDMs.
 For instance, manipulating distribution data to remove the signal of dispersal limitation before model fitting [@elithArtModellingRangeshifting2010], or treating correlative SDM predictions as a proxy for habitat carrying capacity in a secondary spatial population simulation model [@andersonDynamicsRangeMargins2009].
 Whilst these approaches can be effective, they often make strong and questionable assumptions (for example about the relationship between habitat suitability and carrying capacity) and require the modeller to focus on implementation details, rather than the ecology of the species they are studying.
-Models that explicitly consider the demographic processes relevant to distribution modelling task in hand are therefore appealing.
+Models that explicitly consider the demographic processes relevant to the distribution modelling task in hand are therefore appealing.
 
-A number of models have been made to improve the realism in species distribution models, whilst still enabling them to be fitted to the distribution data that are commonly available.
-For example, dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes and spatial variation in population growth rates [@pagelForecastingSpeciesRanges2012].
+Models have been proposed to improve the realism in species distribution models, whilst still enabling them to be fitted to the distribution data that are commonly available.
+Dynamic range models (DRMs) extend correlative SDMs by explicitly modelling dispersal processes and spatial variation in population growth rates [@pagelForecastingSpeciesRanges2012].
 DRMs model population growth rates using a correlative model and environmental covariates, just as a correlative SDM models the abundance or probability of presence of a species. 
@@
 
-Because they do not model vital rates directly DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change &mdash; for example a conservation intervention like predator control &mdash; that affects a single vital rate.
+Because they do not model vital rates or population dynamic processes directly, DRMs, like correlative SDMs, are unable to predict the impact on distributions of a change that affects a single vital rate (e.g.
+only survival or only fecundity).
+It is also not possible to extend DRMs to incorporate important population dynamic processes other than dispersal - such as density dependence, seed-banks, biotic interactions, and evolution.
 
-<explicitly modelling vital rates, because it's useful>
+By contrast to DRMs, matrix population models and their recently developed extension integral projection models (IPMs) are demographic models of the dynamics of a population at a single location.
+Both approach focus on the construct and analysis of a transition matrix (or kernel) representing key demographic processes [@merowAdvancingPopulationEcology2014].
+Spatial IPMs^[Note that @merowUsingIntegralProjection2014 use the term 'demographic distribution model' to describe their spatial application of an integral projection model, however since that title could be applied to a very broad range of models including the approach introduced in this paper, we use the term spatial integral projection model to refer to their approach] extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the distribution of a species [@merowUsingIntegralProjection2014].
+Spatial IPMs require data on vital rates at a number of sites, in order to be able to characterise how vital rates vary as a function of multiple spatial covariates.
+Such data are unlikely to be available for modelling most species, restricting the applicability of this approach to a small subset of distribution modelling applications.
 
-<MPMs & IPMs do this at a single location - widely used in ecology, and even whole databases available>
-
-By contrast, matrix population models and integral projection models (IPMs) are demographic models of single populations that construct and analyse a transition matrix representing key demographic processes [@merowAdvancingPopulationEcology2014].
-Spatial IPMs extend this idea to map distributions by modelling demographic rates as a function of covariates, then analysing transition matrices under different conditions to map emergent demographic quantities (such as the intrinsic growth rate) to characterize the distribution of a species [@merowUsingIntegralProjection2014].
-
-<SIPMs (DDMs in Merow - note on nomenclature) apply these to large spatial scales.
-Whilst they rely on vast amnounts field data (so are not applicable in most SDM situations), they provide a link between demography and distribution, making it possible to perform statistical inference on them using distribution data.
-This is what I call a DSDM - the subject of this manuscript.>
+However, spatial IPMs provide a mechanistic mathematical link between environmental covariates, demography, and distributions, making it possible to perform statistical inference on these spatially-varying vital rates from distribution data.
+I refer to this framework of fitting a spatially-varying MPM or IPM to distribution data as a demographic species distribution model.
+This approach draws on the strengths of both DRMs and spatial IPMs (and indeed includes both as special cases), enabling the user to make use of both demographic data and distribution data, to the extent that either is available for the target species.
 
@@
 Nor is it obvious that such a model can be fitted to data with sufficiently ease and speed to be useful to ecologists.
-The rest of this manuscript addresses these issues, introducing the basic model and describing a number of potential extensions.
-Example 1 illustrates a basic DSDM for presence/absence distribution data, requiring minimal prior knowledge and demonstrates that spatial variation in the two different vital rates can indeed be disentangled.
-Example 2 then develops a more complex DSDM incorporating detailed knowledge of a species' ecology, and illustrates that the relationships between vital rates and environmental variables learned from distribution data do broadly agree with field-derived estimates.
+The rest of this manuscript addresses these issues, introducing both a basic model and a more complex example, and describing a number of potential extensions.
 
@@
 
-<general equations - a number of options for observation models>
-<blame link between growth rate and probability of presence on Jorn, point reader to section on extensions for density-dependent solution>
-<suggest doing Bayesian inference - propagate uncertainty and incorporate any prior information to help the model disentangle the spatial signal>
+## Defining the model
+
+In correlative species distribution modelling, the analyst begins by selecting a statistical or machine-learning modeling framework that is appropriate to the data type and is thought to perform well.
+By contrast, the construction of a DSDM begins with the selection of a population matrix structure that is appropriate to represent relevant aspects of the population dynamics of the species, and the questions being asked of the model.
+For example, in a species where vital rates strongly depend on the age or size of individuals, a matrix model representing multiple life stages structured by age or size would likely be appropriate.
+If an aim of the analysis relates to the dispersal of the species (e.g. a range-shifting species), then explicit consideration of of dispersal in the momdel would be warranted.
+
+The intention of this paper is to introduce and demonstrate a general modelling approach and related class of models, rather than proposing a single matrix structure that would be expected to apply in all cases.
+In this sense, the proposed DSDM framework is akin to mathematical modelling, rather than correlative statistical modelling.
+However, a number of different features of population dynamics can easily be included in DSDMs, in a modular way.
+The DSDM structure given in equations 2-3 is simple version for presence-absence data which makes a number of assumptions about the species ecology.
+This structure can easily be extended to accommodate alternative data types and demographic processes, and answer a range of ecological questions, by extending this basic model in a modular way.
+These extensions are discussed below.
+The appendix provides notation and code for fitting a number of models that can be fitted within this framework, capturing a wide range of ecological processes including: density dependence, non-stationary population dynamics, dispersal and non-equilibrium distributions, biotic interactions and evolutionary dynamics.
+
+Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are both based on transition matrices, and amenable to efficient MCMC methods.
+By taking a Bayesian approach, analysts are able to incorporate ecological knowledge into models, but also account for uncertainty in this knowledge and allow assumptions to be updated on fitting to distribution data.
+Fitting by MCMC also fully captures uncertainty about model parameters (poor model identifiability) and propagates it into predictions, which can be important where a species' dstribution can be equally well explained by the same spatial pattern applied to one or other vital rate models.
+This constrains the considered class of models to those with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values.
+This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], though it necessarily restricts the scope of these models to those with only continuous parameters.
+The model variations I propose encompass a wide range of features whilst remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that they should scale well to the large datasets commonly used for fitting SDMs.
 
@@
 
-# Example 1. A DSDM for the Carolina Chickadee
+## Example 1. A DSDM for the Carolina Chickadee
 
@@
 K_i &= (\lambda_i - 1) \alpha \\ \nonumber
-\lambda_i &= eigval(\mathbf{A}_i)
+\lambda_i &= \rho(\mathbf{A}_i)
 \end{align}
 
-where $eigval()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the population at location $i$.
+where $\rho()$ denotes the dominant eigenvalue of a symmetric matrix, yielding the intrinsic growth rate $\lambda_i$ of the population at location $i$.
 The theoretical carrying capacity at each site $K_i$ is computed from the carrying capacity relationship of the Beverton-Holt model.
@@
 
-# Alternative DSDM specifications
-
-The DSDM structure given in equations 2-3 is simple version for presence-absence data which makes a number of assumptions about the species ecology.
-This structure can easily be extended to accommodate alternative data types and demographic processes, and answer a range of ecological questions.
-I detail a number of proposed variants below, and provide full notation and example code in Appendix A.
-
-Whilst statistical inference against distribution data could be performed on almost any demographic model, I restrict the models discussed here to those that are amenable to efficient MCMC methods.
-Specifically: models with likelihoods that are tractable (the exact value of likelihood can be computed from the parameters and data), and can be differentiated with respect to parameter values.
-This enables the use of modern gradient-based MCMC samplers [@monnahanFasterEstimationBayesian2017], though it necessarily restricts the scope of these models to those with only continuous parameters.
-The model variations I propose below encompass a wide range of features whilst remaining within these constraints, and can also make use of efficient vectorised and linear algebra routines, meaning that should scale well to the large datasets commonly used for fitting SDMs.
-
 ## Alternative observation models
@@
 
+<blame link between growth rate and probability of presence on Jorn, point reader to section on extensions for density-dependent solution>
+
 Since these observation models all relate to $\Lambda_i$, multiple likelihoods could be specified to form an *integrated* DSDM to be fitted to multiple datasets.
@@
 
-# Temporal dynamics
+## Temporal dynamics
 
@@
 
-# Continuous stages
+## Continuous stages
 
